### 2016-07-25 latex/input/abstract.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

%abs

We demonstrate a computational method to predict the clinical
phenotypes of a patient from raw metagenomic sequence read data. We  compared 
two state of the art programs for annotating the sequence data, 
UCLUST and Kraken, and using their output for feature generation. 
We apply these programs 
to a set of over 1.3 million reads from 
904 patients, some of whom have liver cirrhosis, encephalopathy due to liver cirrhosis, or 
neither disease. Once the reads have been processed by UCLUST or 
Kraken, we use  Support Vector Machines to setup the 
clinical phenotype prediction problem.  
%
We find that too many false negatives are being predicted by 
the classifier. In order to address the issue, we scale features to
improve the classification model and evaluate the 
end results on a held-out test set. We
demonstrate our approach works quickly and 
accurately with an 85.64\% success rate when we use the 
UCLUST representation. We also find that UCLUST generally performs better than Kraken, with the latter having an 80.66\% success rate. We also test our classifier on several subsets of the data, with success rates ranging from 69.81\% to 96.72\%.

### 2016-07-25 latex/input/conclusion.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

%Conclusion

We have demonstrated a computational method 
to predict clinical phenotypes using metagenomic sequence data with intermediate 
OTU representation using state-of-the-art binning and taxonomic profiling approaches. 
%
%
Our results showed that 
the UCLUST representation 
achieved 85.64\% accuracy and 70.46\% F1-Score. Kraken performed less well, with 80.66\% accuracy and 53.34\% F1-Score. 
For the 1.3 million sequence reads UCLUST ran in 180 seconds and Kraken for 120 seconds, indicating that both performed efficiently enough to feasibly use them with larger data sets.

We then used UCLUST to test three one-versus-one clinical phenotype classifiers,
which generally produced stronger results than our original 
comparison of the "Encephalopathy" class versus all other 
classes. In particular, the classification of 
"Encephalopathy" versus "Control" performed the best. This intuitively seems reasonable. Whereas 
the original comparison requires making distinctions between 
those who have both encephalopathy and liver cirrhosis, those who have only the latter, and those who have neither, this pairwise comparison focuses on the simpler distinction 
between patients who have both diseases and those who have neither. The other two pairwise comparisons also similarly simplify the classification problem. When Kraken was used for the three pairwise comparisons, it had the peculiar effect of drastically increasing the recall while maintaining roughly the same accuracy and precision. This is most likely due to the fact that the pairwise comparisons involved data sets that were less negatively-skewed and thus didn't cause the classifier to predict as negatively based on Kraken's results. This is combined with the relatively strong accuracy and precision numbers in Kraken's original run, versus its poor recall numbers that left plenty of room for improvement.

Our technique of applying a multiplier to feature values for positive examples produced the effect of improving the recall in most cases. These results were generally corroborated by similar improvements in the cross validation results. Another effect was that precision and accuracy results were also often improved after applying this technique. This intuitively makes sense, as switching false negatives to true positives also positively affects the accuracy and precision ratios. Overall, this feature engineering process appeared to have a generally positive effect on most of the results, suggesting that similar techniques for improving classification results merit further investigation.

Based on our results demonstrating over 85\% accuracy for the original run and even better results for some of the pairwise comparisons, we believe the use of machine learning
classification techniques to predict clinical phenotypes 
merits continued research. There are many potential practical 
uses of such technology for medical purposes, including 
diagnostics and research into diseases and their 
relationship with the human microbiome.

\subsection{Future Work}

Kraken is developed to be fast, accurate, and precise, all 
of which we found to be true. However, it was not as accurate as a representation in comparison to 
UCLUST. Kraken's poor performance was mostly due to  poor matching of $k$-mers in the reads to taxonomic classes in its database. Often, a read was only a one or two percent match to the class that Kraken labeled it as. This 
may be due in part to database constraints. Kraken's full database is very computationally expensive to download and build. We 
used a 4 GB custom subset of the full 160 GB database. It is 
possible that the full database would have led to improved performance, and this is 
a future research question.

There is also an opportunity to use a much larger set of sequence reads. Having more training examples for the classifier would  lead to a stronger and more generalizable model for predicting clinical phenotypes. The use of other annotation and classification methods could also be explored. Finally, a limited set of feature selection and scaling techniques were explored, and future research could investigate further how such techniques could be used to improve predictions.

### 2016-07-25 latex/input/experiments.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

%EXP

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{./diagrams/PreFeatEng}
\caption{This diagram shows the results for the comparison between UCLUST and Kraken when classifying "Encephalopathy" versus all other classes. UCLUST had an accuracy and F1-Score of 84.53\% and 67.44\% respectively. Kraken didn't perform as well, with an accuracy of 79.56\% and an F1-Score of 43.08\%. \label{prefeature}}
\end{figure}

\begin{table*}[t]
\begin{center}
\caption{Comparative Run Time of UCLUST and Kraken Representations. \label{tab:time}}
\begin{tabular}{|c|c|}\hline
Representation & Time\\\hline
UCLUST & 180s\\\hline
Kraken & 120s\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Comparative Performance of UCLUST and Kraken Representations. 
\label{tab:comp}}
\begin{tabular}{|c|cccc|cccc|}\hline
Representation & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
UCLUST & 84.53 & 78.38 & 59.18 & 67.44 & 88.26 & 83.22 & 67.39 & 74.47\\\hline
Kraken & 79.56 & 87.5 & 28.57 & 43.08 & 78.18 & 70.31 & 24.46 & 36.29\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using UCLUST. \label{tab:pairuclust}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 88.68 & 90 & 86.54 & 88.23 & 84.47 & 81.77 & 81.77 & 81.77\\\hline
Encephalopathy vs. Control & 96.72 & 95.35 & 100 & 97.62 & 94.26 & 93.2 & 100 & 96.48\\\hline
No Encephalopathy vs. Control & 83.78 & 82.61 & 100 & 90.48 & 87.5 & 86.96 & 99.59 & 92.85\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using Kraken. \label{tab:pairkraken}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 67.92 & 82.14 & 44.23 & 57.50 & 68.94 & 72.07 & 44.20 & 54.79\\\hline
Encephalopathy vs. Control & 78.69 & 76.92 & 97.56 & 86.02 & 79.51 & 81.14 & 96.35 & 88.09\\\hline
No Encephalopathy vs. Control & 77.03 & 77.03 & 100 & 87.02 & 81.42 & 81.42 & 100 & 89.76\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Comparative Performance After Feature Engineering. \label{tab:pfe}}
\begin{tabular}{|c|cccc|cccc|}\hline
Representation & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
UCLUST & 85.64 & 79.49 & 63.27 & 70.46 & 88.54 & 82.58 & 69.57 & 75.52\\\hline
Kraken & 80.66 & 76.92 & 40.82 & 53.34 & 80.8 & 75.86 & 35.87 & 48.93\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using UCLUST After Feature Engineering. \label{tab:pairuclustfe}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 88.68 & 90 & 86.54 & 88.23 & 86.35 & 83.24 & 85.08 & 84.15\\\hline
Encephalopathy vs. Control & 96.72 & 95.35 & 100 & 97.62 & 95.08 & 94.12 & 100 & 96.97\\\hline
No Encephalopathy vs. Control & 87.84 & 86.36 & 100 & 92.68 & 91.22 & 90.57 & 99.59 & 95.05\\\hline
\end{tabular}
\end{center}
\end{table*}

\begin{table*}[t]
\begin{center}
\caption{Three Pairwise Comparisons Using Kraken After Feature Engineering. \label{tab:pairkrakenfe}}
\begin{tabular}{|c|cccc|cccc|}\hline
Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
Encephalopathy vs. No Encephalopathy & 69.81 & 83.33 & 48.08 & 60.98 & 77.18 & 80.88 & 60.77 & 69.40\\\hline
Encephalopathy vs. Control & 86.89 & 86.67 & 95.12 & 90.70 & 87.30 & 88.15 & 96.88 & 92.31\\\hline
No Encephalopathy vs. Control & 77.03 & 77.78 & 98.25 & 86.82 & 82.77 & 82.53 & 100 & 90.43\\\hline
\end{tabular}
\end{center}
\end{table*}

Table \ref{tab:time} shows 
the run time of each representation in seconds. We had total of 1.3 million sequence reads. UCLUST unsupervised clustering 
took 180 seconds, while Kraken's OTU database lookups took 120 seconds.

Figure \ref{prefeature} and Table \ref{tab:comp} show 
the classification performance of UCLUST and 
Kraken in distinguishing patients in the "Encephalopathy" class versus 
the other classes. UCLUST correctly predicted the clinical 
phenotype 84.53\% of the time, and had a precision of 78.38\% and recall of 59.18\% on the held-out set. LOOCV results were 
slightly better for each metric than results on the test 
set by about 4-8\%. Kraken had an accuracy of 79.56\% with an even 
larger disparity between precision and recall (87.5\% for the former and 
28.57\% for the latter), which led to a relatively poor F1-score of 43.08\%. LOOCV results for Kraken 
were worse by 1.38\% for accuracy, 17.19\% for precision, 4.11\% for recall, and 6.79\% for F1-Score. As 
the results show, UCLUST generally performed better as a feature 
representation, with 4.97\% higher accuracy and 24.36\% higher F1-score.

Table \ref{tab:pairuclust} displays the performance of UCLUST-based classification 
with regards to training three one-versus-one clinical phenotypic classifiers.  The first comparison 
distinguished between the "Encephalopathy" class and the "No Encephalopathy" class, effectively determining whether or not a patient who had liver cirrhosis also had encephalopathy. Results were much improved when compared with the findings in table \ref{tab:comp}. The accuracy, precision, recall, and F1-score were all  86-90\%. LOOCV
results were worse by about 4-8\% per metric. The 
second comparison distinguished between the "Encephalopathy" class and 
the "Control" class, effectively determining whether a 
patient had encephalopathy caused by liver cirrhosis or had neither disease. This comparison had the best results of the ones we examined, with greater than 95\% performance in all metrics and 100\% recall. The cross validation
results were about 1-2\% worse for each metric except recall, which was still 100\%. The 
third comparison distinguished between the "No Encephalopathy and liver cirrhosis" class and the "Control" class, effectively determining which among the patients lacking encephalopathy had liver cirrhosis. This comparison had an accuracy of 83.78\% and precision of 82.61\%, but had a 100\% recall which led to a high F1-score of 90.48\%. Cross validation results were better by 3.72\% for accuracy, 4.35\% for precision, and 2.37\% better for F1-Score, with essentially the same recall.  Apart from the accuracy and precision of the third comparison, these three pairwise comparisons generally performed better across the board than the results in Table \ref{tab:comp}, with the second pairwise comparison having the best performance.

Table \ref{tab:pairkraken} displays the performance of Kraken with regards to the same three pairwise comparisons described above. For the Encephalopathy vs. No Encephalopathy run, the accuracy was 67.92\%, with 82.14\% precision, 44.23\% recall, and 57.50\% F1-Score. Cross validation produced a precision that was 10.07\% lower but largely the same accuracy and recall. The Encephalopathy vs. Control run produced good results, with 78.69\% accuracy, 76.92\% precision, 97.56\% recall, and 86.02\% F1-Score. LOOCV results mostly confirmed those results, except the precision, which was 4.22\% higher. The No Encephalopathy vs. Control run had quite similar results to the Encephalopathy vs. Control run. The main pattern that appears to emerge from these results is that the recall for Kraken was generally far superior in the pairwise runs than in the original run, which increased the F1-Scores but resulted in similar accuracy and precision. The Encephalopathy vs. No Encephalopathy run was significantly worse than the other two.

Table \ref{tab:pfe} shows the results for the same setup as 
in Table \ref{tab:comp} after we performed feature engineering. For both representations in Table \ref{tab:comp}, the recall was significantly lower than the precision, indicating that there were many more false negatives than false positives. We thus engineered the features in the training set. Certain features in the training set were most strongly present in positive patients, and as such were correlated with positive examples. Some examples in the test set had slightly lower values for those features, and were thus labelled by the classifier as negative examples, when in reality they were positive. We aimed to influence some of these borderline false negative predictions by lowering the threshold for those positively correlated features that would lead to an example being predicted as positive. In order to reduce the threshold for those positively correlated features, we applied a multiplier of 0.9 to the values of each feature of every positive patient in the training set. This had the effect of causing some previously borderline negative predictions to become positive predictions, primarily reducing the number of false negatives and improving the recall scores, but also positively impacting most other metrics as well.  Attempts to use a multiplier of 
less than 0.9 did not appear to further 
improve the results.

With the multiplier of 0.9 applied to the 
feature values for positive training examples, the recall 
increased by 4.09\% for UCLUST and by 12.25\% for Kraken. This was confirmed in the 
cross validation results as well, 
in which UCLUST's recall increased by 2.18\% and 
Kraken's recall increased by 11.41\%. Because of the improvement 
in recall, both representations also had similarly improved F1-Scores after feature engineering. Accuracy and precision remained generally consistent. 
%
Kraken's LOOCV precision 
increased by 5.55\%. This suggests that Kraken's very high precision was tied to its lack of many positive phenotype predictions in general, resulting in a very low number of false positives. While Kraken improved more than UCLUST did due to feature engineering, UCLUST still performed better in all metrics.

Tables \ref{tab:pairuclustfe} and \ref{tab:pairkrakenfe} display the results after the same feature engineering techniques were applied to the pairwise results in \ref{tab:pairuclust} and \ref{tab:pairkraken}, respectively. UCLUST's No Encephalopathy vs. Control run saw an increase of 4.06\% in accuracy, 3.75\% in precision, and similar gains in its cross validation results for accuracy and precision. UCLUST's other two runs were largely unaffected, albeit with slightly better cross validation results. Kraken's Encephalopathy vs. Control run was greatly improved after feature engineering, with an 8.20\% increase in accuracy and a 9.75\% increase in precision, with similar increases in cross validation results. This more than offset a slight decrease in recall. The Encephalopathy vs. No Encephalopathy run had mostly unchanged results, albeit with much improved cross validation scores, and the No Encephalopathy vs. Control run was not significantly affected.

### 2016-07-25 latex/input/intro.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

%Intro

The human body is one of the 
most densely populated microbial 
environments in the world, with human host cells interacting 
with  more than $10^{14}$ microbial cells. Collectively, the human 
and microbial cells are referred to as the 
\emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}. 
%
%
%
Biotechnological  advances 
can 
interrogate the microbial 
communities present across three primary dimensions: (i) clinically different
patients like those suffering from liver disease versus healthy controls (ii) different
patient sites and tissues like skin and the gastrointestinal tract 
and (iii) across different patient visits~\cite{Costello12182009,Qin:2010fk} and before/after treatment. 
%
Methods to query this  microbial
data include DNA sequencing of 
specific marker genes such as the 16S rRNA marker genes~\cite{Woese97}, 
sequencing
the entire pool of microbial genomes at once (metagenomics), or 
sequencing the mixture of 
expressed gene transcripts (metatranscriptomes).
%
These sequencing methods along 
with proteomics and mass spectrometry
allow us to study the
biotransformations caused by the microbial communities. 
%
Several researchers and clinicians have embarked on
studies of the pathogenic and clinical 
role played by the microbiome with respect to
human health and disease conditions.

In this study, we specifically developed
a computational method to predict the clinical phenotype
of a patient i.e., a pipeline to 
predict whether or not the patient has a specific type and level of  a disease.  This 
pipeline
is able to make the phenotypic
predictions from input metagenomic sequences using post-processing 
with unsupervised binning approaches (clustering) or  supervised search based 
taxa profiling approaches.  This method is designed to 
run quickly and produce accurate and sensitive 
results. We  compare and evaluate two state-of-the-art 
approaches, one for binning (UCLUST \cite{Edgar10}) and one for taxonomic profiling (KRAKEN \cite{Wood14}) 
to represent the input metagenome sequences as features for classification. The classification
formulation uses a binary support vector machine based classifier \cite{vap95}. We also 
evaluate the performance of preliminary 
feature scaling and engineering  for classification purposes.

We specifically evaluated our approach on 
microbiome samples of patients suffering from hepatic encephalopathy due 
to liver cirrhosis. 
%
We found that 
UCLUST performed better than Kraken on our data 
set, although Kraken was faster. We were also able to 
engineer the data to improve the accuracy of the classifier. Our 
pipeline had 85.64\% accuracy using UCLUST and 80.66\% accuracy 
using Kraken.

%For our data set with 1.3 million reads, UCLUST took 
%180 seconds to run and Kraken took 120 seconds. 
%

### 2016-07-25 latex/input/materials.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

\subsection{Dataset Description}

We obtained 904
clinical microbiome samples from a study by Dr. Patrick Gillevet 
that relates to patients who suffered from hepatic encephalopathy due 
to liver cirrhosis. Specifically, the classification formulation 
was setup to distinguish between patients suffering with a specific 
clinical phenotype or not. There were a total of 239 
patients with Encephalopathy   due to liver cirrhosis (denoted 
by the ``Encephalopathy'' class), 590 patients with liver cirrhosis 
but no Encephalopathy (denoted by ``No Encephalopathy'') 
and 75 patients who were considered as 
control and did not have either of the clinical conditions (denoted by ``Control'').
%
16S rRNA metagenomic sequence read data was 
obtained from patient stool samples. On average 
there were a total of 1,464  sequence reads of length 200-400 obtained per 
patient, with 1,323,016 total reads across all patients in our dataset.

\subsection{Evaluation Metrics}
We assess the performance of our classification pipeline 
in terms of correctness and execution time.
%
Given the imbalanced nature of class distributions,  the 
performance of  binary classifiers was measured by 
F1 score, precision and recall, along with
the standard accuracy metric. We discuss 
these standard metrics in brief.

Accuracy measures the percentage of instances 
that are classified correctly and can 
be represented by 
\begin{equation}
Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
\end{equation}
where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.

Accuracy as an evaluation metric can be biased if one of the classes 
(positive or negative)  has a larger number of examples than the other.   
Precision measures the percentage of positive predictions that 
were correct, 
whereas recall measures the percentage of positive 
examples that were correctly predicted (or retrieved). 
We can represent Precision and Recall by \cite{Goutte}:

\begin{equation}
Precision = TP / (TP + FP). \label{eqn:prec}
\end{equation}
\begin{equation}
Recall = TP / (TP +FN). \label{eqn:roc}
\end{equation}

The F1 score captures the trade offs between precision and recall in a
single metric and is the harmonic mean of precision and recall \cite{Goutte}, given 
by:
\begin{equation}
F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
\end{equation}

\subsection{Software and Hardware Details}
We used the Argo computing cluster available at George Mason University. 
%
The representation generation phase using UCLUST and Kraken and and the classification phase
were 
run on one of the  compute nodes available on the cluster. The cluster is configured with 
35 Dell C8220 Compute Nodes, each with dual 
Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores – 
528 and 1056 total threads, RAM \textgreater 2TB)\cite{ORC}.

Source codes for UCLUST \cite{Edgar10} and KRAKEN \cite{Wood14} were downloaded from 
their respective websites\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html     
Kraken: https://ccb.jhu.edu/software/kraken/} and compiled on the Argo platform. 
%
Kraken aligns reads to an OTU database. The standard Kraken database is 160GB in size. Due to 
computational limitations we used the custom, small sized 4GB
database  in this feasibility study \cite{Kraken}.

For the  SVM-based classification\cite{Joachims08}  we used the popular 
SVM-Light \cite{Joachims08}  source code, which is publicly 
available \footnote{http://svmlight.joachims.org/}.  The linear kernel was used
and the regularization parameters were set to their default values.

\subsection{Experimental Protocol}

For evaluating the performance of our binary phenotypic classifiers, we 
split the patient samples into a training set containing 80\% of the 
patients and the test set containing 20\% of the patient samples. Every fifth sample was placed in the test set, so the relative class sizes were roughly consistent between the training and test sets.
We also performed leave-one-out cross validation (LOOCV).

In the next section, we discuss the accuracy of our 
classifier with regards to predicting the clinical 
phenotype of a given patient using 
either the unsupervised 
clustering representation with UCLUST or 
supervised OTU representation 
with Kraken. We present results for classifiers 
distinguishing patients  in the ``Encephalopathy'' class versus
the other classes, effectively 
determining whether or not a patient had Encephalopathy 
regardless of whether they had liver cirrhosis. We also 
present pairwise one-versus-one classification results that 
compared two phenotypes within the dataset, i.e.  Encephalopathy versus
No Encephalopathy, Encephalopathy versus
Control and No Encephalopathy versus Control.

### 2016-07-25 latex/input/methods.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

%methods

\subsection{Overview}

Figure \ref{fig1} provides an overview of the 
developed computational pipeline. 
%
We obtain 16S metagenomic sequence reads from patients who 
had encephalopathy and liver cirrhosis, no encephalopathy but liver cirrhosis, and controls (with neither phenotype). The input sequence 
reads are annotated using a taxonomic profiling approach (Kraken) or 
an unsupervised binning approach (UCLUST). The output of this annotation phase 
features are generated, which are passed on the binary support vector machine 
classifier (SVMs).  We used the  svm-light implementation \cite{Joachims08}  
to perform the classification of whether or not a patient 
had a clinical phenotype. We compared the classification performance of the 
SVM classifier with the two intermediate sequence representations obtained from 
UCLUST and Kraken. We also applied feature scaling and 
selection techniques to improve the classification results.

\begin{figure*}[t]
\begin{center}
\caption{Pipeline Overview. \label{fig1}}
\includegraphics[scale=0.7]{./diagrams/concept}
\end{center}
\end{figure*}

\subsection{Unsupervised Clustering using UCLUST}
\label{uclustmethods}

In this work we evaluate the use of an unsupervised approach 
for representing our input metagenome sequence reads. Specifically, we 
  use UCLUST \cite{edgar2010search} due to its superior performance in terms 
  of run time and accuracy.
%
%
%
UCLUST \cite{edgar2010search} follows a greedy, iterative clustering approach. As a first 
step, this approach identifies exact matches of fixed length between sequence pairs known 
as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
the aligned pairs. Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and 
used for further processing. As such, the step eliminates a lot of pairwise comparisons.

Then UCLUST follows an 
incremental approach for assigning the input sequences to the different bins. 
%
%
The clustering solution is initialized as an empty list. Sequences are then 
incrementally added to the clusters existing within the list. 
%
Each unassigned input sequence is compared to the cluster representatives within the list using 
the fast indexing search technique that uses the HSPs. If a match is found with one of the 
cluster representatives, then that sequence is assigned to that particular cluster or the 
input sequence forms a new cluster. 
%
UCLUST ensures
that the cluster representatives are sequences with the largest length.

\subsection{Supervised Taxonomic Identification using Kraken}

We also evaluate the use of a search based 
approach to taxonomically label our sequence reads. We use 
Kraken for this, due to its strong performance in terms of run time 
and precision. Kraken finds $k$-mers in an input 
sequence read and queries a database of OTUs. It then associates the 
read with the lowest common ancestor (LCA) of the genomes that contains $k$-mers from the read \cite{Kraken}. In order to reduce 
the number of false positives, Kraken does not classify reads for 
which it does not find sufficient evidence of any match.

This approach is designed to perform quickly and with 
high precision. However, it also involves downloading and building 
a large database to query. It is often infeasible to download and 
build the database in its entirety, so a smaller custom  
subset of the database is available for testing 
purposes. We used this 
subset of the entire database in this study (referred to as kraken-light).

\subsection{Classification using Support Vector Machines (SVMs)}

Support Vector Machines (SVMs) \cite{vap95} are 
one of the most powerful and versatile binary classifiers used 
in myriad applications. 
%
For a  binary classification problem, the support 
vector machine (SVM) \cite{vap95} framework 
maximizes the separation between the two classes.
%

Given an input set of positive training instances and 
negative training instances, the SVM classifier learns a 
mathematical function $f$  that maximizes the separation between 
the given positive and negative training instances. The function $f$ will then be able to predict whether a new instance (considered a test instance) should have a positive or negative class label.
%
For our formulation we use SVMs to make clinical phenotype prediction for each 
patient sample 
based on their  input metagenome sequence samples represented with their 
clustering or taxa membership features.

For all the available microbiome samples,  we have a set of reads processed via UCLUST or Kraken. 
%
Each patient is associated with a vector that is used by the SVM to help build $f$. Each row in the vector, called a feature, corresponds to a certain species, and the value for that row corresponds to how closely the read from the patient matched what the species was expected to look like. The exact process of assembling a feature vector for a 
patient depends on which data representation is being used. For 
UCLUST, we use each cluster seed 
as a feature. For each input 
sequence, we use the percentage match to its cluster seed as 
the value for that feature and average the 
sequence match if multiple reads from the metagenome samples 
are assigned to the same cluster by UCLUST. 
%
For Kraken, we use the taxonomic label associated 
with the read as the feature for the SVM. The percentage match of the 
read to that label is used as the feature value. Similar 
to UCLUST, we average the similarity score if multiple reads align to the 
same taxonomic class label. 
%
Our feature selection process involves discarding Kraken's unlabeled 
reads and UCLUST's clusters with only one read (singletons).

We used cross-validation and a held out set procedure (discussed in the next 
Section) to assess the performance of SVM classification using UCLUST and Kraken
as intermediate representations.

### 2016-07-25 latex/input/related.tex
completed fixes of gicf with mini batch size randomization, proper dot products, top k instances, randomized weight initialization; added temporary latex template for paper

%RELATED

\subsection{Human Microbiome and Metagenomics}

The combination of human host cells and microbial 
cells that govern several facets of human health and pathology are referred 
to as collectively by the term \emph{human microbiome} \cite{hugenholtz2008microbiology}. 
%
Using today's sequencing technologies we have the capacity to determine 
the DNA sequences of these co-existing microbial communities. 
However, 
current  genomic  technologies 
do not provide the complete genome for each individual microbe, but 
  short, contiguous subsequences from random positions of the genome. These 
  short subsequences are referred by ``reads''. 
%
Metagenome sequence assembly is defined as the process of taking the different 
sequence reads from the various microbial organisms to produce long contiguous 
sequence of DNA for each individual organism within the community mixture. 
%
The assembly process involves identifying overlapping parts of different reads 
to order them and separate them in organism-specific larger sequences.
%

The metagenome assembly problem is known to be challenging due 
  to the similarity of genomes from the different microbes, 
differing 
  abundance, diversity, complexity and varying
  genome lengths of never-sequenced before microbes within 
  the different microbiome samples. DNA sequencing machines  
  are very high throughput and produce Terabytes of data per run and 
  also produce reads with poor quality  \cite{hugenholtz2008microbiology}. Several studies 
  have highlighted the challenges associated with the metagenome analysis and assembly problem by performing a simulation study to determine the feasibility of solving the metagenome assembly problem \cite{charuvaka2011evaluation}.

  %
  Targeted metagenomics or 16S rRNA gene sequencing provides
  a
  first step for the quick and accurate characterization of
  microbial communities. 16S sequences are marker genes, which
  exists in most microbial genomes and
  have a conserved portion
  for detection (primer development) and a variable portion that allows for
    categorization within  different  taxonomic groups \cite{petrosino2009metagenomic}.
        %
        Targeted metagenomics are also effective in detecting species with low 
        abundances. However, they may not be good at discovering 
        unique species (orphans) that have never been sequenced before. % ones that are considered to be orphan (i.e., never sequenced before).

\subsection{Microbiome Informatics}

Since the human microbiome project \footnote{http://hmpdacc.org/} and 
release of publicly available metagenomic datasets began, a host 
of methods have been developed for the analysis of 
assembled metagenomes and  input  sequence reads (before assembly)
obtained from 16S rRNA genes and whole metagenomes.  
%
%
The relationship between microbiome and human health can be characterized by first
identifying the content, abundance, and functionality of the microbes within the samples. Several
computational approaches (surveyed here \cite{:mo}) have been developed for handling the vast
amount of metagenomic data to solve two related problems: (i) clustering or binning, and (ii) taxonomy profiling methods.
%

\subsubsection{Binning Methods}

The ``binning'' problem  involves grouping input short reads  such that reads within a group are 
similar to each other. This process may lead to groups that are organism-specific and is 
unsupervised in nature, being referred by clustering in the data mining community. 
%
The binning process does not attempt to provide an automated labeling of the input reads. 
%
The clusters/bins/groups obtained from an input metagenome sample 
is referred to as  Operational Taxonomic Units (OTUs), and the number of 
OTUs gives an approximation of species diversity in a sample \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
%
These  approaches are not constrained due to the absence of a complete coverage in taxonomic databases. Some 
environmental samples contain microbial organisms that have never been cultured in a laboratory, and thus those organisms do not exist in genomic databases.
As such,  binning of 
sequence reads has several advantages: (i)
        it can lead to an improved  metagenome assembly, (ii) it can be used 
        for computing species diversity metrics \cite{bibm2012} 
        and (iii)
        the  reduced computational
        complexity within several work-flows that analyze only
        cluster representatives, instead of individual sequences
        within a sample.

CD-HIT \cite{Li01072006}, UCLUST \cite{Edgar10}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a} and 
MC-LSH \cite{bibm2012}
are some of the popular 
metagenome/sequence clustering approaches used for binning. 
%
UCLUST, MC-MinH and MC-LSH  are
greedy approaches that achieve computational efficiency 
by using either hash-based indexing or matching of gap-less sequences called seeds (instead 
of expensive sequence alignment) and followup with an incremental clustering approach 
that does not involve comparing all pairs of input sequences. In Section \ref{uclustmethods} we discuss 
in detail the UCLUST \cite{Edgar10}  sequence clustering approach. We use UCLUST within 
our study, known to be a state-of-the-art metagenome clustering approach in terms of 
computational performance and accuracy of results.

\subsubsection{Taxonomy Profiling Methods}
\label{sec:tax}

The taxonomy profiling problem involves 
assigning a specific label (i.e., a phylogenetic group label) to sequence 
reads or assembled metagenome contigs~\cite{phymm}. A traditional 
approach for taxonomy profiling is to formulate  a 
classification problem (supervised) that uses marker genes for
  identification of source organism of a sequence read or fragment~\cite{Woese97}. Marker
  genes are highly conserved and provide accurate identification
  of the taxonomy class~\cite{ZongzhiLiu10012008}. These approaches 
  rely on a previously annotated 
  reference dataset like RDP~\cite{Cole2005,Cole2007,Cole2008} or GreenGenes~\cite{greengenes}. These methods
  provide
  valuable community estimates but are limited to specific 
  reads or contigs (marker genes constitute a small fraction of a metagenomic
  sequence set) and have low sensitivity due to reliance
  on an incomplete and taxon-biased reference genome database.The RDP database
  implements a naive Bayes classifier 
  that uses DNA-composition features to classify 16S RNA sequence reads into taxonomic classes as defined
  by Bergey's taxonomy~\cite{ZongzhiLiu10012008}.

  Several comparative methods have been developed for the assignment of phylogenetic class based on principles of homology. Such methods  align
  reads or contigs using BLAST~\cite{altschul90a}, and assign taxonomy based on the best match with a reference database~\cite{Tringe04222005}. MEGAN~\cite{megan} and MARTA~\cite{marta} are
  metagenomic analysis and visualization programs that  make the assignment based on multiple BLAST hits and optimized parameters.  The MG-RAST~\cite{mgrast}  web server
  provides taxonomic and functional annotation by comparative searches performed across multiple reference databases. GAAS~\cite{gaas} is a novel BLAST-based tool
  that includes genome length normalization along with a similarity weighting for multiple BLAST hits to provide improved estimates.

  Composition-based methods have also been developed. They extract key sequence features such as GC composition and $k$-mer frequencies and build supervised
  classification models using those features. PhyloPythia~\cite{McHardy2007} uses a support vector machine framework~\cite{vap95} to classify long reads into taxonomic groups
  using a $k$-mer based kernel function. TETRA~\cite{tetra} correlates the $k$-mer pattern feature to different taxonomic groups. Kraken, which is used in this paper, searches a taxonomic database for the lowest common ancestor (LCA) of the genomes that contain $k$-mers from a sequence read. Phymm~\cite{phymm} trains an interpolated Markov model
  to characterize variable length subsequences  specific to different taxonomic subgroups. In combination with BLAST, Phymm shows improved classification
  accuracy for short reads of 100 base pair (bp) length. Such Markovian models have been very successful in gene finding algorithms like Glimmer~\cite{delc98nucl}.

### 2016-07-25 bak/gicf.py

@@ -4,7 +4,6 @@
 
 import argparse
 import math
-import numpy as np
 import random
 import sys
 import time
@@ -14,7 +13,7 @@ def similarity(i, j):   # similarity measure (0 to 1) between instances i and j
     norm = 0.0    # the squared 2-norm of the difference between the two instance vectors, computed in for loop below
     for k in range(len(i)):
         norm += (i[k] - j[k]) ** 2
-    return math.exp(-norm)  # / len(i) / len(j)      # returns value for kernel described in GICF paper
+    return math.exp(-norm)      # returns value for kernel described in GICF paper
 
 
 def instance_penalty(i, j):     # non-negative penalty on the difference between predictions for instances i and j
@@ -22,23 +21,20 @@ def instance_penalty(i, j):     # non-negative penalty on the difference between
 
 
 def logistic(w, x):     # implementation of the classifier for the instances; currently logistic regression
-    dot = w[0]     # dot product between weights and x, computed in for loop below
+    dot = 0.0     # dot product between weights and x, computed in for loop below
     for i in range(len(x)):
-        dot += w[i+1]*x[i]
-    if dot > 709:    # prevent math overflows
+        dot += w[i]*x[i]
+    if dot > 20:    # prevent math overflows
         return 1
-    if dot < -709:   # prevent math overflows
-        return -1
-    # print 1 / (1 + math.exp(-dot))
-    return 1 / (1 + math.exp(-dot))
-    # print ((1 / (1 + math.exp(-dot))) - 0.5) * 2
-    # return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return logistic regression, except from -1 to 1
+    if dot < -20:   # prevent math overflows
+        return 0
+    return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return normalized logistic regression formula using dot
 
 
 def derivative_logistic(w, x):   # derivative of logistic regression function
-    dot = w[0]     # dot product between weights and x, computed in for loop below
+    dot = 0.0     # dot product between weights and x, computed in for loop below
     for i in range(len(x)):
-        dot += w[i+1] * x[i]
+        dot += w[i]*x[i]
     if dot > 20 or dot < -20:    # prevent math overflows
         return 0
     e = math.exp(dot)   # e ^ (dot product of w and x)
@@ -48,17 +44,6 @@ def derivative_logistic(w, x):   # derivative of logistic regression function
     norm = math.sqrt(norm)
[…]

### 2016-07-25 gicf.py

@@ -4,6 +4,7 @@
 
 import argparse
 import math
+import numpy as np
 import random
 import sys
 import time
@@ -13,7 +14,7 @@ def similarity(i, j):   # similarity measure (0 to 1) between instances i and j
     norm = 0.0    # the squared 2-norm of the difference between the two instance vectors, computed in for loop below
     for k in range(len(i)):
         norm += (i[k] - j[k]) ** 2
-    return math.exp(-norm)      # returns value for kernel described in GICF paper
+    return math.exp(-norm)  # / len(i) / len(j)      # returns value for kernel described in GICF paper
 
 
 def instance_penalty(i, j):     # non-negative penalty on the difference between predictions for instances i and j
@@ -21,20 +22,23 @@ def instance_penalty(i, j):     # non-negative penalty on the difference between
 
 
 def logistic(w, x):     # implementation of the classifier for the instances; currently logistic regression
-    dot = 0.0     # dot product between weights and x, computed in for loop below
+    dot = w[0]     # dot product between weights and x, computed in for loop below
     for i in range(len(x)):
-        dot += w[i]*x[i]
-    if dot > 20:    # prevent math overflows
+        dot += w[i+1]*x[i]
+    if dot > 709:    # prevent math overflows
         return 1
-    if dot < -20:   # prevent math overflows
-        return 0
-    return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return normalized logistic regression formula using dot
+    if dot < -709:   # prevent math overflows
+        return -1
+    # print 1 / (1 + math.exp(-dot))
+    return 1 / (1 + math.exp(-dot))
+    # print ((1 / (1 + math.exp(-dot))) - 0.5) * 2
+    # return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return logistic regression, except from -1 to 1
 
 
 def derivative_logistic(w, x):   # derivative of logistic regression function
-    dot = 0.0     # dot product between weights and x, computed in for loop below
+    dot = w[0]     # dot product between weights and x, computed in for loop below
     for i in range(len(x)):
-        dot += w[i]*x[i]
+        dot += w[i+1] * x[i]
     if dot > 20 or dot < -20:    # prevent math overflows
         return 0
     e = math.exp(dot)   # e ^ (dot product of w and x)
@@ -44,6 +48,17 @@ def derivative_logistic(w, x):   # derivative of logistic regression function
     norm = math.sqrt(norm)
[…]

### 2016-07-25 cirrhosisTests/gicf.py
created script to limit amount of reads per cluster per patient in a file

@@ -339,9 +339,9 @@ def grid_search(verbose, w, data, train, test, res, limit):
 
     rate = [0.0001]  # , 0.001]  # , 0.01, 0.1]
     l = [float(num)]  # [float(num) * 0.1, float(num), float(num) * 10]
-    top_k = [0.9]
-    batch_size = [0.01]  # [num * 4]  # [num * 4, num * 10, num * 100]
-    iterations = [3]  # , 10, 100, 1000]
+    top_k = [0.8]
+    batch_size = [0.1]  # [num * 4]  # [num * 4, num * 10, num * 100]
+    iterations = [30]  # , 10, 100, 1000]
     accuracy = -1.0             # best accuracy on test data
     parameters = [0, 0, 0, 0, 0]   # best parameter settings
     results = []                # best results

### 2016-07-25 clusterLimiter.py

@@ -1,0 +1,78 @@
+# Author: Nathan LaPierre
+# Date: July 25, 2016
+
+import argparse
+import os
+import sys
+import time
+
+
+def parseargs():    # handle user arguments
+    parser = argparse.ArgumentParser(description='Limit the number of reads per cluster in patient files.')
+    parser.add_argument('--fasta', default='NONE', help='Location of input FASTA file. Required.')
+    parser.add_argument('--uclust', default='NONE', help='Location of input UCLUST file. Required.')
+    parser.add_argument('--limit', default=-1, help='Limit the number of selections per cluster. Default: 1.')
+    parser.add_argument('--output', default='limited', help='Files to write lines to. Default: limited{.fasta/.uc}')
+    parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
+    args = parser.parse_args()
+    return args
+
+
+def main():
+    start = time.time()
+    args = parseargs()
+    if args.fasta == 'NONE' or args.uclust == 'NONE':
+        print "Location of input FASTA and UCLUST files with data must be specified with --fasta and --uclust."
+        sys.exit()
+    try:
+        limit = int(args.limit)
+    except ValueError:
+        print "Error: Limit must be an integer greater than 0."
+        sys.exit()
+    if limit < 1:
+        print "Error: Limit must be an integer greater than 0."
+        sys.exit()
+
+    outfasta = open(args.output + '.fasta', 'w')
+    outuclust = open(args.output + '.uc', 'w')
+    cluster_counts = {}     # keeps track of counts of reads from each cluster for each patient
+    try:
+        infasta = open(args.fasta, 'r')
+        inuclust = open(args.uclust, 'r')
+        for line in inuclust:
+            if line.startswith('H') or line.startswith('S'):
+                fastaline = infasta.next()
+                fields = line.split('\t')       # fields in clustering output lines
+                tag = fields[8].split(' ')[0]   # gets the tag that matches a patient
+                if tag not in cluster_counts:
+                    cluster_counts[tag] = [0]
+                counts = cluster_counts[tag]
+                clustnum = int(fields[1])    # cluster number
+                while len(counts) < clustnum + 1:
+                    counts.append(0)
+                if counts[clustnum] < limit:
+                    counts[clustnum] += 1
+                    outfasta.write(fastaline)
+                    fastaline = infasta.next()
+                    outfasta.write(fastaline)
+                    outuclust.write(line)
[…]

### 2016-07-28 cirrhosisTests/gicf.py
cleaned up gicf.py file

@@ -29,36 +29,7 @@ def logistic(w, x):     # implementation of the classifier for the instances; cu
         return 1
     if dot < -709:   # prevent math overflows
         return -1
-    # print 1 / (1 + math.exp(-dot))
     return 1 / (1 + math.exp(-dot))
-    # print ((1 / (1 + math.exp(-dot))) - 0.5) * 2
-    # return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return logistic regression, except from -1 to 1
-
-
-def derivative_logistic(w, x):   # derivative of logistic regression function
-    dot = w[0]     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i+1] * x[i]
-    if dot > 20 or dot < -20:    # prevent math overflows
-        return 0
-    e = math.exp(dot)   # e ^ (dot product of w and x)
-    norm = 0.0        # norm of vector x times scalar e, calculated below
-    for k in range(len(x)):
-        norm += (x[k] * e) ** 2
-    norm = math.sqrt(norm)
-    value = math.fabs(norm / ((1 + e) ** 2))    # value for derivative of logistic function
-    return ((1 / (value + 1)) - 0.5) * 2    # normalized result
-
-
-def validate_logistic(w, x):
-    dot = w[0]     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i+1]*x[i]
-    if dot > 20:    # prevent math overflows
-        return 1
-    if dot < -20:   # prevent math overflows
-        return 0
-    return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return logistic regression, except from -1 to 1
 
 
 def label_penalty(predicted, actual):   # non-negative penalty on difference between prediction & true label for group
@@ -73,18 +44,6 @@ def predict_label(group, instance_labels):
     return total / len(group)   # return average label value in the group
 
 
-'''
-def predict_label (group, instance_labels):  # predict label, top-K version
-    # k = 0.2
-    k = 0.2
-    total = 0.0
-    x = sort (instance_labels).desc()
-    for i in range(0.2 * len (instance_labels)):
-        total += instance_labels[i]
-    return total / (0.2 * len (instance_labels))
-'''
-
-
 def d_group_penalty(group, instance_labels, d_instance_labels, group_label):
     # calculate group penalty in derivative of cost function
[…]

### 2016-07-28 gicf.py

@@ -29,36 +29,7 @@ def logistic(w, x):     # implementation of the classifier for the instances; cu
         return 1
     if dot < -709:   # prevent math overflows
         return -1
-    # print 1 / (1 + math.exp(-dot))
     return 1 / (1 + math.exp(-dot))
-    # print ((1 / (1 + math.exp(-dot))) - 0.5) * 2
-    # return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return logistic regression, except from -1 to 1
-
-
-def derivative_logistic(w, x):   # derivative of logistic regression function
-    dot = w[0]     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i+1] * x[i]
-    if dot > 20 or dot < -20:    # prevent math overflows
-        return 0
-    e = math.exp(dot)   # e ^ (dot product of w and x)
-    norm = 0.0        # norm of vector x times scalar e, calculated below
-    for k in range(len(x)):
-        norm += (x[k] * e) ** 2
-    norm = math.sqrt(norm)
-    value = math.fabs(norm / ((1 + e) ** 2))    # value for derivative of logistic function
-    return ((1 / (value + 1)) - 0.5) * 2    # normalized result
-
-
-def validate_logistic(w, x):
-    dot = w[0]     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i+1]*x[i]
-    if dot > 20:    # prevent math overflows
-        return 1
-    if dot < -20:   # prevent math overflows
-        return 0
-    return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return logistic regression, except from -1 to 1
 
 
 def label_penalty(predicted, actual):   # non-negative penalty on difference between prediction & true label for group
@@ -73,18 +44,6 @@ def predict_label(group, instance_labels):
     return total / len(group)   # return average label value in the group
 
 
-'''
-def predict_label (group, instance_labels):  # predict label, top-K version
-    # k = 0.2
-    k = 0.2
-    total = 0.0
-    x = sort (instance_labels).desc()
-    for i in range(0.2 * len (instance_labels)):
-        total += instance_labels[i]
-    return total / (0.2 * len (instance_labels))
-'''
-
-
 def d_group_penalty(group, instance_labels, d_instance_labels, group_label):
     # calculate group penalty in derivative of cost function
[…]

### 2016-08-03 latex/input/abstract.tex
added script for generating plots of GICF runs; updated paper

@@
 
-We demonstrate a computational method to predict the clinical
-phenotypes of a patient from raw metagenomic sequence read data. We  compared 
-two state of the art programs for annotating the sequence data, 
-UCLUST and Kraken, and using their output for feature generation. 
-We apply these programs 
-to a set of over 1.3 million reads from 
-904 patients, some of whom have liver cirrhosis, encephalopathy due to liver cirrhosis, or 
-neither disease. Once the reads have been processed by UCLUST or 
-Kraken, we use  Support Vector Machines to setup the 
-clinical phenotype prediction problem.  
-%
-We find that too many false negatives are being predicted by 
-the classifier. In order to address the issue, we scale features to
-improve the classification model and evaluate the 
-end results on a held-out test set. We
-demonstrate our approach works quickly and 
-accurately with an 85.64\% success rate when we use the 
-UCLUST representation.
-We also find that UCLUST generally performs better than Kraken, with the latter having an 80.66\% success rate.
-We also test our classifier on several subsets of the data, with success rates ranging from 69.81\% to 96.72\%.
+The recent advent of Metagenome-Wide Association Studies (MGWAS) has allowed for increased accuracy in the prediction of patient phenotype (disease), but has also presented big data challenges.
+Meanwhile, Multiple Instance Learning (MIL) is useful in the domain of bioinformatics because, in addition to classifying patient phenotype, it can also identify individual parts of the metagenome that are indicative of that phenotype, leading to better understanding of the disease.
+We demonstrate a novel, efficient, and effective MIL-based computational pipeline to predict patient phenotype from MGWAS data.
+Specifically, we use a Distance-based Bag of Words method, which has been shown to be one of the most effective and efficient MIL methods.
+This involves assembly of the metagenomic data, clustering of the assembled contigs, extracting features from the contigs, and using a standard SVM classifier to predict patient labels and identify the most relevant read clusters.
+With the exception of the given labels for the patients, this entire process is \emph{de novo}.
+We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper.

### 2016-08-03 latex/input/conclusion.tex
added script for generating plots of GICF runs; updated paper

@@
 
-We have demonstrated a computational method 
-to predict clinical phenotypes using metagenomic sequence data with intermediate 
-OTU representation using state-of-the-art binning and taxonomic profiling approaches. 
-%
-%
-Our results showed that 
-the UCLUST representation 
-achieved 85.64\% accuracy and 70.46\% F1-Score. Kraken performed less well, with 80.66\% accuracy and 53.34\% F1-Score. 
-For the 1.3 million sequence reads UCLUST ran in 180 seconds and Kraken for 120 seconds, indicating that both performed efficiently enough to feasibly use them with larger data sets.
-
-We then used UCLUST to test three one-versus-one clinical phenotype classifiers,
-which generally produced stronger results than our original 
-comparison of the "Encephalopathy" class versus all other 
-classes. In particular, the classification of 
-"Encephalopathy" versus "Control" performed the best. This intuitively seems reasonable. Whereas 
-the original comparison requires making distinctions between 
-those who have both encephalopathy and liver cirrhosis, those who have only the latter, and those who have neither, this pairwise comparison focuses on the simpler distinction 
-between patients who have both diseases and those who have neither.
-The other two pairwise comparisons also similarly simplify the classification problem.
-When Kraken was used for the three pairwise comparisons, it had the peculiar effect of drastically increasing the recall while maintaining roughly the same accuracy and precision.
-This is most likely due to the fact that the pairwise comparisons involved data sets that were less negatively-skewed and thus didn't cause the classifier to predict as negatively based on Kraken's results.
-This is combined with the relatively strong accuracy and precision numbers in Kraken's original run, versus its poor recall numbers that left plenty of room for improvement.
-
-Our technique of applying a multiplier to feature values for positive examples produced the effect of improving the recall in most cases.
-These results were generally corroborated by similar improvements in the cross validation results.
-Another effect was that precision and accuracy results were also often improved after applying this technique.
-This intuitively makes sense, as switching false negatives to true positives also positively affects the accuracy and precision ratios.
-Overall, this feature engineering process appeared to have a generally positive effect on most of the results, suggesting that similar techniques for improving classification results merit further investigation.
-
-Based on our results demonstrating over 85\% accuracy for the original run and even better results for some of the pairwise comparisons, we believe the use of machine learning
-classification techniques to predict clinical phenotypes 
-merits continued research. There are many potential practical 
-uses of such technology for medical purposes, including 
-diagnostics and research into diseases and their 
-relationship with the human microbiome.
-
-\subsection{Future Work}
-
-Kraken is developed to be fast, accurate, and precise, all 
-of which we found to be true. However, it was not as accurate as a representation in comparison to 
-UCLUST.
-Kraken's poor performance was mostly due to  poor matching of $k$-mers in the reads to taxonomic classes in its database.
-Often, a read was only a one or two percent match to the class that Kraken labeled it as.
-This
-may be due in part to database constraints. Kraken's full database is very computationally expensive to download and build. We 
-used a 4 GB custom subset of the full 160 GB database. It is 
-possible that the full database would have led to improved performance, and this is 
-a future research question.
-
-There is also an opportunity to use a much larger set of sequence reads.
-Having more training examples for the classifier would  lead to a stronger and more generalizable model for predicting clinical phenotypes.
-The use of other annotation and classification methods could also be explored.
-Finally, a limited set of feature selection and scaling techniques were explored, and future research could investigate further how such techniques could be used to improve predictions.
+We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data.
+We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study.
+Finally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods.
+Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level labels and validate those labels against clinically verified labels.

### 2016-08-03 latex/input/experiments.tex
added script for generating plots of GICF runs; updated paper

@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.5]{./diagrams/PreFeatEng}
-\caption{This diagram shows the results for the comparison between UCLUST and Kraken when classifying "Encephalopathy" versus all other classes.
-UCLUST had an accuracy and F1-Score of 84.53\% and 67.44\% respectively.
-Kraken didn't perform as well, with an accuracy of 79.56\% and an F1-Score of 43.08\%.
-\label{prefeature}}
-\end{figure}
+\subsection{Data, Software, and Hardware Used}
 
-\begin{table*}[t]
-\begin{center}
-\caption{Comparative Run Time of UCLUST and Kraken Representations. \label{tab:time}}
-\begin{tabular}{|c|c|}\hline
-Representation & Time\\\hline
-UCLUST & 180s\\\hline
-Kraken & 120s\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+Our main dataset comes from a GWAS of Type 2 Diabetes (T2D) in Chinese adults, published by Qin et al \cite{qin041012}.
+The data is publicly available on the National Center for Biotechnology Information (NCBI) \cite{ncbi} website using the accession number SRP011011.
+The full dataset contains over 200 patients, but for our study we used a subset of 60 patients due to lengthy download and format conversion times.
+Of these 60 patients, 30 had T2D and 30 were controls.
+For the purpose of testing our cluster sampling methods, we also used a smaller dataset obtained from Dr.
+Patrick Gillevet of GMU, which is not publicly available.
+It contains 904 patients, 239 of whom have encephalopathy and 665 of whom are controls, but is much smaller because it uses 16S rRNA instead of whole genome data.
+The encephalopathy dataset contains about 1.3 million reads, while the T2D dataset contains about 1.3 billion reads.
 
-\begin{table*}[t]
-\begin{center}
-\caption{Comparative Performance of UCLUST and Kraken Representations. 
-\label{tab:comp}}
-\begin{tabular}{|c|cccc|cccc|}\hline
-Representation & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
-UCLUST & 84.53 & 78.38 & 59.18 & 67.44 & 88.26 & 83.22 & 67.39 & 74.47\\\hline
-Kraken & 79.56 & 87.5 & 28.57 & 43.08 & 78.18 & 70.31 & 24.46 & 36.29\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+As previously mentioned, we use the publicly-available UCLUST \cite{Edgar10} software for clustering.
+For classification, we use Support Vector Machines (SVMs) \cite{vap95}, one of the most powerful and versatile binary classifiers used
+in myriad applications. 
+%
+For a  binary classification problem, the SVM framework 
+maximizes the separation between the two classes.
+In addition to the ubiquity and effectiveness of SVMs, they are well established in both supervised and multiple instance learning, making it easy to compare the two approaches.
+For our standard supervised SVM, we use svm-light \cite{joachims08}, a popular and publicly available implementation.
+For multiple instance learning, we use the python implementation by G.
+Doran and S.
+Ray ("MISVM") \cite{doran122013} for its ease of use and public availability.
 
-\begin{table*}[t]
-\begin{center}
-\caption{Three Pairwise Comparisons Using UCLUST. \label{tab:pairuclust}}
-\begin{tabular}{|c|cccc|cccc|}\hline
-Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
-Encephalopathy vs. No Encephalopathy & 88.68 & 90 & 86.54 & 88.23 & 84.47 & 81.77 & 81.77 & 81.77\\\hline
-Encephalopathy vs. Control & 96.72 & 95.35 & 100 & 97.62 & 94.26 & 93.2 & 100 & 96.48\\\hline
-No Encephalopathy vs. Control & 83.78 & 82.61 & 100 & 90.48 & 87.5 & 86.96 & 99.59 & 92.85\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+\subsection{Efficient Unsupervised Clustering using UCLUST}
+\label{uclustmethods}
 
-\begin{table*}[t]
-\begin{center}
-\caption{Three Pairwise Comparisons Using Kraken. \label{tab:pairkraken}}
-\begin{tabular}{|c|cccc|cccc|}\hline
-Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
-Encephalopathy vs. No Encephalopathy & 67.92 & 82.14 & 44.23 & 57.50 & 68.94 & 72.07 & 44.20 & 54.79\\\hline
-Encephalopathy vs. Control & 78.69 & 76.92 & 97.56 & 86.02 & 79.51 & 81.14 & 96.35 & 88.09\\\hline
-No Encephalopathy vs. Control & 77.03 & 77.03 & 100 & 87.02 & 81.42 & 81.42 & 100 & 89.76\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+From early experimental results, UCLUST can still be computationally infeasible for very large datasets such as the one we will be using, despite its generally excellent speed.
+For our T2D dataset, UCLUST was only able to cluster 2.8\% of the reads within 12 hours, at which point ARGO killed the batch job.
+UCLUST slows down as more clusters are formed, as each new read needs to be compared to more existing clusters.
+Thus, as the number of clusters grows, the execution time of UCLUST increases rapidly.
+The only way to feasibly complete the clustering phase in a reasonable amount of time was to cluster the patients individually in parallel, instead of clustering the reads of all patients together.
+However, this leads to the problem of having mismatched clusterings between patients, that is, cluster X from patient A may represent the same OTU as cluster Y from patient B.
+The solution is to compare the clusters from different patients and map them to each other, and then to combine clusters from different patients that represent the same OTU and output a single clustering output file for all patients.
+However, comparing every read from each cluster to every read from all other clusters is also computationally infeasible.
+Thus, we developed methods to sample representative reads from each patient and use them to create the consensus clustering.
+We explored a number of different methods, but two of them in particular were able to achieve significant speed improvements without significantly harming classification results.
+For each of them, we used a parameter "alpha", which was used to select 1 out of every alpha reads.
+For instance, if alpha=10, 1/10 of reads would be used.
 
-\begin{table*}[t]
-\begin{center}
-\caption{Comparative Performance After Feature Engineering. \label{tab:pfe}}
-\begin{tabular}{|c|cccc|cccc|}\hline
-Representation & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
-UCLUST & 85.64 & 79.49 & 63.27 & 70.46 & 88.54 & 82.58 & 69.57 & 75.52\\\hline
-Kraken & 80.66 & 76.92 & 40.82 & 53.34 & 80.8 & 75.86 & 35.87 & 48.93\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+We refer to one of our methods as "seed re-clustering" (UCLUST refers to cluster centroids as "seeds").
+In this method, the reads corresponding to the cluster seeds from each patient are copied into a temporary file, and then those reads are clustered.
+This runs quickly, since clusters in this setting tend to have hundreds to thousands of reads, so the number of cluster seeds is relatively small.
+We then construct a map, which maps each seed from its original cluster number to its new cluster number from the seed clustering.
+We go back to the individual patient clustering output, and select 1 out of every alpha lines to be placed in the consensus clustering file.
+As we write the lines to the consensus file, we use the map we developed to change the cluster number for the line from the original clustering number to the new consensus cluster number.
+We refer to the other method as "two round clustering".
+We select one out of every alpha reads from the clustering output of each patient, place them in a single file, and run UCLUST on this smaller subset to arrive at a consensus clustering.
+The main difference between two round clustering and seed re-clustering is that here we select a distribution of reads from the original clustering to determine the consensus clustering, instead of cluster seeds/centroids.
 
-\begin{table*}[t]
-\begin{center}
-\caption{Three Pairwise Comparisons Using UCLUST After Feature Engineering. \label{tab:pairuclustfe}}
-\begin{tabular}{|c|cccc|cccc|}\hline
-Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
-Encephalopathy vs. No Encephalopathy & 88.68 & 90 & 86.54 & 88.23 & 86.35 & 83.24 & 85.08 & 84.15\\\hline
-Encephalopathy vs. Control & 96.72 & 95.35 & 100 & 97.62 & 95.08 & 94.12 & 100 & 96.97\\\hline
-No Encephalopathy vs. Control & 87.84 & 86.36 & 100 & 92.68 & 91.22 & 90.57 & 99.59 & 95.05\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+We sought to empirically evaluate how these methods affected clustering runtime and classification accuracy, since feature vectors are generated from the clustering output and thus classification results are affected by the clustering methods.
+Without use of these sampling methods, UCLUST didn't finish within ARGO's 12 hour time limit on the T2D dataset, so we couldn't evaluate the classification results on that dataset and compare them to the classification results with the clustering sample methods.
+In order to evaluate these methods, we used the aforementioned Encephalopathy dataset, which is much smaller.
+We measured the time it took to perform UCLUST on all patients at once, as well as the time taken by each of our two sampling methods.
+We then used the respective clustering outputs to generate feature vectors and run svm-light classification (as described in the next section) and compared the results, which are discussed in the "Experimental Results" section.
+For UCLUST, default settings were used, except the threshold for placing two sequences in the same cluster was set to 40\% match (--id 0.40), and the --usersort option was used.
 
-\begin{table*}[t]
-\begin{center}
-\caption{Three Pairwise Comparisons Using Kraken After Feature Engineering. \label{tab:pairkrakenfe}}
-\begin{tabular}{|c|cccc|cccc|}\hline
-Comparison & Accuracy & Precision & Recall & F1-Score & LOOCV Accuracy & LOOCV Precision & LOOCV Recall & LOOCV F1-Score\\\hline
-Encephalopathy vs. No Encephalopathy & 69.81 & 83.33 & 48.08 & 60.98 & 77.18 & 80.88 & 60.77 & 69.40\\\hline
-Encephalopathy vs. Control & 86.89 & 86.67 & 95.12 & 90.70 & 87.30 & 88.15 & 96.88 & 92.31\\\hline
-No Encephalopathy vs. Control & 77.03 & 77.78 & 98.25 & 86.82 & 82.77 & 82.53 & 100 & 90.43\\\hline
-\end{tabular}
-\end{center}
-\end{table*}
+\subsection{Supervised and Multiple Instance Classification using Support Vector Machines (SVMs)}
+\label{svm}
 
-Table \ref{tab:time} shows 
-the run time of each representation in seconds. We had total of 1.3 million sequence reads. UCLUST unsupervised clustering 
-took 180 seconds, while Kraken's OTU database lookups took 120 seconds.
+For our formulation we use SVMs to make clinical phenotype prediction for each patient sample based on their  input metagenome sequence samples represented with their clustering or taxa membership features.
+With our clustering approach,  we have a set of reads that have been assigned to clusters based on UCLUST and our techniques.
+We now generate feature vectors from each patient from this clustering output.
+We follow a different process of feature vector generation for svm-light and MISVM.
+For each method, we read through the consensus clustering output from UCLUST line-by-line, forming feature vectors as we go.
+UCLUST output lines provide the cluster number, the percentage match to the cluster seed, and the patient that read comes from.
 
-Figure \ref{prefeature} and Table \ref{tab:comp} show 
-the classification performance of UCLUST and 
-Kraken in distinguishing patients in the "Encephalopathy" class versus 
-the other classes. UCLUST correctly predicted the clinical 
-phenotype 84.53\% of the time, and had a precision of 78.38\% and recall of 59.18\% on the held-out set. LOOCV results were 
-slightly better for each metric than results on the test 
-set by about 4-8\%. Kraken had an accuracy of 79.56\% with an even 
-larger disparity between precision and recall (87.5\% for the former and 
-28.57\% for the latter), which led to a relatively poor F1-score of 43.08\%. LOOCV results for Kraken 
-were worse by 1.38\% for accuracy, 17.19\% for precision, 4.11\% for recall, and 6.79\% for F1-Score. As 
-the results show, UCLUST generally performed better as a feature 
-representation, with 4.97\% higher accuracy and 24.36\% higher F1-score.
-
-Table \ref{tab:pairuclust} displays the performance of UCLUST-based classification 
-with regards to training three one-versus-one clinical phenotypic classifiers.  The first comparison 
-distinguished between the "Encephalopathy" class and the "No Encephalopathy" class, effectively determining whether or not a patient who had liver cirrhosis also had encephalopathy.
-Results were much improved when compared with the findings in table \ref{tab:comp}.
-The accuracy, precision, recall, and F1-score were all  86-90\%.
-LOOCV
-results were worse by about 4-8\% per metric. The 
-second comparison distinguished between the "Encephalopathy" class and 
-the "Control" class, effectively determining whether a 
-patient had encephalopathy caused by liver cirrhosis or had neither disease.
-This comparison had the best results of the ones we examined, with greater than 95\% performance in all metrics and 100\% recall.
-The cross validation
-results were about 1-2\% worse for each metric except recall, which was still 100\%. The 
-third comparison distinguished between the "No Encephalopathy and liver cirrhosis" class and the "Control" class, effectively determining which among the patients lacking encephalopathy had liver cirrhosis.
-This comparison had an accuracy of 83.78\% and precision of 82.61\%, but had a 100\% recall which led to a high F1-score of 90.48\%.
-Cross validation results were better by 3.72\% for accuracy, 4.35\% for precision, and 2.37\% better for F1-Score, with essentially the same recall.
-Apart from the accuracy and precision of the third comparison, these three pairwise comparisons generally performed better across the board than the results in Table \ref{tab:comp}, with the second pairwise comparison having the best performance.
-
-Table \ref{tab:pairkraken} displays the performance of Kraken with regards to the same three pairwise comparisons described above.
-For the Encephalopathy vs.
-No Encephalopathy run, the accuracy was 67.92\%, with 82.14\% precision, 44.23\% recall, and 57.50\% F1-Score.
-Cross validation produced a precision that was 10.07\% lower but largely the same accuracy and recall.
-The Encephalopathy vs.
-Control run produced good results, with 78.69\% accuracy, 76.92\% precision, 97.56\% recall, and 86.02\% F1-Score.
-LOOCV results mostly confirmed those results, except the precision, which was 4.22\% higher.
-The No Encephalopathy vs.
-Control run had quite similar results to the Encephalopathy vs.
-Control run.
-The main pattern that appears to emerge from these results is that the recall for Kraken was generally far superior in the pairwise runs than in the original run, which increased the F1-Scores but resulted in similar accuracy and precision.
-The Encephalopathy vs.
-No Encephalopathy run was significantly worse than the other two.
-
-Table \ref{tab:pfe} shows the results for the same setup as 
-in Table \ref{tab:comp} after we performed feature engineering.
-For both representations in Table \ref{tab:comp}, the recall was significantly lower than the precision, indicating that there were many more false negatives than false positives.
-We thus engineered the features in the training set.
-Certain features in the training set were most strongly present in positive patients, and as such were correlated with positive examples.
-Some examples in the test set had slightly lower values for those features, and were thus labelled by the classifier as negative examples, when in reality they were positive.
-We aimed to influence some of these borderline false negative predictions by lowering the threshold for those positively correlated features that would lead to an example being predicted as positive.
-In order to reduce the threshold for those positively correlated features, we applied a multiplier of 0.9 to the values of each feature of every positive patient in the training set.
-This had the effect of causing some previously borderline negative predictions to become positive predictions, primarily reducing the number of false negatives and improving the recall scores, but also positively impacting most other metrics as well.
-Attempts to use a multiplier of
-less than 0.9 did not appear to further 
-improve the results.
-
-With the multiplier of 0.9 applied to the 
-feature values for positive training examples, the recall 
-increased by 4.09\% for UCLUST and by 12.25\% for Kraken. This was confirmed in the 
-cross validation results as well, 
-in which UCLUST's recall increased by 2.18\% and 
-Kraken's recall increased by 11.41\%. Because of the improvement 
-in recall, both representations also had similarly improved F1-Scores after feature engineering. Accuracy and precision remained generally consistent. 
-%
-Kraken's LOOCV precision 
-increased by 5.55\%.
-This suggests that Kraken's very high precision was tied to its lack of many positive phenotype predictions in general, resulting in a very low number of false positives.
-While Kraken improved more than UCLUST did due to feature engineering, UCLUST still performed better in all metrics.
-
-Tables \ref{tab:pairuclustfe} and \ref{tab:pairkrakenfe} display the results after the same feature engineering techniques were applied to the pairwise results in \ref{tab:pairuclust} and \ref{tab:pairkraken}, respectively.
-UCLUST's No Encephalopathy vs.
-Control run saw an increase of 4.06\% in accuracy, 3.75\% in precision, and similar gains in its cross validation results for accuracy and precision.
-UCLUST's other two runs were largely unaffected, albeit with slightly better cross validation results.
-Kraken's Encephalopathy vs.
-Control run was greatly improved after feature engineering, with an 8.20\% increase in accuracy and a 9.75\% increase in precision, with similar increases in cross validation results.
-This more than offset a slight decrease in recall.
-The Encephalopathy vs.
-No Encephalopathy run had mostly unchanged results, albeit with much improved cross validation scores, and the No Encephalopathy vs.
-Control run was not significantly affected.
+For svm-light, we have one feature vector per patient.
+We use each cluster/OTU as a feature, meaning that the presence of various microbes are the features for the patient.
+Each number (feature) in the vector corresponds to a cluster/OTU, and the value for that number corresponds to how closely the read from the patient matched the cluster seed.
+For each input
+sequence, we use the percentage match to its cluster seed as 
+the value for that feature and average the 
+sequence match if multiple reads from the metagenome samples 
+are assigned to the same cluster by UCLUST.
+Each sequence read includes a label for the patient that it came from, so each read only affects the feature vector corresponding to the patient that it came from.
+For example, say patient A has a read with 95\% match to OTU 2, a read that is the cluster seed of OTU 3, a read that is a 93\% match to OTU 2, a read that is a 97\% match to OTU 5, and no reads for OTU 1 or 4.
+Patient A's feature vector would then look like: [0 0.94 1 0 0.97].
+We first initialize the patient feature vectors to all 0s and then iterate through the UCLUST output lines and update the feature vectors as we go.
+For MISVM, patients are bags, reads for patients are instances, and each read/instance has a feature vector.
+We iterate through the UCLUST output.
+For each read, we add that read as an instance to the relevant patient bag.
+The feature vector for that read contains one number: the cluster number that the read belongs to.
+We compared the runtime and accuracy of the two implementations, which both used a linear kernel.
+In both cases, the training and test set each contained 30 patients, and each contained 15 patients with type 2 diabetes and 15 control (healthy) patients.
+The results are discussed in the "Experimental Results" section.

### 2016-08-03 latex/input/intro.tex
added script for generating plots of GICF runs; updated paper

@@
 
-The human body is one of the 
-most densely populated microbial 
-environments in the world, with human host cells interacting 
-with  more than $10^{14}$ microbial cells. Collectively, the human 
-and microbial cells are referred to as the 
-\emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}. 
-%
-%
-%
-Biotechnological  advances 
-can 
-interrogate the microbial 
-communities present across three primary dimensions: (i) clinically different
-patients like those suffering from liver disease versus healthy controls (ii) different
-patient sites and tissues like skin and the gastrointestinal tract 
-and (iii) across different patient visits~\cite{Costello12182009,Qin:2010fk} and before/after treatment. 
-%
-Methods to query this  microbial
-data include DNA sequencing of 
-specific marker genes such as the 16S rRNA marker genes~\cite{Woese97}, 
-sequencing
-the entire pool of microbial genomes at once (metagenomics), or 
-sequencing the mixture of 
-expressed gene transcripts (metatranscriptomes).
-%
-These sequencing methods along 
-with proteomics and mass spectrometry
-allow us to study the
-biotransformations caused by the microbial communities. 
-%
-Several researchers and clinicians have embarked on
-studies of the pathogenic and clinical 
-role played by the microbiome with respect to
-human health and disease conditions.
+The human body contains one of the most dense and diverse microbial environments in the world.
+The human and microbial cells are collectively referred to as the \emph{human microbiome} \cite{turnbaugh_human_2007,backhed_host-bacterial_2005}.
+Advances in biotechnology have allowed scientists to directly interrogate the human microbiome, particularly the development of high throughput sequencing technologies, which generate massive amounts of biological data.
+In recent years, improving capabilities in data science have allowed for the study of \emph{metagenomics}, which involves sequencing the entire pool of microbial genomes at once.
+This data is usually gathered via \emph{shotgun sequencing}, which generates a number of short \emph{reads} containing bits of genomic data from the microbes of the host environment \cite{messing81}.
+These reads, represented as strings of nucleotides, represent only small parts of the microbe's full genome, and are not ordered in any way, which presents several challenges that will be discussed further in the paper.
+Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (phylogeny), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes \cite{handelsman04}.
+Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.
 
-In this study, we specifically developed
-a computational method to predict the clinical phenotype
-of a patient i.e., a pipeline to 
-predict whether or not the patient has a specific type and level of  a disease.  This 
-pipeline
-is able to make the phenotypic
-predictions from input metagenomic sequences using post-processing 
-with unsupervised binning approaches (clustering) or  supervised search based 
-taxa profiling approaches.  This method is designed to 
-run quickly and produce accurate and sensitive 
-results. We  compare and evaluate two state-of-the-art 
-approaches, one for binning (UCLUST \cite{Edgar10}) and one for taxonomic profiling (KRAKEN \cite{Wood14}) 
-to represent the input metagenome sequences as features for classification. The classification
-formulation uses a binary support vector machine based classifier \cite{vap95}. We also 
-evaluate the performance of preliminary 
-feature scaling and engineering  for classification purposes.
+In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome.
+This can be viewed as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them.
+In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances).
+Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}.
+MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome.
+However, since datasets in this domains frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.
 
-We specifically evaluated our approach on 
-microbiome samples of patients suffering from hepatic encephalopathy due 
-to liver cirrhosis. 
-%
-We found that 
-UCLUST performed better than Kraken on our data 
-set, although Kraken was faster. We were also able to 
-engineer the data to improve the accuracy of the classifier. Our 
-pipeline had 85.64\% accuracy using UCLUST and 80.66\% accuracy 
-using Kraken.
+We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype.
+This data is very large (hundreds of gigabytes) and high-dimensional (thousands of dimensions).
+Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled.
+\emph{Assembly} is the process of finding pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome, and combining them.
+This process is repeated as much as possible to form long strings called \emph{contigs}.
+Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully.
+Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads.
+The reduction in data size also allows for clustering, which is not feasible for massive datasets.
+Clustering uses string similarity measures to group similar reads into "clusters", which is a way of identifying which species of microbe each read corresponds to.
+The alternative, "aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced.
+From the clustering output, we extract feature vectors, which are then fed into a standard Support Vector Machine (SVM) classifier.
 
-%For our data set with 1.3 million reads, UCLUST took 
-%180 seconds to run and Kraken took 120 seconds. 
-%
+Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM.
+This process is explained in further detail in the methods section.
+We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data.
+We show that our classifier shows significantly improved performance.
+This is discusses further in the Experimental Results section.
+
+The rest of the paper is organized as follows.
+Section 2 presents relevant Background information on  Multiple Instance Learning.
+Section 3 presents the methods we used in the creation of our pipeline.
+Section 4 presents Materials, such as dataset, hardware, and software descriptions.
+Section 5 presents our results, and Section 6 presents our conclusions.

### 2016-08-03 latex/input/materials.tex
added script for generating plots of GICF runs; updated paper

@@
 
-We obtained 904
-clinical microbiome samples from a study by Dr. Patrick Gillevet 
-that relates to patients who suffered from hepatic encephalopathy due 
-to liver cirrhosis. Specifically, the classification formulation 
-was setup to distinguish between patients suffering with a specific 
-clinical phenotype or not. There were a total of 239 
-patients with Encephalopathy   due to liver cirrhosis (denoted 
-by the ``Encephalopathy'' class), 590 patients with liver cirrhosis 
-but no Encephalopathy (denoted by ``No Encephalopathy'') 
-and 75 patients who were considered as 
-control and did not have either of the clinical conditions (denoted by ``Control'').
-%
-16S rRNA metagenomic sequence read data was 
-obtained from patient stool samples. On average 
-there were a total of 1,464  sequence reads of length 200-400 obtained per 
-patient, with 1,323,016 total reads across all patients in our dataset.
-
-\subsection{Evaluation Metrics}
-We assess the performance of our classification pipeline 
-in terms of correctness and execution time.
-%
-Given the imbalanced nature of class distributions,  the 
-performance of  binary classifiers was measured by 
-F1 score, precision and recall, along with
-the standard accuracy metric. We discuss 
-these standard metrics in brief.
-
-Accuracy measures the percentage of instances 
-that are classified correctly and can 
-be represented by 
-\begin{equation}
-Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
-\end{equation}
-where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.
-
-Accuracy as an evaluation metric can be biased if one of the classes 
-(positive or negative)  has a larger number of examples than the other.   
-Precision measures the percentage of positive predictions that 
-were correct, 
-whereas recall measures the percentage of positive 
-examples that were correctly predicted (or retrieved). 
-We can represent Precision and Recall by \cite{Goutte}:
-
-\begin{equation}
-Precision = TP / (TP + FP). \label{eqn:prec}
-\end{equation}
-\begin{equation}
-Recall = TP / (TP +FN). \label{eqn:roc}
-\end{equation}
-
-The F1 score captures the trade offs between precision and recall in a
-single metric and is the harmonic mean of precision and recall \cite{Goutte}, given 
-by:
-\begin{equation}
-F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
-\end{equation}
+We used data from a well-known Metagenome-Wide Association Study by Qin et al.
+of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}.
+The total dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online.
+Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
+and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
+The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}.
+Each patient file was several gigabytes in size, with the total dataset size being over 600 gigabytes.
+Qin et al.
+develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "e10171", respectively \cite{qin041012}.
+They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}.
+While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available.
+The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.
 
 \subsection{Software and Hardware Details}
-We used the Argo computing cluster available at George Mason University. 
-%
-The representation generation phase using UCLUST and Kraken and and the classification phase
-were 
-run on one of the  compute nodes available on the cluster. The cluster is configured with 
-35 Dell C8220 Compute Nodes, each with dual 
-Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores – 
-528 and 1056 total threads, RAM \textgreater 2TB)\cite{ORC}.
-
-Source codes for UCLUST \cite{Edgar10} and KRAKEN \cite{Wood14} were downloaded from 
-their respective websites\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html     
-Kraken: https://ccb.jhu.edu/software/kraken/} and compiled on the Argo platform. 
-%
-Kraken aligns reads to an OTU database. The standard Kraken database is 160GB in size. Due to 
-computational limitations we used the custom, small sized 4GB
-database  in this feasibility study \cite{Kraken}.
-
-For the  SVM-based classification\cite{Joachims08}  we used the popular 
-SVM-Light \cite{Joachims08}  source code, which is publicly 
-available \footnote{http://svmlight.joachims.org/}.  The linear kernel was used
-and the regularization parameters were set to their default values.
-
-\subsection{Experimental Protocol}
-
-For evaluating the performance of our binary phenotypic classifiers, we 
-split the patient samples into a training set containing 80\% of the 
-patients and the test set containing 20\% of the patient samples. Every fifth sample was placed in the test set, so the relative class sizes were roughly consistent between the training and test sets.
-We also performed leave-one-out cross validation (LOOCV).
-
-In the next section, we discuss the accuracy of our 
-classifier with regards to predicting the clinical 
-phenotype of a given patient using 
-either the unsupervised 
-clustering representation with UCLUST or 
-supervised OTU representation 
-with Kraken. We present results for classifiers 
-distinguishing patients  in the ``Encephalopathy'' class versus
-the other classes, effectively 
-determining whether or not a patient had Encephalopathy 
-regardless of whether they had liver cirrhosis. We also 
-present pairwise one-versus-one classification results that 
-compared two phenotypes within the dataset, i.e.  Encephalopathy versus
-No Encephalopathy, Encephalopathy versus
-Control and No Encephalopathy versus Control.
+We used the ARGO computing cluster available at George Mason University \cite{argo}.
+The clustering and classification phases were run on one of the compute nodes available on the cluster.
+The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM.
+(Total Cores 528 and 1056 total threads, RAM$>$2TB) \cite{argo}.
+Source codes for
+SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
+their respective websites
+ and compiled on the Argo platform.

### 2016-08-03 latex/input/methods.tex
added script for generating plots of GICF runs; updated paper

@@
 %methods
+
+\begin{figure*}[t]
+\centering
+\includegraphics[scale=0.5]{./mil-metagenomics-pipeline.png}
+\caption{This diagram illustrates the entire pipeline.
+Patient files with FASTQ metagenomic reads are individually assembled using SOAPdenovo2, then combined into one file and clustered with UCLUST.
+We extract features according to the D-BoW method, and classify the patients using the feature vectors with svm-light.} \label{pipeline}
+\end{figure*}
 
@@
 
-Figure \ref{fig1} provides an overview of the 
-developed computational pipeline. 
-%
-We obtain 16S metagenomic sequence reads from patients who 
-had encephalopathy and liver cirrhosis, no encephalopathy but liver cirrhosis, and controls (with neither phenotype). The input sequence 
-reads are annotated using a taxonomic profiling approach (Kraken) or 
-an unsupervised binning approach (UCLUST). The output of this annotation phase 
-features are generated, which are passed on the binary support vector machine 
-classifier (SVMs).  We used the  svm-light implementation \cite{Joachims08}  
-to perform the classification of whether or not a patient 
-had a clinical phenotype. We compared the classification performance of the 
-SVM classifier with the two intermediate sequence representations obtained from 
-UCLUST and Kraken. We also applied feature scaling and 
-selection techniques to improve the classification results.
+As mentioned in the Introduction, our pipeline involves a number of steps, which serve a variety of purposes.
+For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled.
+The reads that could not be assembled were unlikely to be useful anyway, since they could not be assembled with other reads into useful contigs.
+The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier.
+We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center.
+Finally, we used a standard SVM classifier to predict patient phenotype, and used several metrics to assess its accuracy.
+We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the results section.
+Aside from the patient labels, this process is entirely de novo, and does not consult any external databases.
+An illustration of this pipeline can be seen in Figure \ref{pipeline} on page \pageref{pipeline}.
 
-\begin{figure*}[t]
-\begin{center}
-\caption{Pipeline Overview. \label{fig1}}
-\includegraphics[scale=0.7]{./diagrams/concept}
-\end{center}
-\end{figure*}
+\subsection{Assembly with SOAPdenovo2}
 
-\subsection{Unsupervised Clustering using UCLUST}
-\label{uclustmethods}
+For our assembly step, we used SOAPdenovo2 \cite{luo12}, which is a de novo metagenome assembly tool.
+This tool was also used for assembly in the MGWAS study that our data came from.
+We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers\cite{peng12}.
+We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51.
+The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together.
+Conversely, all contigs need to be in one file for clustering, to avoid inconsistent cluster assignments between different patients.
+Thus, we combined the contigs from each assembled patient file into a single for clustering.
 
-In this work we evaluate the use of an unsupervised approach 
-for representing our input metagenome sequence reads. Specifically, we 
-  use UCLUST \cite{edgar2010search} due to its superior performance in terms 
-  of run time and accuracy.
-%
-%
-%
-UCLUST \cite{edgar2010search} follows a greedy, iterative clustering approach. As a first 
-step, this approach identifies exact matches of fixed length between sequence pairs known 
-as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
-the aligned pairs. Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and 
-used for further processing. As such, the step eliminates a lot of pairwise comparisons.
+\subsection{Clustering with UCLUST}
 
-Then UCLUST follows an 
-incremental approach for assigning the input sequences to the different bins. 
-%
-%
-The clustering solution is initialized as an empty list. Sequences are then 
-incrementally added to the clusters existing within the list. 
-%
-Each unassigned input sequence is compared to the cluster representatives within the list using 
-the fast indexing search technique that uses the HSPs. If a match is found with one of the 
-cluster representatives, then that sequence is assigned to that particular cluster or the 
-input sequence forms a new cluster. 
-%
-UCLUST ensures
-that the cluster representatives are sequences with the largest length.
+We used UCLUST \cite{Edgar10}, which has been shown to be among the simplest, fastest, and most effective clustering methods \cite{bonder090112, sun042711}.
+Since our contigs were not ordered, we used the usersort option, and we set the sequence match id to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster.
+For instance, between two strings of length 100, at least 40 places in each of those strings would have to contain the same nucleotide (represented as A, T, G, or C).
+New contigs that did not match at least 40\% to any of the existing contigs would form the seed of a new cluster.
+Again, this value of 40\% was found to be the best value based on our experiments with this dataset.
 
-\subsection{Supervised Taxonomic Identification using Kraken}
+\subsection{Feature Extraction}
 
-We also evaluate the use of a search based 
-approach to taxonomically label our sequence reads. We use 
-Kraken for this, due to its strong performance in terms of run time 
-and precision. Kraken finds $k$-mers in an input 
-sequence read and queries a database of OTUs. It then associates the 
-read with the lowest common ancestor (LCA) of the genomes that contains $k$-mers from the read \cite{Kraken}. In order to reduce 
-the number of false positives, Kraken does not classify reads for 
-which it does not find sufficient evidence of any match.
+Our feature extraction operation followed the D-BoW method.
+For each patient, we initialized a vector with length equal to the number of clusters.
+The value for each place in that patient's feature vector would then be set to the closest percentage match (0.4 to 1) of the reads from that patient to the relevant cluster seed.
+For instance, if there are three clusters of reads, and patient A has the cluster seed for cluster 1, no reads from cluster 2, and a read with 70\% match to cluster 3, their feature vector would be~~A = [1.0~~~0.0~~~0.7].
+This is illustrated in Figure \ref{pipeline}.
 
-This approach is designed to perform quickly and with 
-high precision. However, it also involves downloading and building 
-a large database to query. It is often infeasible to download and 
-build the database in its entirety, so a smaller custom  
-subset of the database is available for testing 
-purposes. We used this 
-subset of the entire database in this study (referred to as kraken-light).
+\subsection{Classification and Evaluation Metrics}
 
-\subsection{Classification using Support Vector Machines (SVMs)}
+We performed classification with a standard SVM classifier using the generated feature vectors; in this case, we used svm-light \cite{joachims08}.
+The choice of classifier is not very important for D-BoW methods \cite{amores13}.
 
-Support Vector Machines (SVMs) \cite{vap95} are 
-one of the most powerful and versatile binary classifiers used 
-in myriad applications. 
-%
-For a  binary classification problem, the support 
-vector machine (SVM) \cite{vap95} framework 
-maximizes the separation between the two classes.
-%
+We can assess the success of our classifier in several ways. The simplest measure, accuracy, measures the percentage of instances 
+that are classified correctly, represented by 
+\begin{equation}
+Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
+\end{equation}
+where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.
 
-Given an input set of positive training instances and 
-negative training instances, the SVM classifier learns a 
-mathematical function $f$  that maximizes the separation between 
-the given positive and negative training instances. The function $f$ will then be able to predict whether a new instance (considered a test instance) should have a positive or negative class label.
-%
-For our formulation we use SVMs to make clinical phenotype prediction for each 
-patient sample 
-based on their  input metagenome sequence samples represented with their 
-clustering or taxa membership features.
+Accuracy as an evaluation metric can be biased if one of the classes 
+(positive or negative)  has a larger number of examples than the other.   
+Precision measures the percentage of positive predictions that 
+were correct, 
+whereas recall measures the percentage of positive 
+examples that were correctly predicted (or retrieved). 
+We can represent Precision and Recall as:
 
-For all the available microbiome samples,  we have a set of reads processed via UCLUST or Kraken. 
-%
-Each patient is associated with a vector that is used by the SVM to help build $f$.
-Each row in the vector, called a feature, corresponds to a certain species, and the value for that row corresponds to how closely the read from the patient matched what the species was expected to look like.
-The exact process of assembling a feature vector for a
-patient depends on which data representation is being used. For 
-UCLUST, we use each cluster seed 
-as a feature. For each input 
-sequence, we use the percentage match to its cluster seed as 
-the value for that feature and average the 
-sequence match if multiple reads from the metagenome samples 
-are assigned to the same cluster by UCLUST. 
-%
-For Kraken, we use the taxonomic label associated 
-with the read as the feature for the SVM. The percentage match of the 
-read to that label is used as the feature value. Similar 
-to UCLUST, we average the similarity score if multiple reads align to the 
-same taxonomic class label. 
-%
-Our feature selection process involves discarding Kraken's unlabeled 
-reads and UCLUST's clusters with only one read (singletons).
+\begin{equation}
+Precision = TP / (TP + FP). \label{eqn:prec}
+\end{equation}
+\begin{equation}
+Recall = TP / (TP +FN). \label{eqn:roc}
+\end{equation}
 
-We used cross-validation and a held out set procedure (discussed in the next 
-Section) to assess the performance of SVM classification using UCLUST and Kraken
-as intermediate representations.
+The F1 score captures the trade-offs between precision and recall in a
+single metric and is the harmonic mean of precision and recall, given by:
+\begin{equation}
+F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
+\end{equation}
+
+Finally, we also use the Area Under Curve of the Receiver Operating Characteristic (AUC-ROC), which measures the performance of the classifier as the decision boundary threshold is moved.
+The SVM classifier generally predicts a group label to be negative if the predicted label for that group was less than 0 and predicts a group label to be positive otherwise.
+The AUC-ROC measures the performance of the classifier as the threshold is varied to more or less than 0.
+In effect, it measures how far off incorrect predictions were from being correct.
+AUC-ROC plots True Positive Rate versus False Positive Rate, given by:
+\begin{equation}
+True Positive Rate = Recall = TP / (TP + FN). \label{eqn:roc}
+\end{equation}
+\begin{equation}
+False Positive Rate = FP / (FP + TN). \label{eqn:prec}
+\end{equation}

### 2016-08-03 latex/input/related.tex
added script for generating plots of GICF runs; updated paper

@@
 
-\subsection{Human Microbiome and Metagenomics}
+\subsection{Multiple Instance Learning}
 
-The combination of human host cells and microbial 
-cells that govern several facets of human health and pathology are referred 
-to as collectively by the term \emph{human microbiome} \cite{hugenholtz2008microbiology}. 
-%
-Using today's sequencing technologies we have the capacity to determine 
-the DNA sequences of these co-existing microbial communities. 
-However, 
-current  genomic  technologies 
-do not provide the complete genome for each individual microbe, but 
-  short, contiguous subsequences from random positions of the genome. These 
-  short subsequences are referred by ``reads''. 
-%
-Metagenome sequence assembly is defined as the process of taking the different 
-sequence reads from the various microbial organisms to produce long contiguous 
-sequence of DNA for each individual organism within the community mixture. 
-%
-The assembly process involves identifying overlapping parts of different reads 
-to order them and separate them in organism-specific larger sequences.
+The Multiple Instance Learning (MIL) problem was first described by Dietterich in the context of drug activity prediction \cite{dietterich97}.
+In this problem, we want to figure out which of a number of different molecules will bind to a target "binding site".
+Each molecule can assume a number of different 3-dimensional shapes, so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding.
+If even one shape of a given molecule binds to the binding site, it is considered a "good" molecule \cite{dietterich97}.
+Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains all negative instances.
+In the original paper, Dietterich develops a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}.
+In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, and MILES \cite{wang06}.
+A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.
 %
 
-The metagenome assembly problem is known to be challenging due 
-  to the similarity of genomes from the different microbes, 
-differing 
-  abundance, diversity, complexity and varying
-  genome lengths of never-sequenced before microbes within 
-  the different microbiome samples. DNA sequencing machines  
-  are very high throughput and produce Terabytes of data per run and 
-  also produce reads with poor quality  \cite{hugenholtz2008microbiology}. Several studies 
-  have highlighted the challenges associated with the metagenome analysis and assembly problem by performing a simulation study to determine the feasibility of solving the metagenome assembly problem \cite{charuvaka2011evaluation}.
-
-  %
-  Targeted metagenomics or 16S rRNA gene sequencing provides
-  a
-  first step for the quick and accurate characterization of
-  microbial communities. 16S sequences are marker genes, which
-  exists in most microbial genomes and
-  have a conserved portion
-  for detection (primer development) and a variable portion that allows for
-    categorization within  different  taxonomic groups \cite{petrosino2009metagenomic}.
-        %
-        Targeted metagenomics are also effective in detecting species with low 
-        abundances. However, they may not be good at discovering 
-        unique species (orphans) that have never been sequenced before. % ones that are considered to be orphan (i.e., never sequenced before).
-
-\subsection{Microbiome Informatics}
-
-Since the human microbiome project \footnote{http://hmpdacc.org/} and 
-release of publicly available metagenomic datasets began, a host 
-of methods have been developed for the analysis of 
-assembled metagenomes and  input  sequence reads (before assembly)
-obtained from 16S rRNA genes and whole metagenomes.  
-%
-%
-The relationship between microbiome and human health can be characterized by first
-identifying the content, abundance, and functionality of the microbes within the samples. Several
-computational approaches (surveyed here \cite{:mo}) have been developed for handling the vast
-amount of metagenomic data to solve two related problems: (i) clustering or binning, and (ii) taxonomy profiling methods.
+However, since 2010, there has been increased interest in different formulations of the MIL problem.
+For instance, the problem of "key instance detection" \cite{zhou12} revolves around finding the instances that contribute the most to bag labels.
+A recent study focused on a formulation of the MIL problem in which bags with negative labels can contain some "positive" instances, and developed a general cost function for determining individual instance labels from group labels \cite{kotzias15}.
+This is significant in metagenomics because, while some diseases are caused by a single pathogen, many arise from a combination of many factors, and even patients that are "healthy" may contain small amounts of pathogens that are normally associated with disease.
 %
 
-\subsubsection{Binning Methods}
+However, this method treats a group label simply as an aggregation of instance labels.
+Amores, in his taxonomy, called this type of method "Instance-based", and found this paradigm to be ineffective \cite{amores13}.
+Amores identifies two other paradigms, Bag-Space and Embedded-Space.
+An example of the latter are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
+Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient \cite{amores13}.
+The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center.
+Despite these recent developments in MIL, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.
 
-The ``binning'' problem  involves grouping input short reads  such that reads within a group are 
+\subsection{Assembly}
+
+The \emph{assembly} problem involves combining overlapping short reads into longer sequences called \emph{contigs}.
+
+\subsection{Clustering}
+
+The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are 
 similar to each other. This process may lead to groups that are organism-specific and is 
-unsupervised in nature, being referred by clustering in the data mining community. 
+unsupervised in nature. 
 %
-The binning process does not attempt to provide an automated labeling of the input reads. 
+The clustering process does not attempt to provide an automated labelling of the input reads. 
 %
-The clusters/bins/groups obtained from an input metagenome sample 
-is referred to as  Operational Taxonomic Units (OTUs), and the number of 
+The clusters/groups obtained from an input metagenome sample 
+are referred to as  Operational Taxonomic Units (OTUs), and the number of 
 OTUs gives an approximation of species diversity in a sample \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
@@
 environmental samples contain microbial organisms that have never been cultured in a laboratory, and thus those organisms do not exist in genomic databases.
-As such,  binning of 
+As such,  clustering of 
 sequence reads has several advantages: (i)
@@
 are some of the popular 
-metagenome/sequence clustering approaches used for binning. 
+metagenome sequence clustering approaches used for binning. 
 %
@@
 of expensive sequence alignment) and followup with an incremental clustering approach 
-that does not involve comparing all pairs of input sequences. In Section \ref{uclustmethods} we discuss 
-in detail the UCLUST \cite{Edgar10}  sequence clustering approach. We use UCLUST within 
-our study, known to be a state-of-the-art metagenome clustering approach in terms of 
-computational performance and accuracy of results.
+that does not involve comparing all pairs of input sequences.
+We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}.
+%
 
-\subsubsection{Taxonomy Profiling Methods}
-\label{sec:tax}
-
-The taxonomy profiling problem involves 
-assigning a specific label (i.e., a phylogenetic group label) to sequence 
-reads or assembled metagenome contigs~\cite{phymm}. A traditional 
-approach for taxonomy profiling is to formulate  a 
-classification problem (supervised) that uses marker genes for
-  identification of source organism of a sequence read or fragment~\cite{Woese97}. Marker
-  genes are highly conserved and provide accurate identification
-  of the taxonomy class~\cite{ZongzhiLiu10012008}. These approaches 
-  rely on a previously annotated 
-  reference dataset like RDP~\cite{Cole2005,Cole2007,Cole2008} or GreenGenes~\cite{greengenes}. These methods
-  provide
-  valuable community estimates but are limited to specific 
-  reads or contigs (marker genes constitute a small fraction of a metagenomic
-  sequence set) and have low sensitivity due to reliance
-  on an incomplete and taxon-biased reference genome database.The RDP database
-  implements a naive Bayes classifier 
-  that uses DNA-composition features to classify 16S RNA sequence reads into taxonomic classes as defined
-  by Bergey's taxonomy~\cite{ZongzhiLiu10012008}.
-
-  Several comparative methods have been developed for the assignment of phylogenetic class based on principles of homology. Such methods  align
-  reads or contigs using BLAST~\cite{altschul90a}, and assign taxonomy based on the best match with a reference database~\cite{Tringe04222005}. MEGAN~\cite{megan} and MARTA~\cite{marta} are
-  metagenomic analysis and visualization programs that  make the assignment based on multiple BLAST hits and optimized parameters.  The MG-RAST~\cite{mgrast}  web server
-  provides taxonomic and functional annotation by comparative searches performed across multiple reference databases. GAAS~\cite{gaas} is a novel BLAST-based tool
-  that includes genome length normalization along with a similarity weighting for multiple BLAST hits to provide improved estimates.
-
-  Composition-based methods have also been developed. They extract key sequence features such as GC composition and $k$-mer frequencies and build supervised
-  classification models using those features. PhyloPythia~\cite{McHardy2007} uses a support vector machine framework~\cite{vap95} to classify long reads into taxonomic groups
-  using a $k$-mer based kernel function.
-TETRA~\cite{tetra} correlates the $k$-mer pattern feature to different taxonomic groups.
-Kraken, which is used in this paper, searches a taxonomic database for the lowest common ancestor (LCA) of the genomes that contain $k$-mers from a sequence read.
-Phymm~\cite{phymm} trains an interpolated Markov model
-  to characterize variable length subsequences  specific to different taxonomic subgroups. In combination with BLAST, Phymm shows improved classification
-  accuracy for short reads of 100 base pair (bp) length. Such Markovian models have been very successful in gene finding algorithms like Glimmer~\cite{delc98nucl}.
+UCLUST \cite{Edgar10} follows a greedy, iterative clustering approach. As a first 
+step, this approach identifies exact matches of fixed length between sequence pairs known 
+as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
+the aligned pairs.
+Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and used for further processing.
+As such, the step eliminates a lot of pairwise comparisons.
+Then UCLUST follows an incremental approach for assigning the input sequences to the different bins.
+The clustering solution is initialized as an empty list.
+Sequences are then incrementally added to the clusters existing within the list.
+Each unassigned input sequence is compared to the cluster representatives within the list using the fast indexing search technique that uses the HSPs.
+If a match is found with one of the cluster representatives, then that sequence is assigned to that particular cluster; otherwise, the input sequence forms a new cluster.
+Matches are computed by comparing two reads, character by character, and dividing the number of characters at the same position in both reads that match by the length of the longer of two reads.
+UCLUST ensures that the cluster representatives are sequences with the largest length.

### 2016-08-03 cirrhosisTests/gicf.py

@@ -231,7 +231,7 @@ def validate(verbose, w, data, k):     # apply the weights determined by gradien
     return accuracy, precision, tpr_recall, fpr, f1_score, predicted_actual
 
 
-def grid_search(verbose, w, data, train, test, res, limit):
+def grid_search(verbose, w, data, train, test, res, limit, settings):
     # linear grid search to set learning rate, lambda, and mini batch size parameters
     if verbose:
         print "Entering grid search to select best parameters for gradient descent."
@@ -253,11 +253,18 @@ def grid_search(verbose, w, data, train, test, res, limit):
     tr.close()
     te.close()
 
-    rate = [0.0001]  # , 0.001]  # , 0.01, 0.1]
-    l = [float(num)]  # [float(num) * 0.1, float(num), float(num) * 10]
-    top_k = [0.8]
-    batch_size = [0.1]  # [num * 4]  # [num * 4, num * 10, num * 100]
-    iterations = [30]  # , 10, 100, 1000]
+    if settings == 'NONE':
+        rate = [0.0001]  # , 0.001]  # , 0.01, 0.1]
+        l = [float(num)]  # [float(num) * 0.1, float(num), float(num) * 10]
+        top_k = [0.8]
+        batch_size = [0.1]  # [num * 4]  # [num * 4, num * 10, num * 100]
+        iterations = [30]  # , 10, 100, 1000]
+    else:
+        rate = settings[0]
+        l = settings[1]
+        top_k = settings[2]
+        batch_size = settings[3]
+        iterations = settings[4]
     accuracy = -1.0             # best accuracy on test data
     parameters = [0, 0, 0, 0, 0]   # best parameter settings
     results = []                # best results
@@ -315,6 +322,7 @@ def parseargs():    # handle user arguments
     parser.add_argument('--limit', default=-1, help='Limit the number of bags to use. Default: No limit.')
     parser.add_argument('--out_dir', default='./', help='Directory to write files to.')
     parser.add_argument('--results', default='results', help='Where to write results to.')
+    parser.add_argument('--settings', default='NONE', help='Parameter settings can be input.')
     parser.add_argument('--test', default='test.dat', help='Where to write test data to.')
     parser.add_argument('--train', default='train.dat', help='Where to write training data to.')
     parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
@@ -328,6 +336,14 @@ def main():
     if args.infile == 'NONE':
         print "Location of input file with data must be specified."
[…]

### 2016-08-03 gicf.py

@@ -148,15 +148,17 @@ def descent(verbose, rate, l, top_k, batch_size, iterations, w, data):
         # find value of derivative of cost function using weights w, then use learning rate to find adjustment amount
         delta_j = rate * derivative_cost(instances, w, cost_data, l, labels)
         # once slope becomes non-negative (or barely negative), stop; or if we exceed number of iterations
-        print "\nDelta J: " + str(delta_j / rate)
-        print "\nRate: " + str(rate)
+        if verbose:
+            print "\nDelta J: " + str(delta_j / rate)
+            print "\nRate: " + str(rate)
 
         # adjust weights according to slope of cost function
         for i in range(1, len(w)):
             adjustment = delta_j[i-1] + (update_vector[i] * momentum)
             w[i] -= adjustment
             update_vector[i] = adjustment
-        print "\nCost: " + str(cost(instances, w, cost_data, l, labels)) + '\n'
+        if verbose:
+            print "\nCost: " + str(cost(instances, w, cost_data, l, labels)) + '\n'
     if verbose:
         print "Gradient descent final weights: "
         print w
@@ -193,7 +195,8 @@ def validate(verbose, w, data, k):     # apply the weights determined by gradien
         average_predicted += i[0]
     average_predicted /= len(predicted_actual)
     threshold = average_predicted
-    print "Average prediction: " + str(average_predicted)
+    if verbose:
+        print "Average prediction: " + str(average_predicted)
 
     true_positives = 0
     false_positives = 0
@@ -231,7 +234,7 @@ def validate(verbose, w, data, k):     # apply the weights determined by gradien
     return accuracy, precision, tpr_recall, fpr, f1_score, predicted_actual
 
 
-def grid_search(verbose, w, data, train, test, res, limit):
+def grid_search(verbose, w, data, train, test, res, limit, settings):
     # linear grid search to set learning rate, lambda, and mini batch size parameters
     if verbose:
         print "Entering grid search to select best parameters for gradient descent."
@@ -253,11 +256,20 @@ def grid_search(verbose, w, data, train, test, res, limit):
     tr.close()
     te.close()
 
-    rate = [0.0001]  # , 0.001]  # , 0.01, 0.1]
-    l = [float(num)]  # [float(num) * 0.1, float(num), float(num) * 10]
-    top_k = [0.8]
-    batch_size = [0.1]  # [num * 4]  # [num * 4, num * 10, num * 100]
-    iterations = [30]  # , 10, 100, 1000]
+    if settings == 'NONE':
[…]

### 2016-08-03 labelDisplay.py

@@ -1,1 +1,10 @@
-__author__ = 'nathan'
+# Author: Nathan LaPierre
+# Date: July 27, 2016
+
+
+def main():
+    pass
+
+
+if __name__ == "__main__":
+    main()

### 2016-08-03 plotter.py

@@ -1,0 +1,166 @@
+# Author: Nathan LaPierre
+# Date: July 29, 2016
+
+import argparse
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+import os
+import shlex
+import subprocess
+import sys
+import time
+
+
+def graph(title, xaxis, yaxis, names, values, results, outdir, verbose):    # graph the plots
+    if verbose:
+        print "Plotting graph: " + title
+    plt.figure(figsize=(8, 6), dpi=100)
+    plt.subplot(111)
+    plt.title(title)
+    plt.xlabel(xaxis[2])
+    plt.xlim(xaxis[0], xaxis[1])
+    plt.ylabel(yaxis[2])
+    plt.ylim(yaxis[0], yaxis[1])
+    for i in range(len(results)):
+        plt.plot(values[i], results[i], label=names[i], linewidth=2.5, linestyle='-', marker='o')
+    plt.legend(loc='upper left', frameon=False)
+    plt.savefig(outdir + title + '.png')
+
+
+def read_results(result_file, metric, verbose):     # read the results from the program runs
+    if verbose:
+        print "Reading results file " + result_file + " for metric " + metric
+    f = open(result_file, 'r')
+    for line in f:
+        if line.startswith(metric):
+            result = float(line.strip().split(metric + ': ')[1])  # retrieve the value of the desired metric
+            f.close()
+            return result
+
+
+def execute(series, outdir, verbose):  # execute necessary program runs
+    if verbose:
+        print "Executing runs for " + str(series)
+    resultdir = outdir + 'results/temp/'
+    params = {'rate': 0, 'lambda': 1, 'top-k': 2, 'mini-batch-size': 3, 'iterations': 4}    # gicf parameters
+    variable = params[series[1].split(' ')[0]]  # variable in the parameter list to vary
+    values = series[1].split(' ')[1].split(',')  # possible values for the variable
+    series[0] += ' --out_dir ' + resultdir  # line to run
+    line = series[0]
+    metrics = series[2].split(',')
+    valuelist = []  # contains full list of x values needed for graphs
+    results = []    # contains full list of y values needed for graphs
+    for m in range(len(metrics)):
+        valuelist.append([])
+        results.append([])
+        for v in range(len(values)):
+            valuelist[m].append(float(values[v]))
+    for i in range(len(values)):
+        # next three lines: isolate settings array, change the variable to desired value, then place back in line
[…]

### 2016-08-04 latex/input/conclusion.tex
finished related work section of paper and added diagrams; began working on results section

@@
 We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study.
-Finally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods.
-Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level labels and validate those labels against clinically verified labels.
+We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability.
+More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods.
+Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level information and validate that information against clinical understanding of the diseases.

### 2016-08-04 latex/input/experiments.tex
finished related work section of paper and added diagrams; began working on results section

@@
 
-\subsection{Data, Software, and Hardware Used}
+\subsection{Experimental Setup}
 
-Our main dataset comes from a GWAS of Type 2 Diabetes (T2D) in Chinese adults, published by Qin et al \cite{qin041012}.
-The data is publicly available on the National Center for Biotechnology Information (NCBI) \cite{ncbi} website using the accession number SRP011011.
-The full dataset contains over 200 patients, but for our study we used a subset of 60 patients due to lengthy download and format conversion times.
-Of these 60 patients, 30 had T2D and 30 were controls.
-For the purpose of testing our cluster sampling methods, we also used a smaller dataset obtained from Dr.
-Patrick Gillevet of GMU, which is not publicly available.
-It contains 904 patients, 239 of whom have encephalopathy and 665 of whom are controls, but is much smaller because it uses 16S rRNA instead of whole genome data.
-The encephalopathy dataset contains about 1.3 million reads, while the T2D dataset contains about 1.3 billion reads.
+The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section.
+Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
+Assembly for each file took 7-30 minutes, depending on the file size, and each patient was assembled in parallel.
+Combining the files into one file took 12 minutes, and clustering with UCLUST took 10 hours and 51 minutes.
+Without assembly first reducing the size of the data, clustering would not have been feasible.
+SVM-light took 10 minutes and 17 seconds to run.
 
-As previously mentioned, we use the publicly-available UCLUST \cite{Edgar10} software for clustering.
-For classification, we use Support Vector Machines (SVMs) \cite{vap95}, one of the most powerful and versatile binary classifiers used
-in myriad applications. 
-%
-For a  binary classification problem, the SVM framework 
-maximizes the separation between the two classes.
-In addition to the ubiquity and effectiveness of SVMs, they are well established in both supervised and multiple instance learning, making it easy to compare the two approaches.
-For our standard supervised SVM, we use svm-light \cite{joachims08}, a popular and publicly available implementation.
-For multiple instance learning, we use the python implementation by G.
-Doran and S.
-Ray ("MISVM") \cite{doran122013} for its ease of use and public availability.
+\subsection{Results For Bag/Patient Labels}
 
-\subsection{Efficient Unsupervised Clustering using UCLUST}
-\label{uclustmethods}
+Temp
 
-From early experimental results, UCLUST can still be computationally infeasible for very large datasets such as the one we will be using, despite its generally excellent speed.
-For our T2D dataset, UCLUST was only able to cluster 2.8\% of the reads within 12 hours, at which point ARGO killed the batch job.
-UCLUST slows down as more clusters are formed, as each new read needs to be compared to more existing clusters.
-Thus, as the number of clusters grows, the execution time of UCLUST increases rapidly.
-The only way to feasibly complete the clustering phase in a reasonable amount of time was to cluster the patients individually in parallel, instead of clustering the reads of all patients together.
-However, this leads to the problem of having mismatched clusterings between patients, that is, cluster X from patient A may represent the same OTU as cluster Y from patient B.
-The solution is to compare the clusters from different patients and map them to each other, and then to combine clusters from different patients that represent the same OTU and output a single clustering output file for all patients.
-However, comparing every read from each cluster to every read from all other clusters is also computationally infeasible.
-Thus, we developed methods to sample representative reads from each patient and use them to create the consensus clustering.
-We explored a number of different methods, but two of them in particular were able to achieve significant speed improvements without significantly harming classification results.
-For each of them, we used a parameter "alpha", which was used to select 1 out of every alpha reads.
-For instance, if alpha=10, 1/10 of reads would be used.
+\subsection{Deriving Instance "Labels"}
 
-We refer to one of our methods as "seed re-clustering" (UCLUST refers to cluster centroids as "seeds").
-In this method, the reads corresponding to the cluster seeds from each patient are copied into a temporary file, and then those reads are clustered.
-This runs quickly, since clusters in this setting tend to have hundreds to thousands of reads, so the number of cluster seeds is relatively small.
-We then construct a map, which maps each seed from its original cluster number to its new cluster number from the seed clustering.
-We go back to the individual patient clustering output, and select 1 out of every alpha lines to be placed in the consensus clustering file.
-As we write the lines to the consensus file, we use the map we developed to change the cluster number for the line from the original clustering number to the new consensus cluster number.
-We refer to the other method as "two round clustering".
-We select one out of every alpha reads from the clustering output of each patient, place them in a single file, and run UCLUST on this smaller subset to arrive at a consensus clustering.
-The main difference between two round clustering and seed re-clustering is that here we select a distribution of reads from the original clustering to determine the consensus clustering, instead of cluster seeds/centroids.
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.5]{./instance-labels.png}
+\caption{This diagram illustrates why static instance labels are not sufficient for phenotype prediction.
+A patient with 6 of the blue microbe or 6 of the green microbe may be healthy, while a patient with 3 of each is sick.
+Static instance labels cannot capture this relationship.
+This is also explored by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
+\end{figure}
 
-We sought to empirically evaluate how these methods affected clustering runtime and classification accuracy, since feature vectors are generated from the clustering output and thus classification results are affected by the clustering methods.
-Without use of these sampling methods, UCLUST didn't finish within ARGO's 12 hour time limit on the T2D dataset, so we couldn't evaluate the classification results on that dataset and compare them to the classification results with the clustering sample methods.
-In order to evaluate these methods, we used the aforementioned Encephalopathy dataset, which is much smaller.
-We measured the time it took to perform UCLUST on all patients at once, as well as the time taken by each of our two sampling methods.
-We then used the respective clustering outputs to generate feature vectors and run svm-light classification (as described in the next section) and compared the results, which are discussed in the "Experimental Results" section.
-For UCLUST, default settings were used, except the threshold for placing two sequences in the same cluster was set to 40\% match (--id 0.40), and the --usersort option was used.
+One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance "labels".
+In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other.
+For instance, a patient with X amount of microbe A or X amount of microbe B, but not with X/2 amount of microbe A and X/2 amount of microbe B.
+This is illustrated in Figure \ref{instance-labels}.
+This example is very simplified, but it explains why static instance labels are insufficient.
 
-\subsection{Supervised and Multiple Instance Classification using Support Vector Machines (SVMs)}
-\label{svm}
-
-For our formulation we use SVMs to make clinical phenotype prediction for each patient sample based on their  input metagenome sequence samples represented with their clustering or taxa membership features.
-With our clustering approach,  we have a set of reads that have been assigned to clusters based on UCLUST and our techniques.
-We now generate feature vectors from each patient from this clustering output.
-We follow a different process of feature vector generation for svm-light and MISVM.
-For each method, we read through the consensus clustering output from UCLUST line-by-line, forming feature vectors as we go.
-UCLUST output lines provide the cluster number, the percentage match to the cluster seed, and the patient that read comes from.
-
-For svm-light, we have one feature vector per patient.
-We use each cluster/OTU as a feature, meaning that the presence of various microbes are the features for the patient.
-Each number (feature) in the vector corresponds to a cluster/OTU, and the value for that number corresponds to how closely the read from the patient matched the cluster seed.
-For each input
-sequence, we use the percentage match to its cluster seed as 
-the value for that feature and average the 
-sequence match if multiple reads from the metagenome samples 
-are assigned to the same cluster by UCLUST.
-Each sequence read includes a label for the patient that it came from, so each read only affects the feature vector corresponding to the patient that it came from.
-For example, say patient A has a read with 95\% match to OTU 2, a read that is the cluster seed of OTU 3, a read that is a 93\% match to OTU 2, a read that is a 97\% match to OTU 5, and no reads for OTU 1 or 4.
-Patient A's feature vector would then look like: [0 0.94 1 0 0.97].
-We first initialize the patient feature vectors to all 0s and then iterate through the UCLUST output lines and update the feature vectors as we go.
-For MISVM, patients are bags, reads for patients are instances, and each read/instance has a feature vector.
-We iterate through the UCLUST output.
-For each read, we add that read as an instance to the relevant patient bag.
-The feature vector for that read contains one number: the cluster number that the read belongs to.
-We compared the runtime and accuracy of the two implementations, which both used a linear kernel.
-In both cases, the training and test set each contained 30 patients, and each contained 15 patients with type 2 diabetes and 15 control (healthy) patients.
-The results are discussed in the "Experimental Results" section.
+However, we can infer from the SVM decision boundary which clusters appear to be most relevant to the disease diagnosis.
+Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis.
+For instance, if the Ith scalar in the weight vector is has the highest value of any of the weights, then cluster I is likely to play a major role in the disease.
+Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease.
+Because the data is metagenomic, the clusters represent both phylogenetic and functional similarity, so identifying the most relevant clusters can help discover more about the pathology of the disease.
+For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.

### 2016-08-04 latex/input/intro.tex
finished related work section of paper and added diagrams; began working on results section

@@
 These reads, represented as strings of nucleotides, represent only small parts of the microbe's full genome, and are not ordered in any way, which presents several challenges that will be discussed further in the paper.
-Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (phylogeny), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes \cite{handelsman04}.
+
+Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (\emph{phylogeny}), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes (\emph{functional} information) \cite{handelsman04}.
+Summarily, metagenomics allows us to view microbial data that is not accessible to us via traditional laboratory culturing and allows for both phylogenetic and functional profiling of those microbes.
 Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.
@@
 We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype.
-This data is very large (hundreds of gigabytes) and high-dimensional (thousands of dimensions).
+This data is very large (multiple terabytes) and high-dimensional (thousands of dimensions).
 Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled.
@@
 We show that our classifier shows significantly improved performance.
-This is discusses further in the Experimental Results section.
+This is discussed further in the Experimental Results section.
 

### 2016-08-04 latex/input/materials.tex
finished related work section of paper and added diagrams; began working on results section

@@
 The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}.
-Each patient file was several gigabytes in size, with the total dataset size being over 600 gigabytes.
+The total size of these 218 FASTQ files was 2.3 terabytes, with an average size of 10.82 gigabytes per patient file.
 Qin et al.
-develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "e10171", respectively \cite{qin041012}.
+develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "1071", respectively \cite{qin041012}.
 They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}.
@@
 \subsection{Software and Hardware Details}
-We used the ARGO computing cluster available at George Mason University \cite{argo}.
+We used the ARGO computing cluster available at George Mason University\footnote{http://orc.gmu.edu/research-computing/argo-cluster/argo-hardware-specs/}.
 The clustering and classification phases were run on one of the compute nodes available on the cluster.
 The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM.
-(Total Cores 528 and 1056 total threads, RAM$>$2TB) \cite{argo}.
+(Total Cores 528 and 1056 total threads, RAM$>$2TB).
 Source codes for
 SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
-their respective websites
- and compiled on the Argo platform.
+their respective websites and compiled on the Argo platform.

### 2016-08-04 latex/input/related.tex
finished related work section of paper and added diagrams; began working on results section

@@
 
-The \emph{assembly} problem involves combining overlapping short reads into longer sequences called \emph{contigs}.
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{./de-bruijn.png}
+\caption{This diagram illustrates a basic de Bruijn graph, as explained in the text.
+Notice the double arrow from TAT to ATT, allowing a path to go through that route twice.
+Thus, the longest contig that can be created by following a path is GTATTATTGGA, which is what we want.} \label{de-bruijn}
+\end{figure}
+
+The \emph{assembly} problem involves combining overlapping short reads (usually less than 1000 base pairs) into longer sequences called \emph{contigs} (often tens of thousands of base pairs).
+For instance, if one read ends with the same relatively large nucleotide string that another read starts with, the reads are likely to be overlapping fragments from the same genome, and can thus be combined into one contig.
+This can be done either \emph{de novo} (in an unsupervised manner) or by referencing sequences against known contigs.
+We focus on de novo assembly, because we wish to keep our pipeline as unsupervised as possible.
+
+One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}.
+Even if full genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classifiers, because the contigs will contain more phylogenetic and functional information than short reads.
+This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads.
+Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads.
+Metagenomics additionally poses its own set of challenges, due to very large datasets and lack of knowledge about how many species are present and in what relative abundances \cite{namiki12}.
+Thus, metagenome assembly is a relatively new and challenging field.
+
+Some of the popular genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
+We used SOAPdenovo2 in our study, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}.
+SOAPdenovo2 first constructs a type of directed graph called a \emph{de Bruijn} graph that represents the overlaps between different sequences \cite{li10}.
+Reads are divided into strings of length K called \emph{k-mers}; these k-mers are the nodes of the graph \cite{zerbino08}.
+The choice of K is up to the user, and is important for having good assembly results.
+The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}.
+SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating "tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}.
+The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.
+There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.
 
@@
 
-The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are 
-similar to each other. This process may lead to groups that are organism-specific and is 
-unsupervised in nature. 
-%
-The clustering process does not attempt to provide an automated labelling of the input reads. 
-%
-The clusters/groups obtained from an input metagenome sample 
-are referred to as  Operational Taxonomic Units (OTUs), and the number of 
-OTUs gives an approximation of species diversity in a sample \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
-%
-These  approaches are not constrained due to the absence of a complete coverage in taxonomic databases. Some 
-environmental samples contain microbial organisms that have never been cultured in a laboratory, and thus those organisms do not exist in genomic databases.
-As such,  clustering of 
-sequence reads has several advantages: (i)
-        it can lead to an improved  metagenome assembly, (ii) it can be used 
-        for computing species diversity metrics \cite{bibm2012} 
-        and (iii)
-        the  reduced computational
-        complexity within several work-flows that analyze only
-        cluster representatives, instead of individual sequences
-        within a sample.
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{./uclust-similarity.png}
+\caption{This diagram illustrates the similarity computation between two sequences in UCLUST.
+Terminal characters are excluded, so the length of the reads is 10.
+In the first 10 characters of each string, there are 5 matching characters that are in the same position in the string.
+Thus, the similarity is 5/10 = 50\%.} \label{uclust-similarity}
+\end{figure}
 
-CD-HIT \cite{Li01072006}, UCLUST \cite{Edgar10}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a} and 
-MC-LSH \cite{bibm2012}
-are some of the popular 
-metagenome sequence clustering approaches used for binning. 
-%
-UCLUST, MC-MinH and MC-LSH  are
-greedy approaches that achieve computational efficiency 
-by using either hash-based indexing or matching of gap-less sequences called seeds (instead 
-of expensive sequence alignment) and followup with an incremental clustering approach 
-that does not involve comparing all pairs of input sequences.
+The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are similar to each other.
+The clusters obtained from this process are referred to as Operational Taxonomic Units (OTUs).
+OTUs represent a group of equivalent or similar organisms.
+Accordingly, the number of OTUs in a sample gives an approximation of the species diversity in that sample  \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
+In addition to approximating species diversity, clustering has several other key advantages.
+Because clustering is always de novo (unsupervised), it is not limited by the species that are covered in taxonomic databases.
+This detail is important, because it is believed that most microorganisms that reside in the human body have not been laboratory cultured \cite{handelsman04}.
+Clustering also reduces computational costs by allowing analyses to operate on entire clusters instead of on each read.
+Finally, clustering helps the classification process by allowing feature vectors to be built at the OTU level, instead of using individual short reads that contain little biological information.
+
+UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a}, and 
+MC-LSH \cite{bibm2012} are some of the popular  nucleotide sequence clustering approaches.
 We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}.
@@
 
-UCLUST \cite{Edgar10} follows a greedy, iterative clustering approach. As a first 
-step, this approach identifies exact matches of fixed length between sequence pairs known 
-as seeds. These seeds are then extended by allowing for a few mismatches and/or gaps between 
-the aligned pairs.
-Seeds scoring above a certain threshold are chosen as high segment pairs (HSPs)  and used for further processing.
-As such, the step eliminates a lot of pairwise comparisons.
-Then UCLUST follows an incremental approach for assigning the input sequences to the different bins.
-The clustering solution is initialized as an empty list.
-Sequences are then incrementally added to the clusters existing within the list.
-Each unassigned input sequence is compared to the cluster representatives within the list using the fast indexing search technique that uses the HSPs.
-If a match is found with one of the cluster representatives, then that sequence is assigned to that particular cluster; otherwise, the input sequence forms a new cluster.
-Matches are computed by comparing two reads, character by character, and dividing the number of characters at the same position in both reads that match by the length of the longer of two reads.
-UCLUST ensures that the cluster representatives are sequences with the largest length.
+UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}.
+Thus, each \emph{centroid} defines the center of a cluster, and the distance T defines the radius of the cluster, such that any point that has a similarity of greater than T to the centroid is within the radius and is thus part of the cluster.
+UCLUST is a heuristic algorithm that has several optimizations to improve speed, thus condition (i) above is not always guaranteed \cite{Edgar10}.
+UCLUST proceeds in a greedy, iterative manner.
+The first sequence in the input file becomes a new cluster centroid.
+For each new sequence in the file, it is compared with each of the existing cluster centroids in order.
+As soon as it is compared with a centroid that it has a similarity of greater than T with, it becomes part of that cluster.
+If the read is not similar enough with any of the existing cluster centroids, it becomes the centroid of a new cluster.
+The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}.
+This is illustrated in figure \ref{uclust-similarity}.

### 2016-08-04 clean.py

@@ -1,0 +1,42 @@
+import argparse
+import time
+import sys
+
+
+def parseargs():    # handle user arguments
+    parser = argparse.ArgumentParser(description='Clean unwanted whitespace out of FASTA files.')
+    parser.add_argument('-i', '--infile', default='NONE', help='Location of input file')
+    parser.add_argument('-o', '--output', default='NONE', help='Where to write output to')
+    parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
+    args = parser.parse_args()
+    return args
+
+
+def main():
+    start = time.time()
+    args = parseargs()
+    if args.infile == 'NONE' or args.output == 'NONE':
+        print "Location of input and output files must be specified."
+        sys.exit()
+    infile = open(args.infile, 'r')
+    output = open(args.output, 'w')
+    read = ''
+    for line in infile:
+	if line.startswith('>'):
+	    if read != '':
+		output.write(read + '\n')
+		read = ''
+	    output.write(line)
+	else:
+	    read += ''.join(line.split())
+    if read != '':
+	output.write(read + '\n')
+    infile.close()
+    output.close()
+    if args.verbose:
+	print "Program execution time: " + str(time.time() - start) + " seconds"
+
+
+if __name__ == "__main__":
+    main()
+

### 2016-08-08 latex/input/experiments.tex
updated paper results; pipeline now can be d-bow or h-bow

@@
 Assembly for each file took 7-30 minutes, depending on the file size, and each patient was assembled in parallel.
-Combining the files into one file took 12 minutes, and clustering with UCLUST took 10 hours and 51 minutes.
-Without assembly first reducing the size of the data, clustering would not have been feasible.
-SVM-light took 10 minutes and 17 seconds to run.
+Combining the files into one file took 12 minutes.
+Assembly was used for all of the classification methods tested, because the combining of individual reads and reduction in total data size made classification feasible.
+
+\subsection{Methods Tested}
+
+This section provides an overview of the different methods that we compared our pipeline to.
+Our pipeline uses clustering, whereas the other methods do not.
+Instead, those methods directly compare individual instances instead of clusters.
+In order to do this, sequence reads were represented as counts of k-mers.
+k-mers are nucleotide strings of length k.
+Since there are four possible nucleotides, the number of possible k-mers is \(4^k\).
+For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT.
+For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else.
+We wrote a script to represent the reads as vectors representing the k-mer counts in the string, and these vectors were used by the other methods.
+From experimental validation, we found a k-mer value of 3 to be the most effective.
+
+\subsubsection{CAMIL: Our Pipeline}
+
+We refer to our pipeline as "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
+We implemented two different versions of this pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction.
+These results are denoted as "CAMIL D-BoW" and "CAMIL H-BoW", respectively, in the results tables and graphs.
+Clustering for the pipeline with UCLUST took 10 hours and 51 minutes.
+Without the assembly step reducing the size of the data, clustering would not have been feasible.
+Feature extraction and SVM-light classification took 10 minutes and 17 seconds to run for D-BoW, versus 5 minutes and 14 seconds for H-BoW.
+
+\subsubsection{MI-SVM and sbMIL}
+
+MI-SVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the "Instance Space" (IS) methods, in that they only use "local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}.
+Additionally, both of these methods follow the standard MIL assumption that bags with negative labels contain only negative instances, whereas positive bags contain one or more positive instances \cite{amores13}.
+sbMIL specifically assumes that positive bags contain few positive instances \cite{bunescu07}.
+We include these algorithms as an example of many of the early MIL algorithms, which usually fell into the IS paradigm and used the standard MIL assumption.
+For the implementation of these methods, we used an open-source Python implementation by Doran \cite{doran14}, which is available on GitHub\footnote{https://github.com/garydoranjr/misvm}.
+
+\subsubsection{GICF}
+
+The Group-Instance Cost Function (GICF) is a method proposed by Kotzias et al. that learns instance labels in addition to group labels \cite{kotzias15}. The cost function
+
+\subsubsection{Original MGWAS Paper: mRMR + SVM}
+
+Temp
 

### 2016-08-08 latex/input/intro.tex
updated paper results; pipeline now can be d-bow or h-bow

@@
 The rest of the paper is organized as follows.
-Section 2 presents relevant Background information on  Multiple Instance Learning.
-Section 3 presents the methods we used in the creation of our pipeline.
+Section 2 presents relevant Background information on  Multiple Instance Learning, Assembly, and Clustering.
+Section 3 presents the Methods we used in the creation of our pipeline.
 Section 4 presents Materials, such as dataset, hardware, and software descriptions.
-Section 5 presents our results, and Section 6 presents our conclusions.
+Section 5 presents our Results, and Section 6 presents our Conclusions.

### 2016-08-08 latex/input/materials.tex
updated paper results; pipeline now can be d-bow or h-bow

@@
 of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}.
-The total dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online.
+This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies.
+The full dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online.
 Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}

### 2016-08-08 latex/input/related.tex
updated paper results; pipeline now can be d-bow or h-bow

@@
 In the original paper, Dietterich develops a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}.
-In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, and MILES \cite{wang06}.
+In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, sbMIL \cite{bunescu07}, and MILES \cite{wang06}.
 A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.

### 2016-08-08 pipelineT2D.py

@@ -13,6 +13,7 @@ import misvm
 
 def parseargs():    # handle user arguments
     parser = argparse.ArgumentParser(description='Classify fasta-format sequence reads')
+    parser.add_argument('--bow', default='d', choices=['d', 'h'], help='Distance (d) or Histogram (h) Bag of Words.')
     parser.add_argument('-c', '--cluster', default='NONE', choices=['kraken', 'uclust'],
                         help='Which clustering algorithm was used')
     parser.add_argument('-d', '--dir', default='NONE', help='You can specify directory of all parameters if shared')
@@ -63,7 +64,7 @@ def parse_kraken(vectors, infile, split):  # generate feature vectors from krake
     return vectors
 
 
-def parse_uclust(vectors, infile, split):  # generate feature vectors from uclust output
+def parse_uclust(vectors, infile, split, bow):  # generate feature vectors from uclust output
     f = open(infile)
     for line in f:
         if line.startswith('H') or line.startswith('S'):
@@ -76,17 +77,24 @@ def parse_uclust(vectors, infile, split):  # generate feature vectors from uclus
                 if veclen < featnum + 1:        # used to expand feature list if needed
                     for j in range(veclen, featnum + 1):
                         vec.append('')
-                if line.startswith('H'):
-                    value = str(float(fields[3])/100.0)   # "value" of feature, i.e. percentage match to cluster seed
-                    if vec[featnum] == '' or float(value) > float(vec[featnum].split(":")[1]):
-                        vec[featnum] = str(featnum) + ":" + value + ' '
-                elif line.startswith('S'):
-                    vec[featnum] = str(featnum) + ":1.0 "
+                if bow == 'd':
+                    if line.startswith('H'):
+                        value = str(float(fields[3])/100.0)   # "value" of feature i.e. percentage match to cluster seed
+                        if vec[featnum] == '' or float(value) > float(vec[featnum].split(":")[1]):
+                            vec[featnum] = str(featnum) + ":" + value + ' '
+                    elif line.startswith('S'):
+                        vec[featnum] = str(featnum) + ":1.0 "
+                elif bow == 'h':
+                    if line.startswith('H') or line.startswith('S'):
+                        if vec[featnum] == '':
+                            vec[featnum] = str(featnum) + ":1.0 "
+                        else:   # add to the tally for this cluster
[…]

### 2016-08-09 latex/input/abstract.tex
finished paper draft #1

@@
 The recent advent of Metagenome-Wide Association Studies (MGWAS) has allowed for increased accuracy in the prediction of patient phenotype (disease), but has also presented big data challenges.
-Meanwhile, Multiple Instance Learning (MIL) is useful in the domain of bioinformatics because, in addition to classifying patient phenotype, it can also identify individual parts of the metagenome that are indicative of that phenotype, leading to better understanding of the disease.
+Meanwhile, Multiple Instance Learning (MIL) is useful in the domain of bioinformatics because, in addition to classifying patient phenotype, it can also identify individual parts of the microbiome that are indicative of that phenotype, leading to better understanding of the disease.
 We demonstrate a novel, efficient, and effective MIL-based computational pipeline to predict patient phenotype from MGWAS data.
@@
 This involves assembly of the metagenomic data, clustering of the assembled contigs, extracting features from the contigs, and using a standard SVM classifier to predict patient labels and identify the most relevant read clusters.
-With the exception of the given labels for the patients, this entire process is \emph{de novo}.
-We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper.
+With the exception of the given labels for the patients, this entire process is \emph{de novo} (unsupervised).
+We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper, as well as other common MIL methods.
+We call our pipeline "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.

### 2016-08-09 latex/input/conclusion.tex
finished paper draft #1

@@
 More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods.
-Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level information and validate that information against clinical understanding of the diseases.
+CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement.
+Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

### 2016-08-09 latex/input/experiments.tex
finished paper draft #1

@@
 The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section.
+The data was split evenly into training and test sets, with each containing 109 patients.
+The training set had 54 T2D patients and 55 control patients, while the test set had 53 T2D patients and 56 control patients.
 Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
@@
 From experimental validation, we found a k-mer value of 3 to be the most effective.
+Converting the reads from their original string form to k-mer vectors took 3 hours, 17 minutes, and 55 seconds.
 
-\subsubsection{CAMIL: Our Pipeline}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-We refer to our pipeline as "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
-We implemented two different versions of this pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction.
+\subsubsection{CAMIL - Our Pipeline}
+
+We implemented two different versions of the pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction.
 These results are denoted as "CAMIL D-BoW" and "CAMIL H-BoW", respectively, in the results tables and graphs.
@@
 
-\subsubsection{MI-SVM and sbMIL}
+\subsubsection{MISVM and sbMIL}
 
-MI-SVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the "Instance Space" (IS) methods, in that they only use "local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}.
+MISVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the "Instance Space" (IS) methods, in that they only use "local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}.
 Additionally, both of these methods follow the standard MIL assumption that bags with negative labels contain only negative instances, whereas positive bags contain one or more positive instances \cite{amores13}.
@@
 
-The Group-Instance Cost Function (GICF) is a method proposed by Kotzias et al. that learns instance labels in addition to group labels \cite{kotzias15}. The cost function
+The Group-Instance Cost Function (GICF) is a method proposed by Kotzias et al.
+that learns instance labels in addition to group labels \cite{kotzias15}.
+The cost function uses a kernel that measures similarity between instances and a penalty on the difference between instance labels to generate instance labels \cite{kotzias15}.
+It then sets the group label to be the average instance label of all instances in that group, using a penalty on the difference between the predicted group label and the actual group label \cite{kotzias15}.
+Ideally, this would cause instances that are similar to each other to have similar labels, and predicted group labels to correspond closely to reality.
+Unlike MISVM and sbMIL, GICF explicitly does not hold the standard MIL assumption, instead favoring the collective assumption.
+However, because this method compares only individual instances and not entire bags, and treats bag labels simply as aggregations of instance labels, GICF is still an Instance Space method.
 
-\subsubsection{Original MGWAS Paper: mRMR + SVM}
+\subsubsection{Original MGWAS Paper}
 
-Temp
+The methods in the original paper are neither MIL-based, nor are they entirely de novo apart from patient labels.
+The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then use a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}.
+They then combined these genes with an existing gene catalog, MetaHIT \cite{qin030410}, and carried out taxonomic assignment and functional annotation of the genes using the KEGG \cite{kanehisa00} and eggNOG \cite{powell12} databases, as well as 2,890 other reference genomes \cite{qin041012}.
+The authors defined gene markers by mapping the sequence reads from the MGWAS dataset to the updated gene catalog.
+They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the "sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the "e1071" R package for the SVM \cite{qin041012}.
+Thus, the method in the original paper first applies de novo assembly and gene prediction methods, but then uses a number of references to identify the gene markers to be used in classification.
+From their results, the authors generated an Area Under Curve - Receiver Operating Characteristic (AUC-ROC) graph.
+We manually computed the accuracy and F1 score with an ideally-chosen decision boundary based on the results the authors provided in their supplementary tables.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Temp
+\begin{table}[h]
+\begin{center}
+\caption{Comparison of the performance of various methods on the MGWAS dataset. 
+\label{tab:comp}}
+\begin{tabular}{|c|ccc|}\hline
+Method & Accuracy & F1-Score & AUC-ROC\\\hline
+MISVM & 50.00 & 50.00 & 50.00\\\hline
+sbMIL & 50.00 & 50.00 & 50.00\\\hline
+GICF & 60.00 & 71.43 & 66.00\\\hline
+Original & 82.61 & 80.02 & 81.00\\\hline 
+CAMIL H-BoW & 91.74 & 91.43 & 95.00\\\hline
+CAMIL D-BoW & \bf{92.66} & \bf{92.31} & \bf{96.00}\\\hline
+\end{tabular}
+\end{center}
+\end{table}
+
+Table \ref{tab:comp} shows the comparison of classification performance between the various methods described above.
+CAMIL methods significantly outperform the other methods, with the D-BoW variant of CAMIL slightly outperforming the H-BoW variant.
+
+MISVM and sbMIL had the worst results, faring no better than a random guess.
+This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
+Additionally, they are instance space methods that do not leverage bag-level information.
+The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain.
+GICF performs better than MISVM and sbMIL, which makes sense given the fact that it follows the collective assumption.
+It also has the benefit of calculating instance labels, which we explore further in the next section.
+However, GICF is still an instance space method, so it makes sense that it performs worse than CAMIL.
+
+The method used in the original paper is the only one tested here that is not an MIL method, and the only one that is not entirely unsupervised apart from the patient labels.
+Given that the methods from this paper were not de novo, it makes sense that they would outperform many of the unsupervised MIL methods.
+However, CAMIL still significantly outperformed the results reported in the original paper.
+We believe that this is due to the following reasons: (i) the clustering process puts similar contigs into groups that are useful features for the classifier, and (ii) instead of attempting to select the most significant features before performing classification, we allow the classifier itself to determine the most significant features from the entire pool of features.
 
@@
 Static instance labels cannot capture this relationship.
-This is also explored by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
+This is also explained by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{./instance-scatter.png}
+\caption{This diagram illustrates the distribution of the instance weights assigned by CAMIL D-BoW.
+The Y-Axis shows the weights, while the X-Axis shows the ranking of the clusters by weight.
+Clearly, there are a relatively small number of clusters that have disproportionately large weights, while there is a smaller number of clusters with smaller weights.} \label{instance-scatter}
 \end{figure}
@@
 In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other.
-For instance, a patient with X amount of microbe A or X amount of microbe B, but not with X/2 amount of microbe A and X/2 amount of microbe B.
+For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick.
 This is illustrated in Figure \ref{instance-labels}.
@@
 Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis.
-For instance, if the Ith scalar in the weight vector is has the highest value of any of the weights, then cluster I is likely to play a major role in the disease.
+For instance, if the Ith scalar in the weight vector is has the highest value of any of the weights, then cluster I is likely to play a major role in the disease pathology.
 Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease.
@@
 For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.
+
+In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier.
+Conversely, CAMIL uses all of the data to train and test the classifier, and subsequently identifies significant clusters based on the classification results.
+We printed the 25 highest-weighted and 25 lowest-weighted clusters, out of 3188 total clusters.
+The highest-weighted cluster had a weight of 0.012109, while the lowest-weighted cluster had a weight of -0.003668.
+The average of the 25 highest cluster weights was 0.010402, while it was -0.002753 for the lowest 25.
+The average of all 3188 cluster weights was 0.001003, and the median was 0.000254.
+These results indicate that the highest weights were much larger than the lowest weights and skewed the average weight to be much higher than the median weight.
+Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters aren't particularly relevant in this case and whose weights are just noise.
+Thus, the weights obtained by this method appear to be plausible.
+In contrast, the labels obtained by GICF were barely differentiated from each other at all.
+The instance weights determined by CAMIL are shown in Figure \ref{instance-scatter}.

### 2016-08-09 latex/input/intro.tex
finished paper draft #1

@@
 Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}.
+We also show that our pipeline can be used for Histogram-based Bag of Words (H-BoW) methods.
 MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome.
-However, since datasets in this domains frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.
+However, since datasets in this domain frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.
 
 We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype.
-This data is very large (multiple terabytes) and high-dimensional (thousands of dimensions).
+Because MGWAS studies contain many expert-labeled patients and the metagenomic data associated with those patients, they are very useful for phenotype prediction but also cause many computational challenges.
+The data is very large (multiple terabytes) and high-dimensional (thousands of dimensions).
 Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled.
@@
 Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully.
-Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads.
+Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads and many of them are due to sequencing machine errors.
 The reduction in data size also allows for clustering, which is not feasible for massive datasets.
@@
 This process is explained in further detail in the methods section.
+We refer to the pipeline as "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
 We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data.

### 2016-08-09 latex/input/related.tex
finished paper draft #1

@@
 If even one shape of a given molecule binds to the binding site, it is considered a "good" molecule \cite{dietterich97}.
-Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains all negative instances.
+Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains only negative instances.
+This is commonly referred to as the "standard multiple instance assumption" \cite{amores13}.
 In the original paper, Dietterich develops a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}.
@@
 A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.
+Most of these classic methods follow the standard assumption, which is harmful in domains in which negative bags may contain some positive instances.
 %
 
-However, since 2010, there has been increased interest in different formulations of the MIL problem.
+More recently, there has been increased interest in different formulations of the MIL problem.
 For instance, the problem of "key instance detection" \cite{zhou12} revolves around finding the instances that contribute the most to bag labels.
@@
 This is significant in metagenomics because, while some diseases are caused by a single pathogen, many arise from a combination of many factors, and even patients that are "healthy" may contain small amounts of pathogens that are normally associated with disease.
+In contrast with the standard assumption, this can be referred to as the "collective" assumption \cite{amores13}.
+Additionally, it is helpful to discover which microbes and which functional attributes of those microbes lead to disease, making instance level information significant in this domain.
 %
 
-However, this method treats a group label simply as an aggregation of instance labels.
-Amores, in his taxonomy, called this type of method "Instance-based", and found this paradigm to be ineffective \cite{amores13}.
-Amores identifies two other paradigms, Bag-Space and Embedded-Space.
-An example of the latter are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
+While some of the above methods follow the standard assumption and others follow the collective assumption, they all treat bag labels simply as aggregations of instance labels and thus focus only on comparisons between individual instances and not entire bags or groups of instances.
+For methods following the standard assumption, the aggregation function is simply an OR function: if any of the instances in a bag are positive, the entire bag is positive.
+For methods following the collective assumption, the aggregation function is often based on an averaging of instance labels.
+Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as "Instance Space" by Amores, and he showed that this paradigm was generally significantly less effective than the two other paradigms, "Bag Space" and "Embedded Space" \cite{amores13}.
+Bag Space methods define a distance or kernel function that determine the similarity between bags, while Embedded Space methods map bags into feature vectors which can then be used for classifiers \cite{amores13}.
+Embedded Space methods can be further divided into two subcategories: methods that simply aggregate information about all instances in a bag without differentiating them, and "Vocabulary-based" methods that group certain similar instances together and then use those groups to form the feature vector \cite{amores13}.
+We use a vocabulary-based method in this paper, because having information about groups of similar sequence reads can be biologically important, as explained further in the clustering section.
+
+An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
 Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient \cite{amores13}.
 The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center.
-Despite these recent developments in MIL, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.
+Histogram-based Bag of Words (H-BoW) methods count the number of instances in a given cluster instead of keeping track of the closest-matching instance to that cluster.
+H-BoW methods were found to be among the more effective methods, but not quite as good as D-BoW \cite{amores13}.
+
+As we have discussed, the Multiple Instance paradigm fits phenotype prediction well, since we have a set of labeled patients containing unlabeled sequence reads, and we would like to predict both the patient phenotype and which reads are indicative of that phenotype.
+Despite the recent developments in MIL and its potential utility in phenotype prediction, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.
 

### 2016-08-16 latex/input/abstract.tex
paper draft v1.1: most minor edits completed

@@
 Specifically, we use a Distance-based Bag of Words method, which has been shown to be one of the most effective and efficient MIL methods.
-This involves assembly of the metagenomic data, clustering of the assembled contigs, extracting features from the contigs, and using a standard SVM classifier to predict patient labels and identify the most relevant read clusters.
+This involves assembly of the metagenomic sequence data, clustering of the assembled contigs, extracting features from the contigs, and using an SVM classifier to predict patient labels and identify the most relevant read clusters.
 With the exception of the given labels for the patients, this entire process is \emph{de novo} (unsupervised).
 We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper, as well as other common MIL methods.
-We call our pipeline "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
+We call our pipeline ``CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.

### 2016-08-16 latex/input/experiments.tex
paper draft v1.1: most minor edits completed

@@
 Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
-Assembly for each file took 7-30 minutes, depending on the file size, and each patient was assembled in parallel.
+Assembly for each file took 7-30 minutes, depending on the file size.
+The assembly of the patient files is embarrassingly parallel, so the assembly of each file could be done at the same time.
 Combining the files into one file took 12 minutes.
@@
 We implemented two different versions of the pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction.
-These results are denoted as "CAMIL D-BoW" and "CAMIL H-BoW", respectively, in the results tables and graphs.
+These results are denoted as ``CAMIL D-BoW" and ``CAMIL H-BoW", respectively, in the results tables and graphs.
 Clustering for the pipeline with UCLUST took 10 hours and 51 minutes.
@@
 
-MISVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the "Instance Space" (IS) methods, in that they only use "local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}.
+MISVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the ``Instance Space" (IS) methods, in that they only use ``local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}.
 Additionally, both of these methods follow the standard MIL assumption that bags with negative labels contain only negative instances, whereas positive bags contain one or more positive instances \cite{amores13}.
@@
 It then sets the group label to be the average instance label of all instances in that group, using a penalty on the difference between the predicted group label and the actual group label \cite{kotzias15}.
-Ideally, this would cause instances that are similar to each other to have similar labels, and predicted group labels to correspond closely to reality.
+Ideally, this would cause instances that are similar to each other to have similar predicted labels, and predicted group labels to correspond closely to reality.
 Unlike MISVM and sbMIL, GICF explicitly does not hold the standard MIL assumption, instead favoring the collective assumption.
@@
 The authors defined gene markers by mapping the sequence reads from the MGWAS dataset to the updated gene catalog.
-They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the "sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the "e1071" R package for the SVM \cite{qin041012}.
+They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the ``sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the ``e1071" R package for the SVM \cite{qin041012}.
 Thus, the method in the original paper first applies de novo assembly and gene prediction methods, but then uses a number of references to identify the gene markers to be used in classification.
@@
 GICF & 60.00 & 71.43 & 66.00\\\hline
-Original & 82.61 & 80.02 & 81.00\\\hline 
+mRMR + SVM & 82.61 & 80.02 & 81.00\\\hline 
 CAMIL H-BoW & 91.74 & 91.43 & 95.00\\\hline
@@
 
-\subsection{Deriving Instance "Labels"}
+\subsection{Deriving Instance ``Labels"}
 
@@
 
-One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance "labels".
+One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance ``labels".
 In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other.
@@
 This is illustrated in Figure \ref{instance-labels}.
-This example is very simplified, but it explains why static instance labels are insufficient.
+This example is simple, but it explains why static instance labels are insufficient.
 
@@
 Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis.
-For instance, if the Ith scalar in the weight vector is has the highest value of any of the weights, then cluster I is likely to play a major role in the disease pathology.
+For instance, if the \emph{i}th scalar in the weight vector is has the highest value of any of the weights, then cluster \emph{i} is likely to play a major role in the disease pathology.
 Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease.
@@
 These results indicate that the highest weights were much larger than the lowest weights and skewed the average weight to be much higher than the median weight.
-Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters aren't particularly relevant in this case and whose weights are just noise.
+Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters are not particularly relevant in this case and whose weights are just noise.
 Thus, the weights obtained by this method appear to be plausible.

### 2016-08-16 latex/input/intro.tex
paper draft v1.1: most minor edits completed

@@
 
-Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (\emph{phylogeny}), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes (\emph{functional} information) \cite{handelsman04}.
+Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health \cite{handelsman04}; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown \cite{handelsman04}; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (\emph{phylogeny}), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes (\emph{functional} information) \cite{handelsman04}.
 Summarily, metagenomics allows us to view microbial data that is not accessible to us via traditional laboratory culturing and allows for both phylogenetic and functional profiling of those microbes.
@@
 In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome.
-This can be viewed as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them.
+We view this as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them.
 In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances).
@@
 We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype.
-Because MGWAS studies contain many expert-labeled patients and the metagenomic data associated with those patients, they are very useful for phenotype prediction but also cause many computational challenges.
+MGWAS studies contain many expert-labeled patients and the metagenomic data associated with those patients, so they are useful for phenotype prediction but also cause many computational challenges.
 The data is very large (multiple terabytes) and high-dimensional (thousands of dimensions).
 Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled.
-\emph{Assembly} is the process of finding pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome, and combining them.
+\emph{Assembly} is the process of combining pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome.
 This process is repeated as much as possible to form long strings called \emph{contigs}.
 Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully.
-Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads and many of them are due to sequencing machine errors.
 The reduction in data size also allows for clustering, which is not feasible for massive datasets.
-Clustering uses string similarity measures to group similar reads into "clusters", which is a way of identifying which species of microbe each read corresponds to.
-The alternative, "aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced.
-From the clustering output, we extract feature vectors, which are then fed into a standard Support Vector Machine (SVM) classifier.
+\emph{Clustering} uses string similarity measures to group similar reads into ``clusters", which is a way of identifying which species of microbe each read corresponds to.
+The alternative, ``aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced.
+From the clustering output, we extract feature vectors, which are then fed into a Support Vector Machine (SVM) classifier.
 
@@
 This process is explained in further detail in the methods section.
-We refer to the pipeline as "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
+We refer to the pipeline as ``CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
 We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data.

### 2016-08-16 latex/input/materials.tex
paper draft v1.1: most minor edits completed

@@
 Qin et al.
-develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "1071", respectively \cite{qin041012}.
+develop a simple T2D classifier using a minimum redundancy---maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages ``sideChannelAttack" and ``e1071", respectively \cite{qin041012}.
 They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}.
@@
 SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
-their respective websites and compiled on the Argo platform.
+their respective websites and compiled on the ARGO platform.

### 2016-08-16 latex/input/methods.tex
finished related work section of paper and added diagrams; began working on results section
paper draft v1.1: most minor edits completed

@@
 
-As mentioned in the Introduction, our pipeline involves a number of steps, which serve a variety of purposes.
+Our proposed pipeline involves a number of steps, which serve a variety of purposes.
 For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled.
-The reads that could not be assembled were unlikely to be useful anyway, since they could not be assembled with other reads into useful contigs.
 The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier.
 We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center.
-Finally, we used a standard SVM classifier to predict patient phenotype, and used several metrics to assess its accuracy.
-We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the results section.
+Finally, we used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
+We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the Results section.
 Aside from the patient labels, this process is entirely de novo, and does not consult any external databases.
@@
 This tool was also used for assembly in the MGWAS study that our data came from.
-We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers\cite{peng12}.
+We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers \cite{peng12}.
 We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51.
+The average insert size was set to 350, in accordance with the reported average insert size from the MGWAS study that we used data from \cite{qin041012}.
 The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together.
@@
 We used UCLUST \cite{Edgar10}, which has been shown to be among the simplest, fastest, and most effective clustering methods \cite{bonder090112, sun042711}.
-Since our contigs were not ordered, we used the usersort option, and we set the sequence match id to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster.
+Since our contigs were not ordered, we used the usersort option, and we set the sequence match threshold to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster.
 For instance, between two strings of length 100, at least 40 places in each of those strings would have to contain the same nucleotide (represented as A, T, G, or C).
 New contigs that did not match at least 40\% to any of the existing contigs would form the seed of a new cluster.
-Again, this value of 40\% was found to be the best value based on our experiments with this dataset.
+This value of 40\% was found to be the best value based on our experiments with this dataset.
+Other values tried, including 50\%, 75\%, and 90\%, led to many clusters with very few reads per cluster.
 
@@
 This is illustrated in Figure \ref{pipeline}.
+We implemented this feature extraction method in Python.
 

### 2016-08-16 latex/input/related.tex
paper draft v1.1: most minor edits completed

@@
 The Multiple Instance Learning (MIL) problem was first described by Dietterich in the context of drug activity prediction \cite{dietterich97}.
-In this problem, we want to figure out which of a number of different molecules will bind to a target "binding site".
+In this problem, we want to figure out which of a number of different molecules will bind to a target ``binding site".
 Each molecule can assume a number of different 3-dimensional shapes, so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding.
-If even one shape of a given molecule binds to the binding site, it is considered a "good" molecule \cite{dietterich97}.
+If even one shape of a given molecule binds to the binding site, it is considered a ``good" molecule \cite{dietterich97}.
 Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains only negative instances.
-This is commonly referred to as the "standard multiple instance assumption" \cite{amores13}.
+This is commonly referred to as the ``standard multiple instance assumption" \cite{amores13}.
 In the original paper, Dietterich develops a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}.
@@
 A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.
-Most of these classic methods follow the standard assumption, which is harmful in domains in which negative bags may contain some positive instances.
+Most of these classic methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some positive instances.
 %
@@
 More recently, there has been increased interest in different formulations of the MIL problem.
-For instance, the problem of "key instance detection" \cite{zhou12} revolves around finding the instances that contribute the most to bag labels.
-A recent study focused on a formulation of the MIL problem in which bags with negative labels can contain some "positive" instances, and developed a general cost function for determining individual instance labels from group labels \cite{kotzias15}.
-This is significant in metagenomics because, while some diseases are caused by a single pathogen, many arise from a combination of many factors, and even patients that are "healthy" may contain small amounts of pathogens that are normally associated with disease.
-In contrast with the standard assumption, this can be referred to as the "collective" assumption \cite{amores13}.
+For instance, the problem of ``key instance detection" \cite{zhou12} revolves around finding the instances that contribute the most to bag labels.
+A recent study focused on a formulation of the MIL problem in which bags with negative labels can contain some positive instances, and developed a general cost function for determining individual instance labels from group labels \cite{kotzias15}.
+This is significant in metagenomics because, while some diseases are caused by a single pathogen, many arise from a combination of many factors, and even patients that are healthy may contain small amounts of pathogens that are normally associated with disease.
+In contrast with the standard assumption, this can be referred to as the ``collective" assumption \cite{amores13}.
 Additionally, it is helpful to discover which microbes and which functional attributes of those microbes lead to disease, making instance level information significant in this domain.
@@
 For methods following the collective assumption, the aggregation function is often based on an averaging of instance labels.
-Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as "Instance Space" by Amores, and he showed that this paradigm was generally significantly less effective than the two other paradigms, "Bag Space" and "Embedded Space" \cite{amores13}.
+Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as ``Instance Space" by Amores, and he showed that this paradigm was generally significantly less effective than the two other paradigms, ``Bag Space" and ``Embedded Space" \cite{amores13}.
 Bag Space methods define a distance or kernel function that determine the similarity between bags, while Embedded Space methods map bags into feature vectors which can then be used for classifiers \cite{amores13}.
-Embedded Space methods can be further divided into two subcategories: methods that simply aggregate information about all instances in a bag without differentiating them, and "Vocabulary-based" methods that group certain similar instances together and then use those groups to form the feature vector \cite{amores13}.
+Embedded Space methods can be further divided into two subcategories: methods that simply aggregate information about all instances in a bag without differentiating them, and ``Vocabulary-based" methods that group certain similar instances together and then use those groups to form the feature vector \cite{amores13}.
 We use a vocabulary-based method in this paper, because having information about groups of similar sequence reads can be biologically important, as explained further in the clustering section.
@@
 The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}.
-SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating "tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}.
+SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating ``tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}.
+This step helps to reduce the overall size of the data by eliminating some reads that would not have been useful anyway.
 The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.
@@
 UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a}, and 
-MC-LSH \cite{bibm2012} are some of the popular  nucleotide sequence clustering approaches.
+MC-LSH \cite{bibm2012} are some of the popular sequence clustering approaches.
 We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}.

### 2016-08-16 pipelineT2D.py

@@ -30,6 +30,7 @@ def parseargs():    # handle user arguments
     parser.add_argument('--svm', default='', help='Path to svm_light home folder, OR specify "misvm" ')
     parser.add_argument('--train', default='NONE', help='Path to svm training data set')
     parser.add_argument('--test', default='NONE', help='Path to svm test data set')
+    parser.add_argument('--testlist', default='NONE', help='Path to file with list of patients to be put in test set.')
     parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
     parser.add_argument('--write', default='NONE', help='Write arrays to outfile (k-mers only) and exit')
     args = parser.parse_args()
@@ -119,22 +120,33 @@ def parse_cluster(verbose, directory, positive, negative, cluster, infile, mapfi
     return vectors
 
 
-def write_train_test(vectors, train, test):     # write training and test files for SVM
+def write_train_test(vectors, train, test, testset):     # write training and test files for SVM
     tr = open(train, 'w')
     te = open(test, 'w')
     counter = 0     # every Nth patient will be put into the test set instead of the train set
-    for key, value in vectors.iteritems():
-        counter = (counter + 1) % 2
-        if counter != 0:
-            if value[0].startswith('1'):
-                for j in range(1, len(value)):      # feature engineering loop
-                    if value[j] != '':
-                        parts = value[j].split(":")
-                        value[j] = parts[0] + ":" + str(float(parts[1]) * 0.9) + ' '    # apply multiplier to features
+    if testset == 'NONE':
+        for key, value in vectors.iteritems():
+            counter = (counter + 1) % 2
+            if counter != 0:
+                '''if value[0].startswith('1'):
+                    for j in range(1, len(value)):      # feature engineering loop
+                        if value[j] != '':
+                            parts = value[j].split(":")
+                            value[j] = parts[0] + ":" + str(float(parts[1]) * 0.9) + ' '   # apply feature multiplier'''
 
-            tr.write((''.join(value)) + '\n')   # write to training set
-        else:
-            te.write((''.join(value)) + '\n')   # write to test set
+                tr.write((''.join(value)) + '\n')   # write to training set
+            else:
+                te.write((''.join(value)) + '\n')   # write to test set
+    else:
+        testitems = []
+        setfile = open(testset, 'r')
+        for line in setfile:
+            testitems.append(line.strip())
+        for key, value in vectors.iteritems():
[…]

### 2016-08-22 latex/input/conclusion.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

@@
 We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability.
-More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Distance-based Bag of Words methods.
+More generally, we have shown the effectiveness of Multiple Instance Learning methods within metagenomics and phenotype prediction, particularly Bag of Words methods.
 CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement.
-Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than D-BoW), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.
+Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than Bag of Words), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

### 2016-08-22 latex/input/experiments.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

@@
 
-The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section.
-The data was split evenly into training and test sets, with each containing 109 patients.
-The training set had 54 T2D patients and 55 control patients, while the test set had 53 T2D patients and 56 control patients.
+The 367 patient files were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the Supplementary Tables of the MGWAS paper \cite{qin041012}, as described in the Materials section.
+Out of the 367 patients, 182 were diabetic and 185 were healthy controls.
 Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
@@
 The assembly of the patient files is embarrassingly parallel, so the assembly of each file could be done at the same time.
-Combining the files into one file took 12 minutes.
+Combining the files into one file took 2 minutes and 35 seconds.
 Assembly was used for all of the classification methods tested, because the combining of individual reads and reduction in total data size made classification feasible.
@@
 We wrote a script to represent the reads as vectors representing the k-mer counts in the string, and these vectors were used by the other methods.
+We tested possible k values from 1 to 6; since the number of possible k-mers is exponential in k, higher values for k quickly become impractical in both run time and memory usage.
 From experimental validation, we found a k-mer value of 3 to be the most effective.
-Converting the reads from their original string form to k-mer vectors took 3 hours, 17 minutes, and 55 seconds.
+Converting the reads from their original string form to k-mer vectors with k=3 took 24 hours, 8 minutes, and 49 seconds.
 
@@
 From their results, the authors generated an Area Under Curve - Receiver Operating Characteristic (AUC-ROC) graph.
-We manually computed the accuracy and F1 score with an ideally-chosen decision boundary based on the results the authors provided in their supplementary tables.
+The authors did not provide a learned decision boundary in their supplementary tables, only the predicted values for each patient, so we manually computed the accuracy and F1 score with an optimally-chosen decision boundary.
 
@@
 
-\begin{table}[h]
+\begin{table*}[t]
 \begin{center} 
-\caption{Comparison of the performance of various methods on the MGWAS dataset. 
-\label{tab:comp}}
+\begin{minipage}{0.4\textwidth}
+\caption{Performance on training set.}
+\label{tab:train-comp}
 \begin{tabular}{|c|ccc|}\hline
 Method & Accuracy & F1-Score & AUC-ROC\\\hline
-MISVM & 50.00 & 50.00 & 50.00\\\hline
-sbMIL & 50.00 & 50.00 & 50.00\\\hline
-GICF & 60.00 & 71.43 & 66.00\\\hline
-mRMR + SVM & 82.61 & 80.02 & 81.00\\\hline 
-CAMIL H-BoW & 91.74 & 91.43 & 95.00\\\hline
-CAMIL D-BoW & \bf{92.66} & \bf{92.31} & \bf{96.00}\\\hline
+mRMR + SVM & 73.26 & 76.65 & 80.77\\\hline 
+CAMIL D-BoW & 90.70 & 90.91 & 97.45\\\hline
+CAMIL H-BoW & \bf{94.19} & \bf{93.87} & \bf{97.87}\\\hline
 \end{tabular}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\caption{Performance with 23 patient test set.} 
+\label{tab:test-comp}
+\begin{tabular}{|c|ccc|}\hline
+Method & Accuracy & F1-Score & AUC-ROC\\\hline
+%MISVM & 50.00 & 50.00 & 50.00\\\hline
+%sbMIL & 50.00 & 50.00 & 50.00\\\hline
+GICF & 73.91 & 75.00 & 78.03\\\hline
+mRMR + SVM & 82.61 & 82.76 & 87.12\\\hline 
+CAMIL D-BoW & 91.30 & 92.31 & \bf{100.00}\\\hline
+CAMIL H-BoW & \bf{100.00} & \bf{100.00} & \bf{100.00}\\\hline
+\end{tabular}
+\bigskip
+\end{minipage}
+%\hfill
+\begin{minipage}{0.4\textwidth}
+\caption{Performance with even train/test split.} 
+\label{tab:even-comp}
+\begin{tabular}{|c|ccc|}\hline
+Method & Accuracy & F1-Score & AUC-ROC\\\hline
+%MISVM & 50.00 & 50.00 & 50.00\\\hline
+%sbMIL & 50.00 & 50.00 & 50.00\\\hline
+GICF & 63.04 & 68.33 & 66.19\\\hline %59.24,63.05,66.19
+CAMIL D-BoW & 86.34 & 87.18 & 95.93\\\hline
+CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63}\\\hline
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\caption{Classification time with even train/test split.} 
+\label{tab:time-comp}
+\begin{tabular}{|c|c|}\hline
+Method & Classification Time\\\hline
+%MISVM & 1 hour\\\hline
+%sbMIL & 1 hour\\\hline
+GICF & 8 hours, 44 minutes, 27 seconds\\\hline
+CAMIL D-BoW & \bf{6 minutes, 58 seconds}\\\hline
+CAMIL H-BoW & 7 minutes, 56 seconds\\\hline
+\end{tabular}
+\end{minipage}
 \end{center}
-\end{table}
+\end{table*}
 
-Table \ref{tab:comp} shows the comparison of classification performance between the various methods described above.
-CAMIL methods significantly outperform the other methods, with the D-BoW variant of CAMIL slightly outperforming the H-BoW variant.
+%\begin{table}[t]
+%\begin{center}
 
-MISVM and sbMIL had the worst results, faring no better than a random guess.
+%\end{center}
+%\end{table}
+
+%\begin{table}[t]
+%\begin{center}
+
+%\end{center}
+%\end{table}
+
+In the original MGWAS paper, the authors use 344 patients as a training set and 23 as a test set.
+However, in their reported AUC-ROC reflects the performance of of the mRMR + SVM approach on the 344 patients whose data was used to generate the gene markers that were used for classification.
+Essentially, they tested the performance of their approach on the training data.
+We thus applied our CAMIL pipeline in the same manner, using the same 344 patients as both the training and test set.
+The results are shown in Table \ref{tab:train-comp}, which demonstrates that both CAMIL variants significantly outperformed the mRMR + SVM approach in all metrics by about 14-17\%, with the CAMIL H-BoW version performing the best.
+Since this is not a typical way of assessing classification performance, we ran several other experiments, discussed below.
+
+The MGWAS authors also provided the calculated labels for the 23 patients in the test set in their supplementary tables \cite{qin041012}, although they did not explicitly report the AUC-ROC on the test set.
+However, based on the reported patient labels, we were able to calculate all of the relevant metrics.
+Table \ref{tab:test-comp} shows the performance of the various algorithms, including GICF, on the test set.
+GICF was outperformed by mRMR + SVM, which was in turn outperformed by CAMIL D-BoW, which was in turn outperformed by CAMIL H-BoW, which got perfect results.
+
+Since the train/test split in Table \ref{tab:test-comp} is so imbalanced, we wanted to evaluate CAMIL with more balanced training and test sets.
+Since this was no longer using the same training set as the original MGWAS paper, we could not evaluate mRMR + SVM.
+Table \ref{tab:even-comp} shows the performance of CAMIL and GICF with an even split between the training and test sets.
+CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
+
+Finally, Table \ref{tab:time-comp} shows the classification time of GICF and CAMIL.
+mRMR + SVM was not included because the runtime of that method was not reported in the MGWAS paper.
+CAMIL took significantly less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch.
+
+MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances.
+Since this dataset involved millions of instances, these methods crashed with memory errors, even when we used only 10\% of the reads.
+We were able to run these methods when using only 1\% of the reads from each patient, and they achieved only 50\% accuracy, no better than a random guess.
 This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
@@
 The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain.
-GICF performs better than MISVM and sbMIL, which makes sense given the fact that it follows the collective assumption.
+GICF performs better than MISVM and sbMIL (albeit with more data), which makes sense given the fact that it follows the collective assumption.
 It also has the benefit of calculating instance labels, which we explore further in the next section.
@@
 \includegraphics[scale=0.4]{./instance-scatter.png}
-\caption{This diagram illustrates the distribution of the instance weights assigned by CAMIL D-BoW.
+\caption{This diagram illustrates the distribution of the instance weights assigned by CAMIL H-BoW.
 The Y-Axis shows the weights, while the X-Axis shows the ranking of the clusters by weight.
-Clearly, there are a relatively small number of clusters that have disproportionately large weights, while there is a smaller number of clusters with smaller weights.} \label{instance-scatter}
+Clearly, there are a relatively small number of clusters that have disproportionately large weights or small weights, while the vast majority of the 3918 clusters have weights close to 0.} \label{instance-scatter}
 \end{figure}
@@
 For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick.
-This is illustrated in Figure \ref{instance-labels}.
-This example is simple, but it explains why static instance labels are insufficient.
+This simple example is illustrated in Figure \ref{instance-labels}.
 
@@
 In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier.
-Conversely, CAMIL uses all of the data to train and test the classifier, and subsequently identifies significant clusters based on the classification results.
-We printed the 25 highest-weighted and 25 lowest-weighted clusters, out of 3188 total clusters.
-The highest-weighted cluster had a weight of 0.012109, while the lowest-weighted cluster had a weight of -0.003668.
-The average of the 25 highest cluster weights was 0.010402, while it was -0.002753 for the lowest 25.
-The average of all 3188 cluster weights was 0.001003, and the median was 0.000254.
-These results indicate that the highest weights were much larger than the lowest weights and skewed the average weight to be much higher than the median weight.
+Conversely, CAMIL uses all of the data to train and test the classifier, resulting in 3918 clusters, and subsequently identifies significant clusters based on the classification results.
+Figure \ref{instance-scatter} is a visual display of the cluster weights determined by CAMIL, using the 344 patient training set and 23 patient test set.
+Clearly, there are a few clusters with disproportionately high or low weights, while most clusters have weights near 0.
+Concretely, the lowest cluster weight is -0.000333, the highest is 0.000374, the mean is 0.000008, and the median is 0.000006.
 Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters are not particularly relevant in this case and whose weights are just noise.
@@
 In contrast, the labels obtained by GICF were barely differentiated from each other at all.
-The instance weights determined by CAMIL are shown in Figure \ref{instance-scatter}.

### 2016-08-22 latex/input/intro.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

@@
 In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances).
-Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}.
-We also show that our pipeline can be used for Histogram-based Bag of Words (H-BoW) methods.
+Specifically, we use Bag of Words (BoW) methods, discussed further in the Background section, which have been shown to be among the most effective and efficient MIL methods \cite{amores13}.
 MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome.

### 2016-08-22 latex/input/materials.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

@@
 This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies.
-The full dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online.
-Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
+The full dataset used in this study contains 367 patients \cite{qin041012}.
+Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP008047, http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
 and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
-The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}.
-The total size of these 218 FASTQ files was 2.3 terabytes, with an average size of 10.82 gigabytes per patient file.
+The labels were found in the paper's Supplementary Tables \cite{qin041012}.
+The total size of these 367 FASTQ files was 3.29 terabytes, with an average size of 8.97 gigabytes per patient file.
 Qin et al.
@@
 They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}.
-While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available.
 The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

### 2016-08-22 latex/input/methods.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

@@
 \includegraphics[scale=0.5]{./mil-metagenomics-pipeline.png}
-\caption{This diagram illustrates the entire pipeline.
+\caption{This diagram illustrates the entire CAMIL pipeline.
 Patient files with FASTQ metagenomic reads are individually assembled using SOAPdenovo2, then combined into one file and clustered with UCLUST.
-We extract features according to the D-BoW method, and classify the patients using the feature vectors with svm-light.} \label{pipeline}
+We extract features according to either the D-BoW or H-BoW method, and classify the patients using the feature vectors with svm-light.} \label{pipeline}
 \end{figure*}
@@
 Aside from the patient labels, this process is entirely de novo, and does not consult any external databases.
-An illustration of this pipeline can be seen in Figure \ref{pipeline} on page \pageref{pipeline}.
+An illustration of the pipeline is shown in Figure \ref{pipeline} on page \pageref{pipeline}.
 
@@
 
-Our feature extraction operation followed the D-BoW method.
+We test both the Distance-based and Histogram-based feature extraction methods within the CAMIL pipeline.
 For each patient, we initialized a vector with length equal to the number of clusters.
-The value for each place in that patient's feature vector would then be set to the closest percentage match (0.4 to 1) of the reads from that patient to the relevant cluster seed.
-For instance, if there are three clusters of reads, and patient A has the cluster seed for cluster 1, no reads from cluster 2, and a read with 70\% match to cluster 3, their feature vector would be~~A = [1.0~~~0.0~~~0.7].
-This is illustrated in Figure \ref{pipeline}.
+The value for each scalar in that patient's feature vector would then represent either the read that matched most closely to that cluster centroid (Distance-based), or the number of reads from that cluster that were present (Histogram-based.
+An example of this is illustrated as a part of Figure \ref{pipeline}.
 We implemented this feature extraction method in Python.

### 2016-08-22 latex/input/related.tex
updated results in paper, edited plots, wrote script to calculate AUC-ROC

@@
 An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
-Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient \cite{amores13}.
+Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}.
 The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center.
-Histogram-based Bag of Words (H-BoW) methods count the number of instances in a given cluster instead of keeping track of the closest-matching instance to that cluster.
-H-BoW methods were found to be among the more effective methods, but not quite as good as D-BoW \cite{amores13}.
+Histogram-based Bag of Words (H-BoW) methods count the number of instances from each cluster there are in each bag, instead of keeping track of the closest-matching instance to that cluster.
+This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important.
+H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on real problems that it is well-suited for \cite{amores13}.
+The D-BoW and H-BoW feature extraction methods are illustrated as part of Figure \ref{pipeline}.
 

### 2016-08-22 aucroc.py

@@ -1,0 +1,108 @@
+# Author: Nathan LaPierre
+# Date: August 18, 2016
+
+
+import argparse
+import sys
+import time
+
+
+def accuracy(tp, fp, tn, fn):
+    if tp == 0 and fp == 0 and tn == 0 and fn == 0:
+        return 0.0
+    else:
+        return float(tp + tn) / float(tp + fp + tn + fn)
+
+
+def f1score(precis, recall):
+    if precis == 0.0 and recall == 0.0:
+        return 0.0
+    else:
+        return 2.0 * precis * recall / (precis + recall)
+
+
+def true_positive_rate(tp, fn):
+    if tp == 0 and fn == 0:
+        return 0.0
+    else:
+        return float(tp) / float(tp + fn)
+
+
+def false_positive_rate(fp, tn):
+    if fp == 0 and tn == 0:
+        return 0.0
+    else:
+        return float(fp) / float(fp + tn)
+
+
+def precision(tp, fp):
+    if tp == 0 and fp == 0:
+        return 0.0
+    else:
+        return float(tp) / float(tp + fp)
+
+
+def parseargs():    # handle user arguments
+    parser = argparse.ArgumentParser(description='Generates area under curve based on predicted vs actual labels')
+    parser.add_argument('--labels', default='NONE', help='Location of file with predicted labels')
+    parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
+    args = parser.parse_args()
+    return args
+
+
+def main():
+    start = time.time()
+    args = parseargs()
+    if args.labels == 'NONE':
+        print 'File with predicted labels must be specified with --labels. Aborting...'
+        sys.exit()
+
+    f = open(args.labels, 'r')
+    total, positive, negative = 0, 0, 0     # number of total, positive, and negative labels
+    labels = []     # will hold the numerical labels
+    for line in f:
+        line.strip()
+        if not line.startswith('#'):
+            l = int(line)
+            labels.append(l)
+            total += 1
+            if l == 1:
+                positive += 1
+            elif l == -1 or l == 0:
+                negative += 1
+    f.close()
+    tp, fp = 0, 0   # number of true and false positives; they start at 0 since we start with highest possible threshold
+    tn = negative   # since everything is negatively predicted at first, true negatives = actual negatives
+    fn = positive   # since everything is negatively predicted at first, false negatives = actual positives
[…]

### 2016-08-22 gicf.py

@@ -28,7 +28,7 @@ def logistic(w, x):     # implementation of the classifier for the instances; cu
     if dot > 709:    # prevent math overflows
         return 1
     if dot < -709:   # prevent math overflows
-        return -1
+        return 0
     return 1 / (1 + math.exp(-dot))
 
 
@@ -234,7 +234,7 @@ def validate(verbose, w, data, k):     # apply the weights determined by gradien
     return accuracy, precision, tpr_recall, fpr, f1_score, predicted_actual
 
 
-def grid_search(verbose, w, data, train, test, res, limit, settings):
+def grid_search(verbose, w, data, train, test, res, limit, settings, testset):
     # linear grid search to set learning rate, lambda, and mini batch size parameters
     if verbose:
         print "Entering grid search to select best parameters for gradient descent."
@@ -244,12 +244,25 @@ def grid_search(verbose, w, data, train, test, res, limit, settings):
     tr = open(train, 'w')
     te = open(test, 'w')
     num = 0     # count of the number of groups
+    testitems = []
+    if testset != 'NONE':
+        setfile = open(testset, 'r')
+        for line in setfile:
+            testitems.append(int(line.strip()))
+        setfile.close()
+
     for line in f:
         num += 1
-        if num % test_split == 0:
-            te.write(line + '\n')
+        if testset != 'NONE':
+            if num in testitems:
+                te.write(line + '\n')
+            else:
+                tr.write(line + '\n')
         else:
-            tr.write(line + '\n')
+            if num % test_split == 0:
+                te.write(line + '\n')
+            else:
+                tr.write(line + '\n')
         if num == limit:
             break
     f.close()
@@ -259,9 +272,9 @@ def grid_search(verbose, w, data, train, test, res, limit, settings):
     if settings == 'NONE':
         rate = [0.0001]  # , 0.001]  # , 0.01, 0.1]
         l = [float(num)]  # [float(num) * 0.1, float(num), float(num) * 10]
-        top_k = [0.8]
-        batch_size = [0.1]  # [num * 4]  # [num * 4, num * 10, num * 100]
-        iterations = [30]  # , 10, 100, 1000]
+        top_k = [0.9]
+        batch_size = [0.01]  # [num * 4]  # [num * 4, num * 10, num * 100]
+        iterations = [3]  # , 10, 100, 1000]
     else:
         rate = settings[0]
         l = settings[1]
@@ -308,6 +321,13 @@ def grid_search(verbose, w, data, train, test, res, limit, settings):
[…]

### 2016-08-22 pipelineT2D.py

@@ -29,8 +29,10 @@ def parseargs():    # handle user arguments
     parser.add_argument('--split', default='.0', help='Characters to split patient tag on.')
     parser.add_argument('--svm', default='', help='Path to svm_light home folder, OR specify "misvm" ')
     parser.add_argument('--train', default='NONE', help='Path to svm training data set')
+    parser.add_argument('--trainastest', default='False', choices=['True', 'False'], help='Use train data for test set')
     parser.add_argument('--test', default='NONE', help='Path to svm test data set')
-    parser.add_argument('--testlist', default='NONE', help='Path to file with list of patients to be put in test set.')
+    parser.add_argument('--testset', default='NONE', help='Path to file with list of patients to be put in test set.')
+    parser.add_argument('--vectors', default='NONE', help='Path to file with feature vectors.')
     parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
     parser.add_argument('--write', default='NONE', help='Write arrays to outfile (k-mers only) and exit')
     args = parser.parse_args()
@@ -120,10 +122,11 @@ def parse_cluster(verbose, directory, positive, negative, cluster, infile, mapfi
     return vectors
 
 
-def write_train_test(vectors, train, test, testset):     # write training and test files for SVM
+def write_train_test(vectors, train, test, testset, trainastest):     # write training and test files for SVM
     tr = open(train, 'w')
     te = open(test, 'w')
     counter = 0     # every Nth patient will be put into the test set instead of the train set
+    testlabels = []     # labels of the test examples
     if testset == 'NONE':
         for key, value in vectors.iteritems():
             counter = (counter + 1) % 2
@@ -135,20 +138,42 @@ def write_train_test(vectors, train, test, testset):     # write training and te
                             value[j] = parts[0] + ":" + str(float(parts[1]) * 0.9) + ' '   # apply feature multiplier'''
 
                 tr.write((''.join(value)) + '\n')   # write to training set
-            else:
+                if trainastest == 'True':
+                    te.write((''.join(value)) + '\n')   # write to test set
+                    if value[0] == '1 ':
+                        testlabels.append(1)
+                    elif value[0] == '-1 ':
+                        testlabels.append(-1)
+            elif trainastest == 'False':
                 te.write((''.join(value)) + '\n')   # write to test set
[…]

### 2016-08-22 seedPrinter.py

@@ -1,0 +1,47 @@
+# Author: Nathan LaPierre
+# Date: August 21, 2016
+
+import argparse
+import sys
+import time
+
+
+def parseargs():    # handle user arguments
+    parser = argparse.ArgumentParser(description='Generates area under curve based on predicted vs actual labels')
+    parser.add_argument('--fasta', default='NONE', help='Location of fasta file.')
+    parser.add_argument('--output', default='seeds.out', help='Where to write output.')
+    parser.add_argument('--uc', default='NONE', help='Location of uclust file.')
+    parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
+    args = parser.parse_args()
+    return args
+
+
+def main():
+    start = time.time()
+    args = parseargs()
+    if args.fasta == 'NONE' or args.uc == 'NONE':
+        print 'UCLUST and FASTA files must be specified with --uc and --fasta, respectively. Aborting...'
+        sys.exit()
+
+    uc = open(args.uc, 'r')
+    fasta = open(args.fasta, 'r')
+    out = open(args.output, 'w')
+    for line in uc:
+        if not line.startswith('#'):
+            if line.startswith('S'):
+                out.write(line)
+                out.write(fasta.next())
+                out.write(fasta.next())
+            else:
+                fasta.next()
+                fasta.next()
+    uc.close()
+    fasta.close()
+    out.close()
+
+    if args.verbose:
+        print "\nProgram execution time: " + str(time.time() - start) + " seconds"
+
+
+if __name__ == "__main__":
+    main()

### 2016-08-22 latex/input/intro.tex
mods lit intro abs

@@
 
-The rest of the paper is organized as follows.
+% Space
+%The rest of the paper is organized as follows.
 Section 2 presents relevant Background information on  Multiple Instance Learning, Assembly, and Clustering.

### 2016-08-22 latex/input/related.tex
mods lit intro abs

@@
 The Multiple Instance Learning (MIL) problem was first described by Dietterich in the context of drug activity prediction \cite{dietterich97}. 
-In this problem, we want to figure out which of a number of different molecules will bind to a target ``binding site".
-Each molecule can assume a number of different 3-dimensional shapes, so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding.
+%In this problem, we want to figure out which of a number of different molecules will bind to a target ``binding site". 
+Each molecule can assume a number of different 3-dimensional shapes (conformations), so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding.
 If even one shape of a given molecule binds to the binding site, it is considered a ``good" molecule \cite{dietterich97}.
@@
 For methods following the collective assumption, the aggregation function is often based on an averaging of instance labels.
-Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as ``Instance Space" by Amores, and he showed that this paradigm was generally significantly less effective than the two other paradigms, ``Bag Space" and ``Embedded Space" \cite{amores13}.
+Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as ``Instance Space" by Amores, and he showed that this paradigm was generally  less effective than the two other paradigms, ``Bag Space" and ``Embedded Space" \cite{amores13}.
 Bag Space methods define a distance or kernel function that determine the similarity between bags, while Embedded Space methods map bags into feature vectors which can then be used for classifiers \cite{amores13}.
@@
 This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important.
-H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on real problems that it is well-suited for \cite{amores13}.
+H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on real world problems \cite{amores13}.
 The D-BoW and H-BoW feature extraction methods are illustrated as part of Figure \ref{pipeline}.
 
-As we have discussed, the Multiple Instance paradigm fits phenotype prediction well, since we have a set of labeled patients containing unlabeled sequence reads, and we would like to predict both the patient phenotype and which reads are indicative of that phenotype.
+The Multiple Instance paradigm fits phenotype prediction well, since we have a set of labeled patients containing unlabeled sequence reads, and we would like to predict both the patient phenotype and which reads are indicative of that phenotype.
 Despite the recent developments in MIL and its potential utility in phenotype prediction, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.
@@
 One of the main purposes of assembly is to determine the whole genomes of microbial species, the vast majority of which have not or cannot be laboratory cultured, from sequencing reads \cite{zerbino08}.
-Even if full genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classifiers, because the contigs will contain more phylogenetic and functional information than short reads.
+Even if complete genomes cannot be assembled, combining reads into larger contigs can still make them much more useful for clustering and classification, because the contigs will contain more phylogenetic and functional information than short reads.
 This is because short reads of less than 1000 base pairs constitute only a tiny fraction of microbial genomes, which are usually hundreds of thousands to millions of base pairs, making it difficult to ascertain much about the phylogeny of individual reads.
@@
 
-Some of the popular genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
+Some of the popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
 We used SOAPdenovo2 in our study, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}.
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.5]{./uclust-similarity.png}
-\caption{This diagram illustrates the similarity computation between two sequences in UCLUST.
+%\begin{figure}[h]
+%\centering
+%\includegraphics[scale=0.5]{./uclust-similarity.png}
+%\caption{This diagram illustrates the similarity computation between two sequences in UCLUST.
 Terminal characters are excluded, so the length of the reads is 10.
@@
 Thus, the similarity is 5/10 = 50\%.} \label{uclust-similarity}
-\end{figure}
+%\end{figure}
 
-The \emph{clustering} problem in this context  involves grouping input short reads  such that reads within a group are similar to each other.
-The clusters obtained from this process are referred to as Operational Taxonomic Units (OTUs).
-OTUs represent a group of equivalent or similar organisms.
-Accordingly, the number of OTUs in a sample gives an approximation of the species diversity in that sample  \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
-In addition to approximating species diversity, clustering has several other key advantages.
-Because clustering is always de novo (unsupervised), it is not limited by the species that are covered in taxonomic databases.
-This detail is important, because it is believed that most microorganisms that reside in the human body have not been laboratory cultured \cite{handelsman04}.
-Clustering also reduces computational costs by allowing analyses to operate on entire clusters instead of on each read.
-Finally, clustering helps the classification process by allowing feature vectors to be built at the OTU level, instead of using individual short reads that contain little biological information.
+The \emph{clustering} problem in this context  involves grouping input short sequences (reads or contigs)   such that sequences 
+within a group are similar to each other. The clusters 
+obtained from this process are referred to 
+as Operational Taxonomic Units (OTUs). OTUs represent a 
+group of equivalent or similar organisms. Accordingly, the 
+number of OTUs in a sample gives an approximation of the 
+species diversity in that sample  \cite{schloss2009introducing,schloss2005introducing,sun2009esprit}.
+In addition to approximating species diversity, clustering has several other key advantages. Because 
+clustering is always de novo (unsupervised), it 
+is not limited by the species that are 
+covered in taxonomic databases. This is  important 
+because it is believed that most micro-organisms that reside 
+in the human body have not been laboratory cultured \cite{handelsman04}. Clustering also reduces computational 
+costs by allowing analyses to operate on entire clusters instead of 
+on each read/contig. Finally, clustering helps 
+the classification process by allowing feature vectors to be built
+at the OTU level, instead of using individual short reads.
 
@@
 The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}.
-This is illustrated in figure \ref{uclust-similarity}.
+
+% Not needed for a bioinf. audience
+%This is illustrated in figure \ref{uclust-similarity}.

### 2016-08-22 latex/input/experiments.tex
comments methods

@@
 Our pipeline uses clustering, whereas the other methods do not.
-Instead, those methods directly compare individual instances instead of clusters.
+Instead, those methods directly compare individual reads instead of clusters.
 In order to do this, sequence reads were represented as counts of k-mers.
-k-mers are nucleotide strings of length k.
-Since there are four possible nucleotides, the number of possible k-mers is \(4^k\).
-For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT.
-For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else.
-We wrote a script to represent the reads as vectors representing the k-mer counts in the string, and these vectors were used by the other methods.
-We tested possible k values from 1 to 6; since the number of possible k-mers is exponential in k, higher values for k quickly become impractical in both run time and memory usage.
+k-mers are nucleotide strings of
+length k. Since there are four possible nucleotides, the number
+of possible k-mers are \(4^k\). 
+%For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. 
+%For a string "ATACGATA", the count for the k-mers is 2 for ATA, 1 for TAC, ACG, CGA, and GAT, and 0 for everything else. 
+We wrote a script to represent the reads/contigs as vectors representing the k-mer counts in the string, and these feature 
+vectors were used by the other methods. We tested possible k values 
+from 1 to 6; since the number of possible k-mers is 
+exponential in k, higher values for k quickly become impractical in both run time and memory usage.
 From experimental validation, we found a k-mer value of 3 to be the most effective.

### 2016-08-23 latex/input/experiments.tex
moved utility scripts and notes into utility folder

@@
 
-The 367 patient files were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the Supplementary Tables of the MGWAS paper \cite{qin041012}, as described in the Materials section.
-Out of the 367 patients, 182 were diabetic and 185 were healthy controls.
 Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
@@
 \begin{center} 
+%\hfill
 \begin{minipage}{0.4\textwidth}
-\caption{Performance on training set.}
+\caption{Performance with even train/test split.} 
+\label{tab:even-comp}
+\begin{tabular}{|c|ccc|}\hline
+Method & Accuracy & F1-Score & AUC-ROC\\\hline
+MISVM & --- & --- & ---\\\hline
+sbMIL & --- & --- & ---\\\hline
+GICF & 63.04 & 68.33 & 66.19\\\hline %59.24,63.05,66.19
+CAMIL D-BoW & 86.34 & 87.18 & 95.93\\\hline
+CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63}\\\hline
+\end{tabular}
+\end{minipage}
+\bigskip
+\begin{minipage}{0.4\textwidth}
+\caption{Classification time with even train/test split.} 
+\label{tab:time-comp}
+\begin{tabular}{|c|c|}\hline
+Method & Classification Time\\\hline
+MISVM & ---\\\hline
+sbMIL & ---\\\hline
+GICF & 8 hours, 44 minutes, 27 seconds\\\hline
+CAMIL D-BoW & \bf{6 minutes, 58 seconds}\\\hline
+CAMIL H-BoW & 7 minutes, 56 seconds\\\hline
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}
+\caption{Performance on subset of instances.}
 \label{tab:train-comp}
@@
 Method & Accuracy & F1-Score & AUC-ROC\\\hline
-mRMR + SVM & 73.26 & 76.65 & 80.77\\\hline 
+MISVM & 50.8 & ? & ?\\\hline
+sbMIL & 50.8 & ? & ?\\\hline
+GICF & ? & ? & ?\\\hline
 CAMIL D-BoW & 90.70 & 90.91 & 97.45\\\hline
@@
 Method & Accuracy & F1-Score & AUC-ROC\\\hline
-%MISVM & 50.00 & 50.00 & 50.00\\\hline
-%sbMIL & 50.00 & 50.00 & 50.00\\\hline
+MISVM & --- & --- & ---\\\hline
+sbMIL & --- & --- & ---\\\hline
 GICF & 73.91 & 75.00 & 78.03\\\hline
@@
 CAMIL H-BoW & \bf{100.00} & \bf{100.00} & \bf{100.00}\\\hline
-\end{tabular}
-\bigskip
-\end{minipage}
-%\hfill
-\begin{minipage}{0.4\textwidth}
-\caption{Performance with even train/test split.} 
-\label{tab:even-comp}
-\begin{tabular}{|c|ccc|}\hline
-Method & Accuracy & F1-Score & AUC-ROC\\\hline
-%MISVM & 50.00 & 50.00 & 50.00\\\hline
-%sbMIL & 50.00 & 50.00 & 50.00\\\hline
-GICF & 63.04 & 68.33 & 66.19\\\hline %59.24,63.05,66.19
-CAMIL D-BoW & 86.34 & 87.18 & 95.93\\\hline
-CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63}\\\hline
-\end{tabular}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
-\caption{Classification time with even train/test split.} 
-\label{tab:time-comp}
-\begin{tabular}{|c|c|}\hline
-Method & Classification Time\\\hline
-%MISVM & 1 hour\\\hline
-%sbMIL & 1 hour\\\hline
-GICF & 8 hours, 44 minutes, 27 seconds\\\hline
-CAMIL D-BoW & \bf{6 minutes, 58 seconds}\\\hline
-CAMIL H-BoW & 7 minutes, 56 seconds\\\hline
 \end{tabular}
@@
 MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances.
-Since this dataset involved millions of instances, these methods crashed with memory errors, even when we used only 10\% of the reads.
-We were able to run these methods when using only 1\% of the reads from each patient, and they achieved only 50\% accuracy, no better than a random guess.
+Since this dataset involved millions of instances, these methods crashed with memory errors, even when we used only 1\% of the reads.
+This is why they are shown as ``---" in the results tables.
+We were able to run these methods when using only 0.1\% of the reads from each patient, and they achieved only 50.8\% accuracy, barely better than a random guess.
 This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
@@
 
-\subsection{Deriving Instance ``Labels"}
-
-\begin{figure}[t]
-\centering
-\includegraphics[scale=0.5]{./instance-labels.png}
-\caption{This diagram illustrates why static instance labels are not sufficient for phenotype prediction.
-A patient with 6 of the blue microbe or 6 of the green microbe may be healthy, while a patient with 3 of each is sick.
-Static instance labels cannot capture this relationship.
-This is also explained by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
-\end{figure}
+\subsection{Cluster-level ``Labels"}
 
@@
 
-One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance ``labels".
-In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other.
-For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick.
-This simple example is illustrated in Figure \ref{instance-labels}.
-
-However, we can infer from the SVM decision boundary which clusters appear to be most relevant to the disease diagnosis.
-Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis.
-For instance, if the \emph{i}th scalar in the weight vector is has the highest value of any of the weights, then cluster \emph{i} is likely to play a major role in the disease pathology.
-Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease.
-Because the data is metagenomic, the clusters represent both phylogenetic and functional similarity, so identifying the most relevant clusters can help discover more about the pathology of the disease.
-For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.
-
 In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier.

### 2016-08-23 latex/input/materials.tex
moved utility scripts and notes into utility folder

@@
 This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies.
+Additionally, the authors called for more extensive testing of gut microbiota classifiers \cite{qin041012}.
 The full dataset used in this study contains 367 patients \cite{qin041012}.
@@
 The total size of these 367 FASTQ files was 3.29 terabytes, with an average size of 8.97 gigabytes per patient file.
-Qin et al.
+Out of the 367 patients, 182 were diabetic and 185 were healthy controls.
+% Qin et al.
 develop a simple T2D classifier using a minimum redundancy---maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages ``sideChannelAttack" and ``e1071", respectively \cite{qin041012}.
-They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}.
+They achieved an AUC-ROC of 0.81 by training a classifier on 344 of the patients and using the remaining 23 as a test set \cite{qin041012}.
 The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

### 2016-08-23 latex/input/methods.tex
comments methods
moved utility scripts and notes into utility folder

@@
 The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier.
-We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center.
+We  applied our feature extraction method, which uses either the Histogram-based or Distance-based BoW methods.
 Finally, we used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
@@
 In effect, it measures how far off incorrect predictions were from being correct.
-AUC-ROC plots True Positive Rate versus False Positive Rate, given by:
+AUC-ROC plots True Positive Rate (TPR) versus False Positive Rate (FPR), given by:
 \begin{equation}
-True Positive Rate = Recall = TP / (TP + FN). \label{eqn:roc}
+TPR  = TP / (TP + FN). \label{eqn:roc}
 \end{equation}
 \begin{equation}
-False Positive Rate = FP / (FP + TN). \label{eqn:prec}
+FPR = FP / (FP + TN). \label{eqn:prec}
 \end{equation}
+
+\subsection{Deriving Instance ``Labels"}
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.5]{./instance-labels.png}
+\caption{This diagram illustrates why static instance labels are not sufficient for phenotype prediction.
+A patient with 6 of the blue microbe or 6 of the green microbe may be healthy, while a patient with 3 of each is sick.
+Static instance labels cannot capture this relationship.
+This is also explained by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
+\end{figure}
+
+One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance ``labels".
+In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other.
+For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick.
+This simple example is illustrated in Figure \ref{instance-labels}.
+We can infer from the SVM decision boundary which clusters appear to be most relevant to the disease diagnosis.
+Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis.
+For instance, if the \emph{i}th scalar in the weight vector is has the highest value of any of the weights, then cluster \emph{i} is likely to play a major role in the disease pathology.
+Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease.
+Because the data is metagenomic, the clusters represent both phylogenetic and functional similarity, so identifying the most relevant clusters can help discover more about the pathology of the disease.
+For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.

### 2016-08-23 latex/input/related.tex
moved utility scripts and notes into utility folder

@@
 Each molecule can assume a number of different 3-dimensional shapes (conformations), so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding.
-If even one shape of a given molecule binds to the binding site, it is considered a ``good" molecule \cite{dietterich97}.
+If even one shape of a given molecule binds to a binding site, it is considered a ``good" molecule \cite{dietterich97}.
 Thus, in the original formation of MIL, a bag is classified as positive if one or more instances within it is positive, while a negative bag contains only negative instances.
@@
 This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important.
-H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on real world problems \cite{amores13}.
+H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}.
 The D-BoW and H-BoW feature extraction methods are illustrated as part of Figure \ref{pipeline}.
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[scale=0.5]{./de-bruijn.png}
-\caption{This diagram illustrates a basic de Bruijn graph, as explained in the text.
+%\begin{figure}[h]
+%\centering
+%\includegraphics[scale=0.5]{./de-bruijn.png}
+%\caption{This diagram illustrates a basic de Bruijn graph, as explained in the text.
 Notice the double arrow from TAT to ATT, allowing a path to go through that route twice.
 Thus, the longest contig that can be created by following a path is GTATTATTGGA, which is what we want.} \label{de-bruijn}
-\end{figure}
+%\end{figure}
 
@@
 The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.
-There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.
+%There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.
 

### 2016-08-23 labelDisplay.py

@@ -1,10 +1,0 @@
-# Author: Nathan LaPierre
-# Date: July 27, 2016
-
-
-def main():
-    pass
-
-
-if __name__ == "__main__":
-    main()

### 2016-08-23 mousetests/clusterRep.py
removed old directory with tests on mouse metagenome files

@@ -1,200 +1,0 @@
-# Author: Nathan LaPierre
-# Date: March 17, 2016
-
-import argparse
-import os
-import shlex
-import subprocess
-import sys
-import time
-
-
-def option1(indir, output, alpha, uclustid, uclustpath, verbose):
-    print indir + output + alpha + uclustid + uclustpath + verbose
-    print "Option not yet implemented."
-
-
-def option2(indir, output, alpha, uclustid, uclustpath, verbose):
-    print indir + output + alpha + uclustid + uclustpath + verbose
-    print "Option not yet implemented."
-
-
-def option3(indir, output, alpha, uclustid, uclustpath, verbose):
-    tempfilename = 'ccclusterreptempfile'
-    tempfile = open(tempfilename, 'w')      # temporary helper file, deleted at the end
-    try:
-        clustlist = []      # list of patients and cluster numbers, i.e. patient A cluster 1 becomes "A/1"
-        for filename in os.listdir(indir):
-            if filename.endswith('.uc'):
-                if verbose:
-                    print 'Attempting to open ' + filename + ' and ' + filename.split('.')[0] + '.fasta'
-                fasta = open(filename.split('.')[0] + '.fasta')  # open corresponding fasta file for uclust file
-                f = open(filename)
-                for line in f:
-                    if line.startswith('H') or line.startswith('S'):
-                        fastaline = fasta.next()
-                        if line.startswith('S'):        # cluster seed: write to temp file
-                            fields = line.split('\t')
-                            clustlist.append(fields[8].split('.')[0] + '/' + fields[1])  # for patient A clust 1, "A/1"
-                            tempfile.write(fastaline)   # write the defline
-                            fastaline = fasta.next()    # go to read line
-                            tempfile.write(fastaline)   # write the read
-                        else:
-                            fasta.next()    # read nothing
-                f.close()
-                fasta.close()
-        tempfile.close()
-        if verbose:
-            print 'Calling: ' + uclustpath + ' --input ' + os.curdir + '/' + tempfilename \
-                  + ' --uc ' + output + ' --usersort --id ' + str(uclustid)
-        subprocess.call(shlex.split(uclustpath + ' --input ' + os.curdir + '/' + tempfilename
-                                    + ' --uc ' + output + ' --usersort --id ' + str(uclustid)))  # call uclust
-        outfile = open(output)
-        index = 0   # used to index the cluster list created earlier
[…]

### 2016-08-24 latex/input/conclusion.tex
updated results and tables; added utility scripts for generating AUC-ROC from MGWAS supplementary tables

@@
 We have demonstrated an effective and efficient computational pipeline for classifying patient phenotype based on metagenomic data.
-We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study.
+We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study and other common MIL methods.
 We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability.

### 2016-08-24 latex/input/experiments.tex
updated results and tables; added utility scripts for generating AUC-ROC from MGWAS supplementary tables

@@
 However, because this method compares only individual instances and not entire bags, and treats bag labels simply as aggregations of instance labels, GICF is still an Instance Space method.
+We implemented GICF in Python.
 
@@
 
-\begin{table*}[t]
+\begin{table}[h]
 \begin{center} 
 %\hfill
-\begin{minipage}{0.4\textwidth}
 \caption{Performance with even train/test split.} 
@@
 \end{tabular}
-\end{minipage}
-\bigskip
-\begin{minipage}{0.4\textwidth}
-\caption{Classification time with even train/test split.} 
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\begin{center}
+\caption{Classification time and memory usage with even train/test split.} 
 \label{tab:time-comp}
-\begin{tabular}{|c|c|}\hline
-Method & Classification Time\\\hline
-MISVM & ---\\\hline
-sbMIL & ---\\\hline
-GICF & 8 hours, 44 minutes, 27 seconds\\\hline
-CAMIL D-BoW & \bf{6 minutes, 58 seconds}\\\hline
-CAMIL H-BoW & 7 minutes, 56 seconds\\\hline
+\begin{tabular}{|c|cc|}\hline
+Method & Classification Time & Memory Usage\\\hline
+MISVM & --- & Memory Error\\\hline
+sbMIL & --- & Memory Error\\\hline
+GICF & 8 hours, 44 mins, 27 secs & 2.646 GB\\\hline
+CAMIL D-BoW & \bf{6 minutes, 58 seconds} & \bf{545.293 MB}\\\hline
+CAMIL H-BoW & 7 minutes, 56 seconds & 546.297 MB\\\hline
 \end{tabular}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
-\caption{Performance on subset of instances.}
-\label{tab:train-comp}
+\end{center}
+\end{table}
+
+In the original MGWAS paper, the authors use 344 patients as a training set and 23 as a test set.
+However, when comparing to other MIL algorithms, we wished to have a more balanced training vs.
+test set split.
+We put 184 patients in the training set and 183 in the test set.
+Since this used a different training set than the one in the MGWAS paper, we do not compare our results to theirs here.
+Table \ref{tab:even-comp} shows the comparison of results between CAMIL, GICF, MISVM, and sbMIL, while Table \ref{tab:time-comp} compares the classification time and memory usage.
+CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
+CAMIL took significantly less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch.
+MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances.
+Since this dataset involved millions of instances, these methods crashed with memory errors.
+This is why they are shown as ``---" in the results tables.
+
+\begin{table}[h]
+\begin{center}
+\caption{Performance on subset of instances with even train/test split.}
+\label{tab:subset-comp}
 \begin{tabular}{|c|ccc|}\hline
@@
 sbMIL & 50.8 & ? & ?\\\hline
-GICF & ? & ? & ?\\\hline
-CAMIL D-BoW & 90.70 & 90.91 & 97.45\\\hline
-CAMIL H-BoW & \bf{94.19} & \bf{93.87} & \bf{97.87}\\\hline
+GICF & 64.67 & 66.95 & 65.56\\\hline
+CAMIL D-BoW & \bf{84.15} & \bf{85.13} & \bf{90.86}\\\hline
+CAMIL H-BoW & 74.32 & 68.46 & 83.49\\\hline
 \end{tabular}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}
+\end{center}
+\end{table}
+
+We wanted to compare MISVM and sbMIL to GICF and CAMIL, so we used a subset of the instances for each patient so that MISVM and sbMIL would not crash.
+They could only be run on 0.1\% of the reads for each patient, and even then took over 32.5 GB of memory.
+The results are shown in Table \ref{tab:subset-comp}.
+MISVM and sbMIL also only achieved 50.8\% accuracy, only slightly better than a random guess.
+GICF and CAMIL, while experiencing a performance drop due to the massive information loss, still performed much better, with the CAMIL methods outperforming GICF again.
+CAMIL D-BoW outperformed CAMIL H-BoW this time, because the distance calculations are less affected by having fewer reads than the histogram method of counting the number of reads from each cluster for each patient.
+
+\begin{table}[h]
+\begin{center}
 \caption{Performance with 23 patient test set.} 
@@
 Method & Accuracy & F1-Score & AUC-ROC\\\hline
-MISVM & --- & --- & ---\\\hline
-sbMIL & --- & --- & ---\\\hline
-GICF & 73.91 & 75.00 & 78.03\\\hline
-mRMR + SVM & 82.61 & 82.76 & 87.12\\\hline 
-CAMIL D-BoW & 91.30 & 92.31 & \bf{100.00}\\\hline
-CAMIL H-BoW & \bf{100.00} & \bf{100.00} & \bf{100.00}\\\hline
+mRMR + SVM & 80.00 & 81.20 & 82.30\\\hline 
+CAMIL D-BoW & 91.62 & 91.89 & \bf{98.03}\\\hline
+CAMIL H-BoW & \bf{95.59} & \bf{95.20} & 97.93\\\hline
 \end{tabular}
-\end{minipage}
 \end{center}
-\end{table*}
+\end{table}
 
-%\begin{table}[t]
-%\begin{center}
+Finally, Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM.
+For this comparison, we randomly selected 23 patients for the test set 10 times, and averaged the results of each method for all 10 test sets.
+23 patients were used because the test set used by Qin et al.
+contained 23 patients \cite{qin041012}.
+Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
+The results held for the original 23 patient test set, where CAMIL H-BoW actually achieved 100\% accuracy.
 
-%\end{center}
-%\end{table}
-
-%\begin{table}[t]
-%\begin{center}
-
-%\end{center}
-%\end{table}
-
-In the original MGWAS paper, the authors use 344 patients as a training set and 23 as a test set.
-However, in their reported AUC-ROC reflects the performance of of the mRMR + SVM approach on the 344 patients whose data was used to generate the gene markers that were used for classification.
-Essentially, they tested the performance of their approach on the training data.
-We thus applied our CAMIL pipeline in the same manner, using the same 344 patients as both the training and test set.
-The results are shown in Table \ref{tab:train-comp}, which demonstrates that both CAMIL variants significantly outperformed the mRMR + SVM approach in all metrics by about 14-17\%, with the CAMIL H-BoW version performing the best.
-Since this is not a typical way of assessing classification performance, we ran several other experiments, discussed below.
-
-The MGWAS authors also provided the calculated labels for the 23 patients in the test set in their supplementary tables \cite{qin041012}, although they did not explicitly report the AUC-ROC on the test set.
-However, based on the reported patient labels, we were able to calculate all of the relevant metrics.
-Table \ref{tab:test-comp} shows the performance of the various algorithms, including GICF, on the test set.
-GICF was outperformed by mRMR + SVM, which was in turn outperformed by CAMIL D-BoW, which was in turn outperformed by CAMIL H-BoW, which got perfect results.
-
-Since the train/test split in Table \ref{tab:test-comp} is so imbalanced, we wanted to evaluate CAMIL with more balanced training and test sets.
-Since this was no longer using the same training set as the original MGWAS paper, we could not evaluate mRMR + SVM.
-Table \ref{tab:even-comp} shows the performance of CAMIL and GICF with an even split between the training and test sets.
-CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
-
-Finally, Table \ref{tab:time-comp} shows the classification time of GICF and CAMIL.
-mRMR + SVM was not included because the runtime of that method was not reported in the MGWAS paper.
-CAMIL took significantly less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch.
-
-MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances.
-Since this dataset involved millions of instances, these methods crashed with memory errors, even when we used only 1\% of the reads.
-This is why they are shown as ``---" in the results tables.
-We were able to run these methods when using only 0.1\% of the reads from each patient, and they achieved only 50.8\% accuracy, barely better than a random guess.
+MISVM and sbMIL performed the worst overall.
 This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
@@
 It also has the benefit of calculating instance labels, which we explore further in the next section.
-However, GICF is still an instance space method, so it makes sense that it performs worse than CAMIL.
+However, GICF is still an instance space method, so it makes sense that CAMIL outperformed it.
 

### 2016-08-24 gicf.py

@@ -138,11 +138,16 @@ def descent(verbose, rate, l, top_k, batch_size, iterations, w, data):
             else:
                 labels.append(1)
             group = listify(splits[1])
-            cutoff = int(top_k * len(group))     # use only top (100 * (1 - top_k))% of reads
-            for i in range(cutoff, len(group)):
-                if random.random() > chance:
-                    cost_data[len(labels)-1].append(group[i])   # append appropriate instance to mini batch dataset
+            if len(group) < 1000:   # if group is small enough, just do full batch gradient descent
+                for i in range(len(group)):
+                    cost_data[len(labels)-1].append(group[i])
                     instances += 1.0
+            else:
+                cutoff = int(top_k * len(group))     # use only top (100 * (1 - top_k))% of reads
+                for i in range(cutoff, len(group)):
+                    if random.random() > chance:
+                        cost_data[len(labels)-1].append(group[i])   # append appropriate instance to mini batch dataset
+                        instances += 1.0
         d.close()
 
         # find value of derivative of cost function using weights w, then use learning rate to find adjustment amount

### 2016-08-24 pipelineT2D.py

@@ -131,12 +131,6 @@ def write_train_test(vectors, train, test, testset, trainastest):     # write tr
         for key, value in vectors.iteritems():
             counter = (counter + 1) % 2
             if counter != 0:
-                '''if value[0].startswith('1'):
-                    for j in range(1, len(value)):      # feature engineering loop
-                        if value[j] != '':
-                            parts = value[j].split(":")
-                            value[j] = parts[0] + ":" + str(float(parts[1]) * 0.9) + ' '   # apply feature multiplier'''
-
                 tr.write((''.join(value)) + '\n')   # write to training set
                 if trainastest == 'True':
                     te.write((''.join(value)) + '\n')   # write to test set
@@ -194,23 +188,29 @@ def execute_svm(verbose, directory, train, test, model, prediction, svm, output)
             subprocess.call(shlex.split('./'+svm+'svm_classify '+test+' '+model+' '+prediction), stdout=outfile)
 
 
-def order_labels(testlabels, predictions):
+def order_labels(testlabels, predictions, classifier):
     pred_act = []   # predicted vs. actual label
-    f = open(predictions, 'r')
-    linecount = 0
-    for line in f:
-        line = line.strip()
-        if linecount >= 2:
-            pred_act.append([float(line), testlabels[linecount-2]])
-        linecount += 1
-    f.close()
+    if classifier == 'svmlight':
+        f = open(predictions, 'r')
+        linecount = 0
+        for line in f:
+            line = line.strip()
+            if linecount >= 2:
+                pred_act.append([float(line), testlabels[linecount-2]])
+            linecount += 1
+        f.close()
+    elif classifier == 'misvm':
+        for i in range(len(testlabels)):
+            pred_act.append([float(predictions[i]), float(testlabels[i])])
 
-    # a.sort(key=lambda x: x[1])
     pred_act.sort(key=lambda x: x[0])   # sort by predicted label
     print pred_act
-    outfile = open(predictions + '_ordered', 'w')
+    if classifier == 'svmlight':
+        outfile = open(predictions + '_ordered', 'w')
+    else:
+        outfile = open('predictions_ordered', 'w')
     for line in range(len(pred_act)):  # loop to write actual labels in order of predicted value for that label
-        outfile.write(str(pred_act[len(pred_act) - line - 1][1]) + '\n')
+        outfile.write(str(int(pred_act[len(pred_act) - line - 1][1])) + '\n')
     outfile.close()
 
[…]

### 2016-08-24 utility/mgwasOrdered.py

@@ -1,0 +1,57 @@
+# Author: Nathan LaPierre
+# Date: August 23, 2016
+
+import argparse
+import sys
+
+
+def parseargs():    # handle user arguments
+    parser = argparse.ArgumentParser(description='Prints ordered predictions for mRMR + SVM for random test sets.')
+    parser.add_argument('--labels', default='NONE', help='File with MGWAS labels.')
+    parser.add_argument('--patients', default='NONE', help='File with MGWAS patient tags.')
+    parser.add_argument('--testset', default='NONE', help='File with patient tags for test set.')
+    parser.add_argument('-v', '--verbose', default=False, action='store_true', help='Verbose output')
+    args = parser.parse_args()
+    return args
+
+
+def main():
+    args = parseargs()
+    if args.labels == 'NONE' or args.patients == 'NONE' or args.testset == 'NONE':
+        print '--labels, --patients, and --testset must be specified. See -h for help. Aborting...'
+        sys.exit()
+
+    labels = []     # holds patient labels
+    tags = []       # holds patient tags
+    labelfile = open(args.labels, 'r')
+    patientfile = open(args.patients, 'r')
+    for line in labelfile:
+        line = line.strip()
+        if not line.startswith('#'):
+            patienttag = patientfile.next().strip().split(' = ')[1]
+            labels.append(int(line))
+            tags.append(patienttag)
+    labelfile.close()
+    patientfile.close()
+
+    pred_act = []   # holds predicted vs actual labels
+    testtags = open(args.testset, 'r')
+    for line in testtags:
+        line = line.strip()
+        if args.verbose:
+            print line
+        for i in range(len(tags)):
+            if line == tags[i]:
+                if args.verbose:
+                    print "Matched: " + line + ' --> ' + tags[i]
+                pred_act.append([len(tags) - i, labels[i]])
+    testtags.close()
+
+    pred_act.sort(key=lambda x: x[0])   # sort by predicted label
+    for i in range(len(pred_act)):
+        print pred_act[len(pred_act)-i-1][1]      # print actual labels in order of predicted labels
+    if args.verbose:
+        print len(pred_act)
+
+if __name__ == "__main__":
+    main()

### 2016-08-24 utility/randomTestSet.py

@@ -1,0 +1,45 @@
+# Author: Nathan LaPierre
+# Date: August 23, 2016
+
+import argparse
+import random
+import sys
+
+
+def parseargs():    # handle user arguments
+    parser = argparse.ArgumentParser(description='Script that picks a random test set from a set.')
+    parser.add_argument('--map', default='NONE', help='File number to start on. Required.')
+    parser.add_argument('--size', default=23, help='Size of test set to pick.')
+    args = parser.parse_args()
+    return args
+
+
+def main():
+    args = parseargs()
+    if args.map == 'NONE':
+        print 'Map file must be specified with --map. Aborting...'
+        sys.exit()
+    try:
+        size = args.size
+    except ValueError:
+        print 'Size must be a positive integer. Aborting...'
+        sys.exit()
+    if size < 1:
+        print 'Size must be a positive integer. Aborting...'
+        sys.exit()
+
+    tags = []        # holds patient tags from map file
+    testnums = []   # holds the line numbers in the map to choose from
+    mapfile = open(args.map, 'r')
+    for line in mapfile:
+        tags.append(line.split('\t')[3])
+    mapfile.close()
+    while len(testnums) < size:
+        num = random.random() * len(tags)   # pick random number in range of number of patients
+        if num not in testnums:
+            testnums.append(tags[int(num)])    # keep adding unique numbers until we have desired number of test items
+    for t in testnums:
+        print t
+
+if __name__ == "__main__":
+    main()

### 2016-08-24 latex/input/intro.tex
added MIT software license

@@
 
-We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype.
+We used data from a Metagenome-Wide Association Study (MGWAS) \cite{qin041012}, which compares microbial metagenomic data between many patients with or without a given phenotype.
 MGWAS studies contain many expert-labeled patients and the metagenomic data associated with those patients, so they are useful for phenotype prediction but also cause many computational challenges.
@@
 \emph{Clustering} uses string similarity measures to group similar reads into ``clusters", which is a way of identifying which species of microbe each read corresponds to.
-The alternative, ``aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced.
+The alternative, ``aligning" the reads with known genomes, is impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced.
 From the clustering output, we extract feature vectors, which are then fed into a Support Vector Machine (SVM) classifier.
 
-Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM.
+Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into one file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM.
 This process is explained in further detail in the methods section.
 We refer to the pipeline as ``CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
-We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data.
+We then compare the results of our method against the classifier used in the MGWAS study \cite{qin041012} from which we derived our data.
 We show that our classifier shows significantly improved performance.
-This is discussed further in the Experimental Results section.
+We have also released code for our pipeline on GitHub\footnote{https://github.com/nlapier2/multiInstanceLearning},
+under the open-source MIT license.
 

### 2016-08-24 latex/input/experiments.tex
paper edits: moved parts of related work to methods, moved evaluation metrics to materials, other minor fixes

@@
 
-Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
+Each bag of reads was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
 Assembly for each file took 7-30 minutes, depending on the file size.
@@
 
+MISVM and sbMIL performed the worst overall.
+This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
+Additionally, they are instance space methods that do not leverage bag-level information.
+The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain.
+GICF performs better than MISVM and sbMIL (albeit with more data), which makes sense given the fact that it follows the collective assumption.
+It also has the benefit of calculating instance labels, which we explore further in the next section.
+However, GICF is still an instance space method, so it makes sense that CAMIL outperformed it.
+
 \begin{table}[h]
@@
 The results held for the original 23 patient test set, where CAMIL H-BoW actually achieved 100\% accuracy.
-
-MISVM and sbMIL performed the worst overall.
-This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
-Additionally, they are instance space methods that do not leverage bag-level information.
-The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain.
-GICF performs better than MISVM and sbMIL (albeit with more data), which makes sense given the fact that it follows the collective assumption.
-It also has the benefit of calculating instance labels, which we explore further in the next section.
-However, GICF is still an instance space method, so it makes sense that CAMIL outperformed it.
 

### 2016-08-24 latex/input/materials.tex
updated results and tables; added utility scripts for generating AUC-ROC from MGWAS supplementary tables
paper edits: moved parts of related work to methods, moved evaluation metrics to materials, other minor fixes

@@
 
+\subsection{Evaluation Metrics}
+
+We can assess the success of our classifier in several ways. The simplest measure, accuracy, measures the percentage of instances 
+that are classified correctly, represented by 
+\begin{equation}
+Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
+\end{equation}
+where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.
+
+Accuracy as an evaluation metric can be biased if one of the classes 
+(positive or negative)  has a larger number of examples than the other.   
+Precision measures the percentage of positive predictions that 
+were correct, 
+whereas recall measures the percentage of positive 
+examples that were correctly predicted (or retrieved). 
+We can represent Precision and Recall as:
+
+\begin{equation}
+Precision = TP / (TP + FP). \label{eqn:prec}
+\end{equation}
+\begin{equation}
+Recall = TP / (TP +FN). \label{eqn:roc}
+\end{equation}
+
+The F1 score captures the trade-offs between precision and recall in a
+single metric and is the harmonic mean of precision and recall, given by:
+\begin{equation}
+F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
+\end{equation}
+
+Finally, we also use the Area Under Curve of the Receiver Operating Characteristic (AUC-ROC), which measures the performance of the classifier as the decision boundary threshold is moved.
+The SVM classifier generally predicts a group label to be negative if the predicted label for that group was less than 0 and predicts a group label to be positive otherwise.
+The AUC-ROC measures the performance of the classifier as the threshold is varied to more or less than 0.
+In effect, it measures how far off incorrect predictions were from being correct.
+AUC-ROC plots True Positive Rate (TPR) versus False Positive Rate (FPR), given by:
+\begin{equation}
+TPR  = TP / (TP + FN). \label{eqn:roc}
+\end{equation}
+\begin{equation}
+FPR = FP / (FP + TN). \label{eqn:prec}
+\end{equation}
+
 \subsection{Software and Hardware Details}
@@
 their respective websites and compiled on the ARGO platform.
+The source code for our implementations of the H-BoW and D-BoW feature extraction methods and GICF are available on GitHub\footnote{https://github.com/nlapier2/multiInstanceLearning}.

### 2016-08-24 latex/input/methods.tex
paper edits: moved parts of related work to methods, moved evaluation metrics to materials, other minor fixes

@@
 The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier.
-We  applied our feature extraction method, which uses either the Histogram-based or Distance-based BoW methods.
-Finally, we used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
+We then developed a vocabulary-based feature extraction method.
+
+Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
+Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}.
+The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center.
+Histogram-based Bag of Words (H-BoW) methods count the number of instances from each cluster there are in each bag, instead of keeping track of the closest-matching instance to that cluster.
+This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important.
+H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}.
+
+Our feature extraction method uses either the Histogram-based or Distance-based BoW methods.
+We used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
 We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the Results section.
@@
 
-For our assembly step, we used SOAPdenovo2 \cite{luo12}, which is a de novo metagenome assembly tool.
-This tool was also used for assembly in the MGWAS study that our data came from.
-We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers \cite{peng12}.
+For our assembly step, we used SOAPdenovo2, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}.
+SOAPdenovo2 first constructs a type of directed graph called a \emph{de Bruijn} graph that represents the overlaps between different sequences \cite{li10}.
+Reads are divided into strings of length K called \emph{k-mers}; these k-mers are the nodes of the graph \cite{zerbino08}.
+The choice of K is up to the user, and is important for having good assembly results.
+The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}.
+SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating ``tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}.
+This step helps to reduce the overall size of the data by eliminating some reads that would not have been useful anyway.
+The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.
+
 We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51.
@@
 
-We used UCLUST \cite{Edgar10}, which has been shown to be among the simplest, fastest, and most effective clustering methods \cite{bonder090112, sun042711}.
+We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}.
+UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}.
+Thus, each \emph{centroid} defines the center of a cluster, and the distance T defines the radius of the cluster, such that any point that has a similarity of greater than T to the centroid is within the radius and is thus part of the cluster.
+UCLUST is a heuristic algorithm that has several optimizations to improve speed, thus condition (i) above is not always guaranteed \cite{Edgar10}.
+UCLUST proceeds in a greedy, iterative manner.
+The first sequence in the input file becomes a new cluster centroid.
+For each new sequence in the file, it is compared with each of the existing cluster centroids in order.
+As soon as it is compared with a centroid that it has a similarity of greater than T with, it becomes part of that cluster.
+If the read is not similar enough with any of the existing cluster centroids, it becomes the centroid of a new cluster.
+The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}.
+
 Since our contigs were not ordered, we used the usersort option, and we set the sequence match threshold to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster.
@@
 
-\subsection{Feature Extraction}
+\subsection{Feature Extraction and Classification}
 
@@
 We implemented this feature extraction method in Python.
-
-\subsection{Classification and Evaluation Metrics}
-
 We performed classification with a standard SVM classifier using the generated feature vectors; in this case, we used svm-light \cite{joachims08}.
 The choice of classifier is not very important for D-BoW methods \cite{amores13}.
-
-We can assess the success of our classifier in several ways. The simplest measure, accuracy, measures the percentage of instances 
-that are classified correctly, represented by 
-\begin{equation}
-Accuracy = (TP + TN)/ (TP + TN + FP + FN)  \label{eqn:acc} 
-\end{equation}
-where TP, TN, FP and FN represents true positives, true negatives, false positives and false negatives respectively.
-
-Accuracy as an evaluation metric can be biased if one of the classes 
-(positive or negative)  has a larger number of examples than the other.   
-Precision measures the percentage of positive predictions that 
-were correct, 
-whereas recall measures the percentage of positive 
-examples that were correctly predicted (or retrieved). 
-We can represent Precision and Recall as:
-
-\begin{equation}
-Precision = TP / (TP + FP). \label{eqn:prec}
-\end{equation}
-\begin{equation}
-Recall = TP / (TP +FN). \label{eqn:roc}
-\end{equation}
-
-The F1 score captures the trade-offs between precision and recall in a
-single metric and is the harmonic mean of precision and recall, given by:
-\begin{equation}
-F1-Score = 2 * (Precision * Recall)/ (Precision + Recall). \label{eqn:f1}
-\end{equation}
-
-Finally, we also use the Area Under Curve of the Receiver Operating Characteristic (AUC-ROC), which measures the performance of the classifier as the decision boundary threshold is moved.
-The SVM classifier generally predicts a group label to be negative if the predicted label for that group was less than 0 and predicts a group label to be positive otherwise.
-The AUC-ROC measures the performance of the classifier as the threshold is varied to more or less than 0.
-In effect, it measures how far off incorrect predictions were from being correct.
-AUC-ROC plots True Positive Rate (TPR) versus False Positive Rate (FPR), given by:
-\begin{equation}
-TPR  = TP / (TP + FN). \label{eqn:roc}
-\end{equation}
-\begin{equation}
-FPR = FP / (FP + TN). \label{eqn:prec}
-\end{equation}
 

### 2016-08-24 latex/input/related.tex
paper edits: moved parts of related work to methods, moved evaluation metrics to materials, other minor fixes

@@
 %In this problem, we want to figure out which of a number of different molecules will bind to a target ``binding site". 
-Each molecule can assume a number of different 3-dimensional shapes (conformations), so even for molecules that are known to bind to the binding site, it is not necessarily known which formation of the molecule succeeds in binding.
+Each molecule can assume a number of different 3-dimensional shapes (conformations), so even for molecules that are known to bind to the binding site, it is not necessarily known which conformation of the molecule succeeds in binding.
 If even one shape of a given molecule binds to a binding site, it is considered a ``good" molecule \cite{dietterich97}.
@@
 This is commonly referred to as the ``standard multiple instance assumption" \cite{amores13}.
-In the original paper, Dietterich develops a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}.
+In the original paper, Dietterich developed a solution based on axis-parallel rectangles to solve this problem, and the MIL approach was shown to be significantly more effective than a standard supervised learning approach \cite{dietterich97}.
 In the late 1990s and early 2000s, a number of different approaches were developed for the original MIL problem, such as Diverse Density (DD) \cite{perez98}, EM-DD \cite{zhang01}, MI-SVM \cite{andrews02}, sbMIL \cite{bunescu07}, and MILES \cite{wang06}.
 A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.
-Most of these classic methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some positive instances.
+Most of these classic methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some proportion of positive instances.
 %
@@
 For methods following the collective assumption, the aggregation function is often based on an averaging of instance labels.
-Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as ``Instance Space" by Amores, and he showed that this paradigm was generally  less effective than the two other paradigms, ``Bag Space" and ``Embedded Space" \cite{amores13}.
+Regardless of the problem assumption, methods that rely only on comparisons between individual instances are referred to as ``Instance Space" by Amores; this paradigm was generally  less effective than the two other paradigms, ``Bag Space" and ``Embedded Space" \cite{amores13}.
 Bag Space methods define a distance or kernel function that determine the similarity between bags, while Embedded Space methods map bags into feature vectors which can then be used for classifiers \cite{amores13}.
@@
 We use a vocabulary-based method in this paper, because having information about groups of similar sequence reads can be biologically important, as explained further in the clustering section.
-
-An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
-Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}.
-The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center.
-Histogram-based Bag of Words (H-BoW) methods count the number of instances from each cluster there are in each bag, instead of keeping track of the closest-matching instance to that cluster.
-This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important.
-H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}.
-The D-BoW and H-BoW feature extraction methods are illustrated as part of Figure \ref{pipeline}.
 
@@
 Thus, metagenome assembly is a relatively new and challenging field.
-
 Some of the popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
-We used SOAPdenovo2 in our study, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}.
-SOAPdenovo2 first constructs a type of directed graph called a \emph{de Bruijn} graph that represents the overlaps between different sequences \cite{li10}.
-Reads are divided into strings of length K called \emph{k-mers}; these k-mers are the nodes of the graph \cite{zerbino08}.
-The choice of K is up to the user, and is important for having good assembly results.
-The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}.
-SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating ``tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}.
-This step helps to reduce the overall size of the data by eliminating some reads that would not have been useful anyway.
-The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.
 %There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.
@@
 at the OTU level, instead of using individual short reads.
-
 UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a}, and
 MC-LSH \cite{bibm2012} are some of the popular sequence clustering approaches.
-We use UCLUST within our study, which is one of the most widely used and cited metagenome clustering methods and has been shown to be amongst the most effective in terms of speed and accuracy in benchmarking studies \cite{bonder090112, sun042711}.
-%
-
-UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}.
-Thus, each \emph{centroid} defines the center of a cluster, and the distance T defines the radius of the cluster, such that any point that has a similarity of greater than T to the centroid is within the radius and is thus part of the cluster.
-UCLUST is a heuristic algorithm that has several optimizations to improve speed, thus condition (i) above is not always guaranteed \cite{Edgar10}.
-UCLUST proceeds in a greedy, iterative manner.
-The first sequence in the input file becomes a new cluster centroid.
-For each new sequence in the file, it is compared with each of the existing cluster centroids in order.
-As soon as it is compared with a centroid that it has a similarity of greater than T with, it becomes part of that cluster.
-If the read is not similar enough with any of the existing cluster centroids, it becomes the centroid of a new cluster.
-The similarity measure T is defined as a string similarity between the two nucleotide sequences that counts the number of character placements that they have in common and then divides that number by the length of the reads, with terminal characters excluded \cite{Edgar10}.
 

### 2016-08-24 latex/input/experiments.tex
final paper edits before bibm submission

--- 2016-08-24 latex/input/experiments.tex
@@
 
-Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}.
-Assembly for each file took 7-30 minutes, depending on the file size.
+Each bag of reads was assembled with SOAPdenovo2, which took 7-30 minutes per file, depending on the file size.
 The assembly of the patient files is embarrassingly parallel, so the assembly of each file could be done at the same time.
@@
 k-mers are nucleotide strings of
-length k. Since there are four possible nucleotides, the number
-of possible k-mers are \(4^k\). 
+length k. Since there are four possible nucleotides, the number of possible k-mers is \(4^k\). 
 %For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. 
@@
 From experimental validation, we found a k-mer value of 3 to be the most effective.
-Converting the reads from their original string form to k-mer vectors with k=3 took 24 hours, 8 minutes, and 49 seconds.
+Converting the reads from their string form to k-mer vectors with k=3 took 24 hours, 8 minutes, 49 seconds.
 
@@
 
-We implemented two different versions of the pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction.
+We implemented two different versions of the pipeline in Python: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction.
 These results are denoted as ``CAMIL D-BoW" and ``CAMIL H-BoW", respectively, in the results tables and graphs.
-Clustering for the pipeline with UCLUST took 10 hours and 51 minutes.
+Clustering for the pipeline with UCLUST took 14 hours, 7 minutes, 44 seconds.
 Without the assembly step reducing the size of the data, clustering would not have been feasible.
-Feature extraction and SVM-light classification took 10 minutes and 17 seconds to run for D-BoW, versus 5 minutes and 14 seconds for H-BoW.
 
@@
 However, because this method compares only individual instances and not entire bags, and treats bag labels simply as aggregations of instance labels, GICF is still an Instance Space method.
-We implemented GICF in Python.
+GICF is a generalized cost function, but the authors also use a specific version of it in their paper with squared loss for bag and instance level errors and logistic regression for classification \cite{kotzias15}; this is the specific version that we implemented in Python.
+Like Kotzias et al.
+\cite{kotzias15}, we used mini-batch stochastic gradient descent with momentum to train the classifier and linear grid search to pick the parameters.
 
@@
 
-The methods in the original paper are neither MIL-based, nor are they entirely de novo apart from patient labels.
-The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then use a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}.
+The methods used by Qin et al.
+\cite{qin041012} are neither MIL-based, nor are they entirely de novo apart from patient labels.
+The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then used a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}.
 They then combined these genes with an existing gene catalog, MetaHIT \cite{qin030410}, and carried out taxonomic assignment and functional annotation of the genes using the KEGG \cite{kanehisa00} and eggNOG \cite{powell12} databases, as well as 2,890 other reference genomes \cite{qin041012}.
@@
 Method & Accuracy & F1-Score & AUC-ROC\\\hline
-MISVM & 50.8 & ? & ?\\\hline
-sbMIL & 50.8 & ? & ?\\\hline
+MISVM & 50.8 & --- & 48.47\\\hline
+sbMIL & 50.8 & --- & 48.47\\\hline
 GICF & 64.67 & 66.95 & 65.56\\\hline
@@
 MISVM and sbMIL also only achieved 50.8\% accuracy, only slightly better than a random guess.
+The F1 score could not be computed because there were no True Positives or False Positives.
 GICF and CAMIL, while experiencing a performance drop due to the massive information loss, still performed much better, with the CAMIL methods outperforming GICF again.
-CAMIL D-BoW outperformed CAMIL H-BoW this time, because the distance calculations are less affected by having fewer reads than the histogram method of counting the number of reads from each cluster for each patient.
+CAMIL D-BoW outperformed CAMIL H-BoW this time, because the distance calculations are less affected by having fewer reads than the histogram method.
+
+MISVM and sbMIL performed the worst overall.
+This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
+Additionally, they are instance space methods that do not leverage bag-level information.
+The performance of these two methods serves to illustrate why many of the classic MIL algorithms with standard assumptions will not be effective in this domain.
+GICF performs better than MISVM and sbMIL, which makes sense given the fact that it follows the collective assumption.
+It also has the benefit of calculating instance labels, which we explore further in the next section.
+However, GICF is still an instance space method, so it makes sense that CAMIL outperformed it.
 
@@
 Finally, Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM.
-For this comparison, we randomly selected 23 patients for the test set 10 times, and averaged the results of each method for all 10 test sets.
+For this comparison, we averaged the results of 10 trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set.
 23 patients were used because the test set used by Qin et al.

### 2016-08-24 latex/input/intro.tex
final paper edits before bibm submission

@@
 We refer to the pipeline as ``CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.
-We then compare the results of our method against the classifier used in the MGWAS study \cite{qin041012} from which we derived our data.
+We then compare the results of our method against the classifier used in the MGWAS study \cite{qin041012} from which we derived our data, as well as other popular MIL methods.
 We show that our classifier shows significantly improved performance.

### 2016-08-24 latex/input/materials.tex
final paper edits before bibm submission

@@
 of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}.
-This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies.
+This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is one of the largest among those studies.
 Additionally, the authors called for more extensive testing of gut microbiota classifiers \cite{qin041012}.
@@
 SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
-their respective websites and compiled on the ARGO platform.
-The source code for our implementations of the H-BoW and D-BoW feature extraction methods and GICF are available on GitHub\footnote{https://github.com/nlapier2/multiInstanceLearning}.
+were downloaded from their respective websites and compiled on the ARGO platform.
+The source code for our implementations of the H-BoW and D-BoW feature extraction methods and GICF are available on GitHub\footnote{https://github.com/nlapier2/multiInstanceLearning}
+under the open-source MIT license.

### 2016-08-24 latex/input/methods.tex
final paper edits before bibm submission

@@
 
-Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
+An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
 Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}.
@@
 
-Our feature extraction method uses either the Histogram-based or Distance-based BoW methods.
+Our feature extraction method uses either the Histogram-based or Distance-based BoW method.
 We used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
-We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the Results section.
+We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in subsection \ref{deriving-instances}.
 Aside from the patient labels, this process is entirely de novo, and does not consult any external databases.
@@
 For each patient, we initialized a vector with length equal to the number of clusters.
-The value for each scalar in that patient's feature vector would then represent either the read that matched most closely to that cluster centroid (Distance-based), or the number of reads from that cluster that were present (Histogram-based.
+The value for each scalar in that patient's feature vector would then represent either the read that matched most closely to that cluster centroid (Distance-based), or the number of reads from that cluster that were present (Histogram-based).
 An example of this is illustrated as a part of Figure \ref{pipeline}.
@@
 We performed classification with a standard SVM classifier using the generated feature vectors; in this case, we used svm-light \cite{joachims08}.
-The choice of classifier is not very important for D-BoW methods \cite{amores13}.
+The choice of classifier is not very important for BoW methods \cite{amores13}.
 
 \subsection{Deriving Instance ``Labels"}
+\label{deriving-instances}
 

### 2016-08-24 latex/input/related.tex
final paper edits before bibm submission

@@
 For instance, the problem of ``key instance detection" \cite{zhou12} revolves around finding the instances that contribute the most to bag labels.
-A recent study focused on a formulation of the MIL problem in which bags with negative labels can contain some positive instances, and developed a general cost function for determining individual instance labels from group labels \cite{kotzias15}.
+A recent study by Kotzias et al.
+focused on a formulation of the MIL problem in which bags with negative labels can contain some positive instances, and developed a general cost function for determining individual instance labels from group labels \cite{kotzias15}.
 This is significant in metagenomics because, while some diseases are caused by a single pathogen, many arise from a combination of many factors, and even patients that are healthy may contain small amounts of pathogens that are normally associated with disease.

### 2016-10-24 latex/input/conclusion.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

@@
 We have demonstrated that even relatively simple de novo assembly and clustering methods, when used within this pipeline, lead to significantly better performance results than the standard classifier used in the original Metagenome-Wide Association Study and other common MIL methods.
+We would like to emphasize that, while we used a particular type 2 diabetes dataset in this paper, the methods described here are general-purpose and could be used for any metagenomic dataset.
 We have shown how to infer the most important OTUs in the disease pathology by using the SVM decision boundary and discussed the clinical importance of this ability.
@@
 CAMIL is a relatively simple and easy to implement pipeline that has both shown strong results and significant potential for even further improvement.
-Future work could revolve around improving individual parts of the pipeline, such as better assembly and clustering methods, application of different multiple instance learning methods (other than Bag of Words), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.
+Future work could revolve around improving individual parts of the pipeline, such as using better assembly and clustering methods, application of different multiple instance learning methods (other than Bag of Words), and further attempts to generate more specific instance level or cluster level information and validate that information against the known pathology of various diseases.

### 2016-10-24 latex/input/experiments.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

@@
 In order to do this, sequence reads were represented as counts of k-mers.
-k-mers are nucleotide strings of
-length k. Since there are four possible nucleotides, the number of possible k-mers is \(4^k\). 
+%k-mers are nucleotide strings of length k.
+Since there are four possible nucleotides, the number of possible k-mers is \(4^k\).
 %For instance, if k=3, there are \(4^3\) = 64 possible k-mers, which are AAA, AAC, AAG, AAT, ACA, ACC, ..., TTT. 
@@
 Clustering for the pipeline with UCLUST took 14 hours, 7 minutes, 44 seconds.
-Without the assembly step reducing the size of the data, clustering would not have been feasible.
+All 367 patients were clustered together in order to keep the features consistent.
 
@@
 Finally, Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM.
-For this comparison, we averaged the results of 10 trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set.
-23 patients were used because the test set used by Qin et al.
-contained 23 patients \cite{qin041012}.
+We use 23 patients in the test set here because that is what Qin et al.
+did \cite{qin041012}.
+We averaged the results of 10 trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set.
 Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
-The results held for the original 23 patient test set, where CAMIL H-BoW actually achieved 100\% accuracy.
+The results held for the original 23 patient test set used by Qin et al., where CAMIL H-BoW actually achieved 100\% accuracy.
 

### 2016-10-24 latex/input/methods.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

@@
 The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier.
-We then developed a vocabulary-based feature extraction method.
-
-An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
-Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}.
-The distinguishing feature in D-BoW methods is that the values for the feature vector represent the instance that has the smallest distance to the cluster center.
-Histogram-based Bag of Words (H-BoW) methods count the number of instances from each cluster there are in each bag, instead of keeping track of the closest-matching instance to that cluster.
-This intuitively has appeal in the domain of microbiome analysis, as the relative quantities of different species of bacteria is important.
-H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}.
-
-Our feature extraction method uses either the Histogram-based or Distance-based BoW method.
-We used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
+We then developed a vocabulary-based feature extraction method, discussed further in subsection \ref{feature-extraction}.
+Using the extracted feature vectors, we trained an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy.
 We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in subsection \ref{deriving-instances}.
@@
 
-For our assembly step, we used SOAPdenovo2, because it was the assembler used in the original MGWAS study \cite{qin041012} and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}.
-SOAPdenovo2 first constructs a type of directed graph called a \emph{de Bruijn} graph that represents the overlaps between different sequences \cite{li10}.
-Reads are divided into strings of length K called \emph{k-mers}; these k-mers are the nodes of the graph \cite{zerbino08}.
-The choice of K is up to the user, and is important for having good assembly results.
-The k-mer nodes in the graph have an edge between them if a read contains those k-mers in order with an overlap of K-1 nucleotides, and the direction of the edge indicates in which order the k-mers appear \cite{zerbino08}.
-SOAPdenovo2 then cleans this graph by removing nodes/sequences with few or no connections with other sequences, eliminating ``tips" that represent likely sequencing machine errors, and removing redundant edges \cite{li10}.
-This step helps to reduce the overall size of the data by eliminating some reads that would not have been useful anyway.
-The contigs are then formed by combining reads according to the de Bruijn graph: each contig represents a directed path in the graph \cite{zerbino08}.
-
+For our assembly step, we used SOAPdenovo2, because it was the assembler used in the MGWAS study \cite{qin041012} that we compare our results with and because it has been shown to be one of the fastest assembly algorithms \cite{peng12}.
+It should be noted that SOAPdenovo2 was not originally intended for metagenome assembly, but is often tuned for that application, as was done by us and Qin et al.
+\cite{qin041012}.
 We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51.
@@
 \subsection{Feature Extraction and Classification}
+\label{feature-extraction}
 
-We test both the Distance-based and Histogram-based feature extraction methods within the CAMIL pipeline.
-For each patient, we initialized a vector with length equal to the number of clusters.
-The value for each scalar in that patient's feature vector would then represent either the read that matched most closely to that cluster centroid (Distance-based), or the number of reads from that cluster that were present (Histogram-based).
-An example of this is illustrated as a part of Figure \ref{pipeline}.
-We implemented this feature extraction method in Python.
-We performed classification with a standard SVM classifier using the generated feature vectors; in this case, we used svm-light \cite{joachims08}.
+We used a "vocabulary-based" feature extraction method.
+An example of Vocabulary-based methods are Bag of Words (BoW) methods, which involve the following three-step process: (i) Cluster the instances to create classes of instances; (ii) for each bag, map the clusters of instances in that bag to a feature vector; and (iii) use a standard classifier that uses the feature vectors to predict group labels \cite{amores13}.
+Step (i) is covered by our assembly and clustering process, while step (iii) is covered by performing classification with a standard SVM classifier based on the extracted feature vectors.
+In this case, we used svm-light \cite{joachims08}.
 The choice of classifier is not very important for BoW methods \cite{amores13}.
+Below, we describe our feature selection methods for step (ii), which we implemented in Python, as well as the rationale for using these methods.
+
+Amores found the Distance-based Bag of Words (D-BoW) method to be the second most effective of all tested methods, and the most effective one that was also time-efficient (linear, rather than quadratic, in the number of bags and number of instances per bag) \cite{amores13}.
+H-BoW methods were found by Amores to be somewhat less effective than D-BoW methods on average, but performed the best out of all algorithms on several datasets, indicating that this method performs very well on some real world problems \cite{amores13}.
+Thus, we tested our pipeline using both of these feature extraction methods.
+
+Either way, the input is a set of clusters for each patient.
+The D-BoW method creates a feature vector based on the contig for each cluster that was the closest match to the cluster seed.
+For instance, say Patient A's reads include the centroid of cluster 1, another contig that has a 45\% match to the centroid of cluster 1, no contigs from cluster 2, and two contigs that match to the centroid of cluster 3, one with a 57\% match and one with an 82\% match.
+The string match percentage is determined by UCLUST, as described in the previous subsection.
+Then, D-BoW would extract the feature vector [1, 0, 0.82], indicating the contigs for Patient A that match most closely to the cluster centroid for each cluster.
+The H-BoW method, instead of using the closest match to each cluster, counts the number of contigs for a patient that belong to each cluster.
+For the above example, the H-BoW method would extract the feature vector [2, 0, 2], since Patient A has 2 representatives from clusters 1 and 3, but no representatives from cluster 2.
+This example is illustrated in part of Figure \ref{pipeline}.
 

### 2016-10-24 latex/input/related.tex
initial paper updated further explaining feature extraction method and reducing explanation of assembly

@@
 A recent review of MIL by Amores created a taxonomy of these various methods and compared their effectiveness for classification \cite{amores13}.
-Most of these classic methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some proportion of positive instances.
+Most of these methods follow the standard assumption, which is not useful in real domains in which negative bags may contain some proportion of positive instances.
 %
@@
 This can be done either \emph{de novo} (in an unsupervised manner) or by referencing sequences against known contigs.
-We focus on de novo assembly, because we wish to keep our pipeline as unsupervised as possible.
+We focus on de novo assembly, in order to keep our pipeline as unsupervised as possible.
 
@@
 Thus, metagenome assembly is a relatively new and challenging field.
-Some of the popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
+Some popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
+Here we use SOAPdenovo2 because it was the assembler used by Qin et al.
+in their metagenomics study \cite{qin041012}.
 %There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.
@@
 at the OTU level, instead of using individual short reads.
-UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, MC-MinH \cite{sdm2013a}, and
-MC-LSH \cite{bibm2012} are some of the popular sequence clustering approaches.
+UCLUST \cite{Edgar10}, CD-HIT \cite{Li01072006}, mothur \cite{schloss2009introducing}, DOTUR \cite{schloss2005introducing}, CROP \cite{Hao01032011}, and MC-MinH \cite{sdm2013a} are some of the popular sequence clustering approaches.
 

### 2016-10-27 latex/input/experiments.tex
results section comparison to mRMR+SVM updated

@@
 
-Finally, Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM.
-We use 23 patients in the test set here because that is what Qin et al.
-did \cite{qin041012}.
-We averaged the results of 10 trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set.
-Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
-The results held for the original 23 patient test set used by Qin et al., where CAMIL H-BoW actually achieved 100\% accuracy.
+Table \ref{tab:test-comp} compares CAMIL to the method used by Qin et al., mRMR + SVM.
+We initially tested CAMIL on the same 23 patient test set that Qin et al.
+used, for which CAMIL H-BoW had 100\% accuracy and AUC, while mRMR + SVM had 0.81 AUC as reported by Qin et al.
+\cite{qin041012}.
+We know CAMIL is not 100\% accurate, so we validated it by averaging the results of 10 independent trials in which we selected 23 patients randomly out of the 367 to serve as the test set, with the other 344 of the training set.
+Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM on these trials, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
 
-The method used in the original paper is the only one tested here that is not an MIL method, and the only one that is not entirely unsupervised apart from the patient labels.
-Given that the methods from this paper were not de novo, it makes sense that they would outperform many of the unsupervised MIL methods.
-However, CAMIL still significantly outperformed the results reported in the original paper.
-We believe that this is due to the following reasons: (i) the clustering process puts similar contigs into groups that are useful features for the classifier, and (ii) instead of attempting to select the most significant features before performing classification, we allow the classifier itself to determine the most significant features from the entire pool of features.
+Unlike the other MIL methods, Qin et al.
+use reference genomes to inform their classifier, so it makes sense that their method performs better than the MIL methods that only use de novo techniques.
+However, CAMIL still significantly outperformed the results reported in the MGWAS paper.
+We believe that the primary reason for this is that Qin et al.
+relied on alignments of sequences to reference genomes and attempted to select the 50 most significant genes for the phenotype before building the classifier.
+There are two primary problems with this approach: (i) many microbes found in the gut do not exist in reference databases and would thus be unusable for their classifier, and (ii) by only using 50 genes to inform the classifier, a lot of potentially valuable data is left out.
+CAMIL avoids these issues by using as much data as can be assembled and not relying on reference databases.
+The tradeoff is that we don't know exactly what genes are being used by the classifier to form the decision boundary.
+We also believe that the clustering process of putting similar contigs into groups forms useful features for the classifier.
 

### 2016-10-29 latex/input/experiments.tex
incorporated most of reviewer suggestions in paper

@@
 \label{tab:even-comp}
-\begin{tabular}{|c|ccc|}\hline
-Method & Accuracy & F1-Score & AUC-ROC\\\hline
-MISVM & --- & --- & ---\\\hline
-sbMIL & --- & --- & ---\\\hline
-GICF & 63.04 & 68.33 & 66.19\\\hline %59.24,63.05,66.19
-CAMIL D-BoW & 86.34 & 87.18 & 95.93\\\hline
-CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63}\\\hline
+\begin{tabular}{|c|ccccc|}\hline
+Method & Accuracy & F1-Score & AUC-ROC & CV Acc. & CV F1\\\hline
+MISVM & --- & --- & --- & --- & ---\\\hline
+sbMIL & --- & --- & --- & --- & ---\\\hline
+GICF & 63.04 & 68.33 & 66.19 & --- & ---\\\hline %59.24,63.05,66.19
+GICF-Cluster & 79.31 & 82.35 & 82.38 & --- & ---\\\hline
+CAMIL D-BoW & 86.34 & 87.18 & 95.93 & 82.07 & 83.07\\\hline
+CAMIL H-BoW & \bf{90.71} & \bf{89.70} & \bf{97.63} & \bf{89.13} & \bf{88.09}\\\hline
 \end{tabular}
@@
 GICF & 8 hours, 44 mins, 27 secs & 2.646 GB\\\hline
-CAMIL D-BoW & \bf{6 minutes, 58 seconds} & \bf{545.293 MB}\\\hline
-CAMIL H-BoW & 7 minutes, 56 seconds & 546.297 MB\\\hline
+GICF-Cluster & 24 minutes, 51 secs & \bf{500.07 MB}\\\hline 
+CAMIL D-BoW & 5 minutes, 33 seconds & 545.293 MB\\\hline
+CAMIL H-BoW & \bf{5 minutes, 25 seconds} & 546.297 MB\\\hline
 \end{tabular}
@@
 
-In the original MGWAS paper, the authors use 344 patients as a training set and 23 as a test set.
+Qin et al.
+use 344 patients as a training set and 23 as a test set.
 However, when comparing to other MIL algorithms, we wished to have a more balanced training vs.
@@
 CAMIL methods significantly outperform GICF, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
-CAMIL took significantly less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch.
-MISVM and sbMIL require computing a kernel matrix of size N*N, where N = the number of instances.
-Since this dataset involved millions of instances, these methods crashed with memory errors.
-This is why they are shown as ``---" in the results tables.
+We attempted to improve on GICF's results by selecting only one read from each cluster for each patient and discarding the other reads.
+This significantly reduced the computation time and improved results, showing how important the clustering step is.
+These results are listed as ``GICF-Cluster" in the tables above.
+Even with this step, CAMIL was faster and more effective than GICF-Cluster, demonstrating the effectiveness of our feature extraction method.
+We performed Leave One Out Cross Validation for the Accuracy and F1-Score metrics (denoted in the table as CV Acc.
+and CV F1, respectively).
+CAMIL's cross validation results were slightly worse than its test set results but still significantly outperformed GICF.
+Cross validation results for GICF and GICF-Cluster were infeasible to calculate due to extremely long computaion time.
+CAMIL took much less time than GICF for two main reasons: (i) GICF requires representing each read as an array of length 64 (for k-mer length 3), while CAMIL reduces the data size with clustering and feature extraction; (ii) GICF requires expensive pairwise comparisons between each pair of instances in a mini-batch.
+MISVM and sbMIL require computing a kernel matrix of size N*N, where N = number of instances.
+Since this dataset had millions of instances, these methods crashed with memory errors, and are shown as ``---" in the tables.
+We attempted to resolve these issues by applying the same methods that we used for GICF-Cluster, but still received memory errors.
 
@@
 The results are shown in Table \ref{tab:subset-comp}.
-MISVM and sbMIL also only achieved 50.8\% accuracy, only slightly better than a random guess.
-The F1 score could not be computed because there were no True Positives or False Positives.
+MISVM and sbMIL only achieved 50.8\% accuracy.
+The F1 score could not be computed because there were no True or False Positives.
 GICF and CAMIL, while experiencing a performance drop due to the massive information loss, still performed much better, with the CAMIL methods outperforming GICF again.
@@
 MISVM and sbMIL performed the worst overall.
-This makes sense, as they make the standard MIL assumption, which doesn't make sense in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
+This makes sense, as they make the standard MIL assumption, which is not helpful in the context of phenotype prediction, in which even healthy patients can host a small number of pathogens.
 Additionally, they are instance space methods that do not leverage bag-level information.
@@
 Table \ref{tab:test-comp} shows that CAMIL significantly outperformed mRMR + SVM on these trials, with the H-BoW variant of CAMIL slightly outperforming the D-BoW variant.
+Leave-one-out cross validation generally yielded similar results to the test set results, except precision was sometimes lower.
 

### 2016-10-29 latex/input/methods.tex
results section comparison to mRMR+SVM updated
incorporated most of reviewer suggestions in paper

@@
 UCLUST seeks to ensure that, for some similarity T, the following conditions hold: (i) all cluster centroids have a similarity of less than T to each other; and (ii) all points in a cluster have a similarity of greater than T to the cluster centroid \cite{Edgar10}.
-Thus, each \emph{centroid} defines the center of a cluster, and the distance T defines the radius of the cluster, such that any point that has a similarity of greater than T to the centroid is within the radius and is thus part of the cluster.
-UCLUST is a heuristic algorithm that has several optimizations to improve speed, thus condition (i) above is not always guaranteed \cite{Edgar10}.
 UCLUST proceeds in a greedy, iterative manner.
@@
 In this case, we used svm-light \cite{joachims08}.
-The choice of classifier is not very important for BoW methods \cite{amores13}.
 Below, we describe our feature selection methods for step (ii), which we implemented in Python, as well as the rationale for using these methods.
@@
 Static instance labels cannot capture this relationship.
-This is also explained by Amores in his MIL taxonomy \cite{amores13}.} \label{instance-labels}
+This is also explained by Amores \cite{amores13}.} \label{instance-labels}
 \end{figure}

### 2016-10-29 latex/input/related.tex
results section comparison to mRMR+SVM updated
incorporated most of reviewer suggestions in paper

@@
 Despite the recent developments in MIL and its potential utility in phenotype prediction, we have not found any literature that specifically applies MIL to classifying patient phenotype based on metagenomic data.
+We present our MIL-based feature extraction method in section \ref{feature-extraction}.
 
@@
 Many modern sequence reads are produced by Next-Generation and High-Throughput Sequencers, which usually produce these short reads.
-Metagenomics additionally poses its own set of challenges, due to very large datasets and lack of knowledge about how many species are present and in what relative abundances \cite{namiki12}.
-Thus, metagenome assembly is a relatively new and challenging field.
-Some popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, ABySS \cite{simpson09}, and Ray Meta \cite{boisvert12}.
-Here we use SOAPdenovo2 because it was the assembler used by Qin et al.
-in their metagenomics study \cite{qin041012}.
+Metagenomics poses its own set of challenges, due to large datasets and lack of knowledge about how many species are present and in what abundances \cite{namiki12}.
+Thus, metagenome assembly is a new and challenging field.
+Some popular single genome and metagenome assembly approaches include SOAPdenovo2 \cite{luo12}, IDBA-UD \cite{peng12}, Velvet \cite{zerbino08}, MetaVelvet \cite{namiki12}, and Ray Meta \cite{boisvert12}.
 %There are variations on how exactly de Bruijn graphs are represented and implemented, but Figure \ref{de-bruijn} illustrates the general process.

### 2016-10-29 bak/gicf.py
directory cleaning

@@ -1,379 +1,0 @@
-# Author: Nathan LaPierre
-# Date: June 6, 2016
-
-
-import argparse
-import math
-import random
-import sys
-import time
-
-
-def similarity(i, j):   # similarity measure (0 to 1) between instances i and j
-    norm = 0.0    # the squared 2-norm of the difference between the two instance vectors, computed in for loop below
-    for k in range(len(i)):
-        norm += (i[k] - j[k]) ** 2
-    return math.exp(-norm)      # returns value for kernel described in GICF paper
-
-
-def instance_penalty(i, j):     # non-negative penalty on the difference between predictions for instances i and j
-    return (i - j) ** 2     # using squared loss as penalty, as described in GICF paper
-
-
-def logistic(w, x):     # implementation of the classifier for the instances; currently logistic regression
-    dot = 0.0     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i]*x[i]
-    if dot > 20:    # prevent math overflows
-        return 1
-    if dot < -20:   # prevent math overflows
-        return 0
-    return ((1 / (1 + math.exp(-dot))) - 0.5) * 2  # return normalized logistic regression formula using dot
-
-
-def derivative_logistic(w, x):   # derivative of logistic regression function
-    dot = 0.0     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i]*x[i]
-    if dot > 20 or dot < -20:    # prevent math overflows
-        return 0
-    e = math.exp(dot)   # e ^ (dot product of w and x)
-    norm = 0.0        # norm of vector x times scalar e, calculated below
-    for k in range(len(x)):
-        norm += (x[k] * e) ** 2
-    norm = math.sqrt(norm)
-    value = math.fabs(norm / ((1 + e) ** 2))    # value for derivative of logistic function
-    return ((1 / (value + 1)) - 0.5) * 2    # normalized result
-
-
-def label_penalty(predicted, actual):   # non-negative penalty on difference between prediction & true label for group
-    return (predicted - actual) ** 2     # using squared loss as penalty, as described in GICF paper
-
-
-def predict_label(group, instance_labels):
-    # real scalar representing predicted label for group based on aggregation of instance labels
-    total = 0.0     # combined value of all labels in group, computed in for loop below
-    for i in range(len(group)):
-        total += instance_labels[i]
-    return total / len(group)   # return average label value in the group
-
-
[…]

### 2016-10-29 gicf.py

@@ -1,412 +1,0 @@
-# Author: Nathan LaPierre
-# Date: June 6, 2016
-
-
-import argparse
-import math
-import numpy as np
-import random
-import sys
-import time
-
-
-def similarity(i, j):   # similarity measure (0 to 1) between instances i and j
-    norm = 0.0    # the squared 2-norm of the difference between the two instance vectors, computed in for loop below
-    for k in range(len(i)):
-        norm += (i[k] - j[k]) ** 2
-    return math.exp(-norm)  # / len(i) / len(j)      # returns value for kernel described in GICF paper
-
-
-def instance_penalty(i, j):     # non-negative penalty on the difference between predictions for instances i and j
-    return (i - j) ** 2     # using squared loss as penalty, as described in GICF paper
-
-
-def logistic(w, x):     # implementation of the classifier for the instances; currently logistic regression
-    dot = w[0]     # dot product between weights and x, computed in for loop below
-    for i in range(len(x)):
-        dot += w[i+1]*x[i]
-    if dot > 709:    # prevent math overflows
-        return 1
-    if dot < -709:   # prevent math overflows
-        return 0
-    return 1 / (1 + math.exp(-dot))
-
-
-def label_penalty(predicted, actual):   # non-negative penalty on difference between prediction & true label for group
-    return (predicted - actual) ** 2     # using squared loss as penalty, as described in GICF paper
-
-
-def predict_label(group, instance_labels):
-    # real scalar representing predicted label for group based on aggregation of instance labels
-    total = 0.0     # combined value of all labels in group, computed in for loop below
-    for i in range(len(group)):
-        total += instance_labels[i]
-    return total / len(group)   # return average label value in the group
-
-
-def d_group_penalty(group, instance_labels, d_instance_labels, group_label):
-    # calculate group penalty in derivative of cost function
-    total1 = 0.0    # the part of the group penalty using the actual group labels
-    total2 = np.array([0.0 for i in range(len(d_instance_labels[0]))])   # part using only predicted instance labels
-    for i in range(len(group)):
-        total1 += instance_labels[i]
-        total2 += d_instance_labels[i]
-    total1 = 2.0 * ((total1 / len(group)) - group_label)
-    total2 /= len(group)
-    return total1 * total2
-
-
-def cost(num_instances, w, data, l, group_labels):
-    # calculate cost function as defined in GICF paper
-    instance_cost = 0.0     # first part of cost function, based on instance differences
-    group_cost = 0.0    # second part of cost function, based on group labels
-
[…]
